## Script to clean UKB_master data

```{r}
#import libraries
library(tidyverse)
library(dplyr)
library(ggplot2)

#find wd and load data
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\Core Datasets\\UKB_master.Rda")
```

#### Exploring core demographic variables
For UKB_master (all participants).
The following code can be used a template for exploring further variables.

## Age

Age is right-skewed, no outliers/errors

```{r}
#age
hist(UKB_master$age,
  xlab = "age",
  main = "Histogram of age",
  breaks = sqrt(nrow(UKB_master))
)
```

Age Range:
```{r}
range(UKB_master$age)
```
Mean age +- s.d.
```{r}
c(mean(UKB_master$age), sd(UKB_master$age))
```
Number of NAs:
```{r}
sum(is.na(UKB_master$age))
```

## Sex

Sex is biased towards women (54%).

```{r}
#create sex count dataframe
sex_df <- UKB_master %>%
  group_by(sex) %>%
  summarise(counts = n())

#plot sex as barplot
ggplot(sex_df, aes(x = sex, y = counts)) +
  geom_bar(fill = "#0073C2FF", stat = "identity") +
  geom_text(aes(label = counts), vjust = -0.3)
```

Number of NAs:
```{r}
sum(is.na(UKB_master$sex))
```
Percentage women:
```{r}
sex_df$counts[1]/sum(sex_df$counts)*100
```
Percentage men:
```{r}
sex_df$counts[2]/sum(sex_df$counts)*100
```

## Ethnicity

Majority British ethnicity. 2778 not disclosed.

```{r}
ethnicity_summary <- UKB_master %>%
    group_by(ethnic) %>%
    summarise(counts = n()) 

print(ethnicity_summary)

#count number of NA 
sum(is.na(UKB_master$ethnic), UKB_master$ethnic['Prefer not to answer'], UKB_master$ethnic['Do not know'])

ethnic_subset <- c('Prefer not to answer', 'NA', 'Do not know')

```

## Disability

Majority ppts not disabled, 31% answered 'yes'.

```{r}
disability_summary <- UKB_master %>%
    group_by(disability) %>%
    summarise(counts = n()) 

print(disability_summary)


#plot disability as barplot
ggplot(disability_summary, aes(x = disability, y = counts)) +
  geom_bar(fill = "#0073C2FF", stat = "identity") +
  geom_text(aes(label = counts), vjust = -0.3)
```

Number of NAs:
```{r}
sum(is.na(UKB_master$disability))
```
Percentage disabled:
```{r}
disability_summary$counts[4]/sum(disability_summary$counts)*100
```
Percentage not disabled:
```{r}
disability_summary$counts[3]/sum(disability_summary$counts)*100
```

## Employment

Majority ppts are employed.

```{r}
employment_summary <- UKB_master %>%
    group_by(employed) %>%
    summarise(counts = n()) 

print(employment_summary)

```