---
title: "seasonal_variation_diabetic_get_ppts"
author: "agpr141"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Seasonal variation in diabetic markers - get participants

**Participant extract script**

This .Rmd is used to extract the final participant dataframe for [paper] by Roguski et al. [date].

#### 1. Load packages

```{r install packages, echo=FALSE}
library(tidyverse)
library(dplyr)
library(readr)
```

#### 2. Load initial df

```{r load df, echo=FALSE}
# set cwd
#setwd("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data")

# load in UKB_master
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\ukb_master.Rda")

# make new df which i will now work with
bdmet_df <- ukb_master
rm(ukb_master)

#  control exclusion vars
control_excl_vars <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\normative mets/control_exclusion_participant.tsv")
control_excl_vars <- control_excl_vars %>% 
  rename('eid' = 'Participant ID')

# merge above 2 datasets into 1
bdmet_df_excl <- merge(bdmet_df,control_excl_vars, by="eid")

# remove bdmet_df and control_excl_vars datasets (no longer needed)
rm(bdmet_df, control_excl_vars)

### remove people who shift/night shift work ####
no_shift <- bdmet_df_excl[!grepl("Prefer not to answer|Do not know|Sometimes|Usually|Always", bdmet_df_excl$shift_work),] 
control_excl_vars <- no_shift[!grepl("Prefer not to answer|Do not know|Sometimes|Usually|Always", no_shift$night_shift),]

# remove columns which are superfluous
control_excl_vars <- subset(control_excl_vars, select = -c(end_DXA, FEV, FVC, PEF, handgrip_l, handgrip_r))
```

#### 3. Correct weather

```{r}
library(pillar)
library(readxl)
setwd("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\biobank-time-dimensions\\Projects\\BD-Metabolomics")
temp_table <- read_excel("./weather_ukb_270923.xlsx", sheet = "monthly temp")
colnames(temp_table) <- c("station","assess_year", "assess_month", "tmax", "tmin", "frost", "rain", "sun", "average_temp")
weather_stations <- read_excel("weather_ukb_270923.xlsx", sheet = "ukb centres - weather stations")
#UKB_master <- read.csv("C:/UKB-time/UKB_master") - use ghaem_df instead
colnames(weather_stations) <- c("station","assess_centre_name", "assess_centre")

#setwd("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\biobank-time-dimensions\\Projects\\BD-Metabolomics")

#check station names & clean up
working_df <- control_excl_vars

# check main df
unique(working_df$station)  #lists 12 stations with 1st letter capital

# check weather_stations and remove shef for revisits
unique(weather_stations$station)  #lists 13 stations with 1st letter capital EXCEPT lower case shef (revisit & imaging)

unique(temp_table$station)  #lists 12 stations with 1st letter lower case EXCEPT upper case Card, Brad
#merge weather stations and centre
#working_df  <- analysis_df
working_df  = subset(working_df , select = -c(station,tmax, tmin,frost, rain, sun, average_temp))
working_df <- merge(working_df,weather_stations, by="assess_centre")

#merge dataframes based on year and month and centre
working_df <- merge(working_df, temp_table, by = c("assess_year", "station", "assess_month"), all.x = TRUE)

#rm(analysis_df)
#analysis_df <- working_df
#save(analysis_df,file="analysis_df.Rda")
```

#### 4. Create additional ICD10 exclusions

ICD-10 exclusions:

| Condition                            | ICD-10 code |
|--------------------------------------|-------------|
| Chronic hepatic failure              | K72.1       |
| Chronic kidney disease - stage 4     | N18.4       |
| Chronic kidney disease - stage 5     | N18.5       |
| Anaemia due to enzyme disorders      | D55         |
| Thalassaemia (category)              | D56         |
| Sickle cell disorders (category)     | D57         |
| Other hereditary haemolytic anaemias | D58         |
| Acquired haemolytic anaemia          | D59         |
| Seropositive rheumatoid arthritis    | M05         |
| Other rheumatoid arthritis           | M06         |
|                                      |             |

```{r icd10 exclusion df}
#load existing icd_10s

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\excluded_numbers.Rda")

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\excluded_icd.Rda")

### amend icd10 exclusion criteria variables ####
#Chronic hepatic failure
excluded_numbers$K72.1 <- as.numeric(NA)
#Chronic kidney disease
excluded_numbers$N18.4 <- as.numeric(NA) # stage 4
excluded_numbers$N18.5 <- as.numeric(NA) # stage 5
####haemoglobinopathies
#Anaemia due to enzyme disorders
excluded_numbers$D55 <- as.numeric(NA)
#Thalassaemia (category)
excluded_numbers$D56 <- as.numeric(NA)
#Sickle cell disorders (category)
excluded_numbers$D57 <- as.numeric(NA)
#Other hereditary haemolytic anaemias
excluded_numbers$D58 <- as.numeric(NA)
#Acquired haemolytic anaemia
excluded_numbers$D59 <- as.numeric(NA)
###Rheumatoid arthritis
#Seropositive rheumatoid arthritis
excluded_numbers$M05 <- as.numeric(NA)
#Other rheumatoid arthritis
excluded_numbers$M06 <- as.numeric(NA)
#delete first blank column
#excluded_numbers <- excluded_numbers[-c(1)]

```

run search

```{r}
#rename working_df
control_excl_vars <- working_df

#Chronic hepatic failure
exclusions_icd$K72.1 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$K72.1[i] <- grepl("K72.1", row)}

#count how many true cases there are
excluded_numbers$K72.1  =  sum(exclusions_icd$K72.1, na.rm=TRUE)

#Chronic kidney disease
exclusions_icd$N18.4 <- as.numeric(NA) # stage 4

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$N18.4[i] <- grepl("N18.4", row)}

#count how many true cases there are
excluded_numbers$N18.4  =  sum(exclusions_icd$N18.4, na.rm=TRUE)

exclusions_icd$N18.5 <- as.numeric(NA) # stage 5

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$N18.5[i] <- grepl("N18.5", row)}

#count how many true cases there are
excluded_numbers$N18.5  =  sum(exclusions_icd$N18.5, na.rm=TRUE)

####haemoglobinopathies
#Anaemia due to enzyme disorders
exclusions_icd$D55 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$D55[i] <- grepl("D55", row)}

#count how many true cases there are
excluded_numbers$D55  =  sum(exclusions_icd$D55, na.rm=TRUE)

#Thalassaemia (category)
exclusions_icd$D56 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$D56[i] <- grepl("D56", row)}

#count how many true cases there are
excluded_numbers$D56  =  sum(exclusions_icd$D56, na.rm=TRUE)

#Sickle cell disorders (category)
exclusions_icd$D57 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$D57[i] <- grepl("D57", row)}

#count how many true cases there are
excluded_numbers$D57  =  sum(exclusions_icd$D57, na.rm=TRUE)

#Other hereditary haemolytic anaemias
exclusions_icd$D58 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$D58[i] <- grepl("D58", row)}

#count how many true cases there are
excluded_numbers$D58  =  sum(exclusions_icd$D58, na.rm=TRUE)

#Acquired haemolytic anaemiaD
exclusions_icd$D59 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$D59[i] <- grepl("D59", row)}

#count how many true cases there are
excluded_numbers$D59  =  sum(exclusions_icd$D59, na.rm=TRUE)

###Rheumatoid arthritis
#Seropositive rheumatoid arthritis
exclusions_icd$M05 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$M05[i] <- grepl("M05", row)}

#count how many true cases there are
excluded_numbers$M05  =  sum(exclusions_icd$M05, na.rm=TRUE)

#Other rheumatoid arthritis
exclusions_icd$M06 <- as.numeric(NA)

for (i in 1:nrow(control_excl_vars)){
  row <- control_excl_vars$'Diagnoses - ICD10'[i]
  exclusions_icd$M06[i] <- grepl("M06", row)}

#count how many true cases there are
excluded_numbers$M06  =  sum(exclusions_icd$M06, na.rm=TRUE)

working_df <- control_excl_vars
```

```{r}
#exclusions_icd <- exclusions_icd %>% 
#  rename('eid' = 'control_excl_vars.eid')
save(excluded_numbers,file="excluded_numbers.Rda")
save(exclusions_icd,file="excluded_icd.Rda")
```

#### Load in excluded above and then extra variables for analysis

```{r}
fasting_time <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\fasting_time.tsv")

fasting_time <- fasting_time %>% 
  rename('fasting_time' = 'p74_i0')

working_df <- merge(working_df,fasting_time, by="eid")

townsend <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\townsend.tsv")

townsend <- townsend %>% 
  rename('townsend' = 'p22189')

working_df <- merge(working_df,townsend, by="eid")


sleep_activity <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\sleep_activity.tsv")

sleep_activity <- sleep_activity %>% 
  rename('sleep_duration_sr' = '1160-0.0',
         'sleep_avg_acti' = '40042-0.0',
         'sedentary_avg_acti' = '40043-0.0',
         'moderate_avg_acti' = '40045-0.0',
         'light_avg_acti' = '40044-0.0',
         'acti_data_qual' = '90015-0.0',
         'IPAQ_group' = '22032-0.0')

working_df <- merge(working_df,sleep_activity, by="eid")
```

#### Load in blood biomarkers and metabolites

```{r}
blood_biochem_df <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\blood_biochemistry.tsv")
blood_datetime <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\blood_datetime_3166_participant.tsv")


#rename tsv variable columns
blood_biochem_df <- blood_biochem_df %>% 
  rename('alanine_at' = '30620-0.0',
         'albumin' = '30600-0.0',
         'alkaline_phos' = '30610-0.0',
         'apolipo_a' = '30630-0.0',
         'apolipo_b' = '30640-0.0',
         'aspartate_at' = '30650-0.0',
         'crp' = '30710-0.0',
         'calcium' = '30680-0.0',
         'cholesterol' = '30690-0.0',
         'creatinine' = '30700-0.0',
         'cystatin_c' = '30720-0.0',
         'dir_bilirubin' = '30660-0.0',
         'gamma_gmt' = '30730-0.0',
         'glucose' = '30740-0.0',
         'g_haem' = '30750-0.0',
         'hdl_chol' = '30760-0.0',
         'igf_1' = '30770-0.0',
         'ldl_direct' = '30780-0.0',
         'lipoprotein_a' = '30790-0.0',
         'oestradiol' = '30800-0.0',
         'phosphate' = '30810-0.0',
         'rheum_factor' = '30820-0.0',
         'shbg' = '30830-0.0',
         'testosterone' = '30850-0.0',
         'tot_bilirubin' = '30840-0.0',
         'tot_protein' = '30860-0.0',
         'triglycerides' = '30870-0.0',
         'urate' = '30880-0.0',
         'urea' = '30670-0.0',
         'vit_d' = '30890-0.0',
         )

#add comments to variable to link with biobank datafield id

comment(blood_biochem_df$alanine_at) <- c("Datafield = 30620-0.0")
comment(blood_biochem_df$albumin) <- c("Datafield = 30600-0.0")
comment(blood_biochem_df$alkaline_phos) <- c("Datafield = 30610-0.0")
comment(blood_biochem_df$apolipo_a) <- c("Datafield = 30630-0.0")
comment(blood_biochem_df$apolipo_b) <- c("Datafield = 30640-0.0")
comment(blood_biochem_df$aspartate_at) <- c("Datafield = 30650-0.0")
comment(blood_biochem_df$crp) <- c("Datafield = 30710-0.0")
comment(blood_biochem_df$calcium) <- c("Datafield = 30680-0.0")
comment(blood_biochem_df$cholesterol) <- c("Datafield = 30690-0.0")
comment(blood_biochem_df$creatinine) <- c("Datafield = 30700-0.0")
comment(blood_biochem_df$cystatin_c) <- c("Datafield = 30720-0.0")
comment(blood_biochem_df$dir_bilirubin) <- c("Datafield = 30660-0.0")
comment(blood_biochem_df$gamma_gmt) <- c("Datafield = 30730-0.0")
comment(blood_biochem_df$glucose) <- c("Datafield = 30740-0.0")
comment(blood_biochem_df$g_haem) <- c("Datafield = 30750-0.0")
comment(blood_biochem_df$hdl_chol) <- c("Datafield = 30760-0.0")
comment(blood_biochem_df$igf_1) <- c("Datafield = 30770-0.0")
comment(blood_biochem_df$ldl_direct) <- c("Datafield = 30780-0.0")
comment(blood_biochem_df$lipoprotein_a) <- c("Datafield = 30790-0.0")
comment(blood_biochem_df$oestradiol) <- c("Datafield = 30800-0.0")
comment(blood_biochem_df$phosphate) <- c("Datafield = 30810-0.0")
comment(blood_biochem_df$rheum_factor) <- c("Datafield = 30820-0.0")
comment(blood_biochem_df$shbg) <- c("Datafield = 30830-0.0")
comment(blood_biochem_df$testosterone) <- c("Datafield = 30850-0.0")
comment(blood_biochem_df$tot_bilirubin) <- c("Datafield = 30840-0.0")
comment(blood_biochem_df$tot_protein) <- c("Datafield = 30860-0.0")
comment(blood_biochem_df$triglycerides) <- c("Datafield = 30870-0.0")
comment(blood_biochem_df$urate) <- c("Datafield = 30880-0.0")
comment(blood_biochem_df$urea) <- c("Datafield = 30670-0.0")
comment(blood_biochem_df$vit_d) <- c("Datafield = 30890-0.0")


# exclude markers that are not of interest to us
metabolites <- blood_biochem_df[c("eid", "alanine_at","albumin","apolipo_a", "apolipo_b", "crp", "cholesterol","creatinine", "cystatin_c", "dir_bilirubin", "gamma_gmt", "glucose", "g_haem", "hdl_chol", "ldl_direct", "lipoprotein_a", "phosphate", "rheum_factor", "tot_bilirubin", "triglycerides", "vit_d")]


# merge with existing bdmet_df
working_df <- merge(working_df,metabolites, by="eid")
working_df <- merge(working_df,blood_datetime, by="eid")

working_df <- working_df %>% 
  rename('blood sample datetime' = '3166-0.0')

```

#### Get blood sample time data

```{r}
# load participant dataframe for analysis
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Metabolite Analyses\\analysis_met.Rda")

# get blood sample timings
bs_vars <- readr::read_tsv("C:/Users/arogusk2/OneDrive - University of Edinburgh./HELIOS-BD/Side Projects/BioBank Project/core dataset/Sample timings/blood_datetime_3166_participant.tsv")

#rename variable
bs_vars <- bs_vars %>% 
  rename('BS_datetime' = '3166-0.0')

# initialise columns
bs_vars$bs_year <- as.numeric(NA)
bs_vars$bs_month <- as.numeric(NA)
#bs_vars$bs_mday <- as.numeric(NA)
#bs_vars$bs_wday <- as.numeric(NA)
bs_vars$bs_time <- as.numeric(NA)

#create forloop to pull out year, month, time and day of week
for (i in 1:nrow(bs_vars)){
  row <- bs_vars[i,]
  
  #convert to POSIXlt
  lt <- as.POSIXlt(bs_vars$BS_datetime[i])
  
  #pull out year from data (lt format = # years since 1900)
  year <- 1900+lt$year
  bs_vars$bs_year[i] <- year
  
  # pull out month (zero indexed, add 1)
  month <- 1+lt$mon
  bs_vars$bs_month[i] <- month
  
  # pull out day of the month
  #mday <- lt$mday
  #bs_vars$bs_mday[i] <- mday
  
  # pull out day of the week
  #wday <- lt$wday
  #bs_vars$bs_wday[i] <- wday
  
  # pull out ToD
  time <- format(as.POSIXct(bs_vars$BS_datetime[i]), format = "%H:%M")
  bs_vars$bs_time[i] <- time
}

#merge bs_vars into main df
working_df <- merge(working_df,bs_vars, by="eid")
```

Tidy data

```{r}
# remove people with NA glycated haemoglobin
ghaem_df <- working_df %>%  filter(!is.na(g_haem))

# remove people with NA glucose
ghaem_df <- ghaem_df %>%  filter(!is.na(glucose))

#remove people with glycated haemoglobin levels outside of manufacturer's reportable range
ghaem_df<- ghaem_df %>%  filter(!g_haem>184)
ghaem_df<- ghaem_df %>%  filter(!g_haem<15)

# remove people with NA variables: bs_datetime (removes 2)
#ghaem_df<- ghaem_df %>%  filter(!is.na(BS_datetime))

# make season numeric
#ghaem_df$bs_season <- as.numeric(as.character(ghaem_df$bs_season))

# tidy time outdoors data to remove '-10', '-3', '-1' values:
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-10] <- 0
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-3] <- NA
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-1] <- NA

ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-10] <- 0
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-3] <- NA
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-1] <- NA

# tidy self-report sleep time to remove -3', '-1' values:
ghaem_df$sleep_duration[ghaem_df$sleep_duration ==-3] <- NA
ghaem_df$sleep_duration[ghaem_df$sleep_duration ==-1] <- NA


# recode ethnicity into 7 groups
library('plyr')
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer
ghaem_df$ethnicity_recode <- revalue(ghaem_df$ethnicity, c("British"="1", "Any other white background"="1", "White"="1", "Irish"="1", "Mixed"="2", "White and Black Caribbean"="2", "White and Black African"="2", "White and Asian"="2", "Any other mixed background"="2", "Asian or Asian British"="3", "Indian"="3", "Pakistani"="3", "Bangladeshi"="3", "Any other Asian background"="3", "Black or Black British"="4", "Caribbean"="4", "African"="4", "Any other Black background"="4", "Chinese"="5", "Other ethnic group"="6", "Do not know"="7", "Prefer not to answer"="7"))
#ghaem_df$ethnicity_recode <- as.numeric(ghaem_df$ethnicity_recode)

#ethnicity dummy vars code
#white
ghaem_df$ethnicity_white <- revalue(ghaem_df$ethnicity_recode, c("1"="1", "2"="0", "3"="0", "4"="0", "5"="0", "6"="0", "7"="0"))

#mixed
ghaem_df$ethnicity_mixed <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="1", "3"="0", "4"="0", "5"="0", "6"="0", "7"="0"))

#asian/asian british
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="1", "4"="0", "5"="0", "6"="0", "7"="0"))

#black/black british
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="1", "5"="0", "6"="0", "7"="0"))

#chinese
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="0", "5"="1", "6"="0", "7"="0"))

#other
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="0", "5"="0", "6"="1", "7"="0"))

# recode employment into 5 groups
#1=employed, 2=retired, 3=unable to work because of sickness/disability, 4=other unemployment, 5=do not know/prefer not to answer
ghaem_df$employment_recode <- revalue(ghaem_df$employed, c("In paid employment or self-employed"="1", "Retired"="2", "Doing unpaid or voluntary work"="4", "Unable to work because of sickness or disability"="3", "Full or part-time student"="4", "Unemployed"="4", "Looking after home and/or family"="4", "None of the above"="4", "Prefer not to answer"="5"))

#remove plyr package as has masking for later functions
detach("package:plyr", unload=TRUE)

library('plyr')
# recode vars
##Group
#control
#ghaem_df$group <- revalue(ghaem_df$Group, c("control"="1", "bipolar"="2"))
#ghaem_df$group_f <- factor(ghaem_df$group, levels= c("1", "2"), labels= c("control", "bipolar"), ordered=FALSE)

### LEVEL 1
#Sex
#sex
#ghaem_df$sex_recode <- revalue(ghaem_df$sex, c("Male"="1", "Female"="2"))
ghaem_df$sex_f <- factor(ghaem_df$sex, ordered=FALSE)

#ghaem_df$sex_fact <- factor(ghaem_df$sex_recode, levels= c("Male", "Female"), labels=c("male", "female"))

#ethnicity
# set values to NA if not relevant
ghaem_df$ethnicity_recode[ghaem_df$ethnicity_recode == -1] <- NA
ghaem_df$ethnicity_recode[ghaem_df$ethnicity_recode == -3] <- NA
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer
ghaem_df$ethnicity_f <- factor(ghaem_df$ethnicity_recode, levels= c("1", "2", "3", "4", "5", "6"), labels= c("white", "mixed", "asian", "black", "chinese", "other"), ordered=FALSE)

### LEVEL 2 
#smoking
# set values to NA if not relevant
ghaem_df$smoking_f <- ghaem_df$smoking_status
ghaem_df$smoking_f[ghaem_df$smoking_f == -3] <- NA
ghaem_df$smoking_f <- factor(ghaem_df$smoking_f, levels= c("Never", "Previous", "Current"), labels= c("Never", "Previous", "Current"), ordered=FALSE)

#alcohol - showing up strange!
#recode: never/special occasions=1 (LOW), 1-3month/1-2week=2 (MODERATE), 3-4week/daily=3 (HIGH)
ghaem_df$alcohol_f <- as.character(ghaem_df$alcohol_intake)
#ghaem_df$alcohol_f[ghaem_df$alcohol_f == 'Prefer not to answer'] <- NA
ghaem_df$alcohol_f <- revalue(ghaem_df$alcohol_f, c("Never"="Low", "Special occasions only"="Low", "One to three times a month"="Moderate", "Once or twice a week"="Moderate","Three or four times a week"="High", "Daily or almost daily"="High", "Prefer not to answer"=NA))
ghaem_df$alcohol_f <- factor(ghaem_df$alcohol_f, levels= c("Low", "Moderate", "High"), labels= c("Low", "Moderate", "High"), ordered=TRUE)

#health self-report
ghaem_df$health_f <- as.character(ghaem_df$health_self_report)
ghaem_df$health_f[ghaem_df$health_f == 'Prefer not to answer'] <- NA
ghaem_df$health_f[ghaem_df$health_f == 'Do not know'] <- NA
ghaem_df$health_f <- factor(ghaem_df$health_f, levels= c("Excellent", "Good", "Fair", "Poor"), labels= c("Excellent", "Good", "Fair", "Poor"), ordered=TRUE)


#disbility
ghaem_df$disability_f <- as.character(ghaem_df$disability_self_report)
ghaem_df$disability_f[ghaem_df$disability_f == 'Prefer not to answer'] <- NA
ghaem_df$disability_f[ghaem_df$disability_f == 'Do not know'] <- NA
ghaem_df$disability_f <- factor(ghaem_df$disability_f, levels= c("No", "Yes"), labels= c("No", "Yes"), ordered=FALSE)


#chronotype
ghaem_df$chrono_f <- as.character(ghaem_df$chronotype)
ghaem_df$chrono_f[ghaem_df$chrono_f == 'Prefer not to answer'] <- NA
ghaem_df$chrono_f[ghaem_df$chrono_f == 'Do not know'] <- NA
ghaem_df$chrono_f <- factor(ghaem_df$chrono_f, ordered=TRUE)

### Level 4 
#assess_centre to factor
ghaem_df$centre_f <- factor(ghaem_df$assess_centre, ordered=FALSE)
```

#### **create participant groups**

```{r}
### remove people with self-reported diabetes dx ####
#unique(no_shift_2$`Diabetes diagnosed by doctor | Instance 0`)
#no_dm <- control_excl_vars[!grepl("Yes", control_excl_vars$'Diabetes diagnosed by doctor | Instance 0'),]
#unique(bdmet_df$`Gestational diabetes only | Instance 0`)
#no_dm_2 <- no_dm[!grepl("Yes", no_dm$'Gestational diabetes only | Instance 0'),]

### remove people with probable single/recurrent major depression ####
no_md <- ghaem_df[!grepl("Probable", ghaem_df$`Bipolar and major depression status | Instance 0`),]

#merge icd search output variables with control_excl_vars ####
study_population <- merge(no_md,exclusions_icd, by="eid")

### CONTROL DIABETES group: remove people who meet any of the exclusion criteria ####
c_dm <- study_population %>%
  filter(F20!='1' & F21!='1'& F22!='1' & F23!='1' & F24!='1' & F25!='1'& F27!='1' &
           F28!='1' & F30!='1' & F31!='1'& F32!='1'& F33!='1' & F34!='1' &
           F38.1!='1' & F39!='1' & E10!='0'| E11!='0' | E12!='0' | E13!='0' |
           E14!='0' & F50!='1'& E70_E72!='1'& E74!='1' & E75!='1' & E76!='1' & 
           E77!='1' & E78!='1' & E79!='1' & E80!='1' & G47.2!='1'& F51.2!='1' &
           K72.1!='1' & N18.4!='1' & N18.5!='1' & D55!='1' & D56!='1'& D57!='1' &
           D58!='1' & D59!='1' & M05!='1' & M06!='1'
           )
#remove people with probable bipolar
c_dm_group <- c_dm[!grepl("Disorder", c_dm$'Bipolar and major depression status | Instance 0'),]

### CONTROL NON-DIABETES group: remove people who meet any of the exclusion criteria ####
c_nondm <- study_population %>%
  filter(F20!='1' & F21!='1'& F22!='1' & F23!='1' & F24!='1' & F25!='1'& F27!='1' &
           F28!='1' & F30!='1' & F31!='1'& F32!='1'& F33!='1' & F34!='1' &
           F38.1!='1' & F39!='1' & E10!='1'& E11!='1' & E12!='1' & E13!='1' &
           E14!='1' & F50!='1'& E70_E72!='1'& E74!='1' & E75!='1' & E76!='1' & 
           E77!='1' & E78!='1' & E79!='1' & E80!='1' & G47.2!='1'& F51.2!='1' &
           K72.1!='1' & N18.4!='1' & N18.5!='1' & D55!='1' & D56!='1'& D57!='1' &
           D58!='1' & D59!='1' & M05!='1' & M06!='1'
           )
#remove people with probable bipolar
c_nondm_gr <- c_nondm[!grepl("Disorder", c_nondm$'Bipolar and major depression status | Instance 0'),]

#remove people with doctor diabetes
c_nondm_group <- c_nondm_gr[!grepl("Yes", c_nondm_gr$'Diabetes diagnosed by doctor | Instance 0'),]

### BIPOLAR DIABETES group: remove people who meet any of the exclusion criteria ####
bd_dm <- study_population %>%
  filter(F20!='1' & F21!='1'& F22!='1' & F23!='1' & F24!='1' & F25!='1'& F27!='1' &
           F28!='1' &
           F32!='1'& F33!='1' & F34!='1' & F38.1!='1' & F39!='1' &
           E10!='0'| E11!='0' | E12!='0' | E13!='0' | E14!='0' &
           F50!='1'&
           E70_E72!='1'& E74!='1' & E75!='1' & E76!='1' & E77!='1' & E78!='1' &
           E79!='1' & E80!='1' &
           G47.2!='1'& F51.2!='1' &
           K72.1!='1' & N18.4!='1' & N18.5!='1' & D55!='1' & D56!='1'& D57!='1' &
           D58!='1' & D59!='1' & M05!='1' & M06!='1')

# create new bipolar group of people who have either: ICD10 bipolar disorder (F31) or have probable bipolar (Pell & Smith criteria)
bd_dm_group <- dplyr::filter(bd_dm, grepl('Bipolar I Disorder|Bipolar II Disorder', `Bipolar and major depression status | Instance 0`)|bd_dm$F31!='0')

### BIPOLAR NON-DIABETES group: remove people who meet any of the exclusion criteria ####
bd_nondm <- study_population %>%
  filter(F20!='1' & F21!='1'& F22!='1' & F23!='1' & F24!='1' & F25!='1'& F27!='1' &
           F28!='1' &
           F32!='1'& F33!='1' & F34!='1' & F38.1!='1' & F39!='1' & 
           E10!='1'& E11!='1' & E12!='1' & E13!='1' & E14!='1' &
           F50!='1'&
           E70_E72!='1'& E74!='1' & E75!='1' & E76!='1' & E77!='1' & E78!='1' &
           E79!='1' & E80!='1' &
           G47.2!='1'& F51.2!='1' &
           K72.1!='1' & N18.4!='1' & N18.5!='1' & D55!='1' & D56!='1'& D57!='1' &
           D58!='1' & D59!='1' & M05!='1' & M06!='1')

# create new bipolar group of people who have either: ICD10 bipolar disorder (F31) or have probable bipolar (Pell & Smith criteria)
bd_nondm_group <- dplyr::filter(bd_nondm, grepl('Bipolar I Disorder|Bipolar II Disorder', `Bipolar and major depression status | Instance 0`)|bd_nondm$F31!='0')

```

#### save 4x groups for analysis

```{r save groups}
setwd("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Seasonal Variation Diabetic Markers - Analysis")
save(bd_dm_group,file="bd_dm_group.Rda")
save(bd_nondm_group,file="bd_nondm_group.Rda")
save(c_dm_group,file="c_dm_group.Rda")
save(c_nondm_group,file="c_nondm_group.Rda")

# Combine both group dataframes into one for future analysis purposes ####
#add new column to both with study group
bd_dm_new <- bd_dm_group %>%
  mutate(group = 'bipolar dm')
bd_nondm_new <- bd_nondm_group %>%
  mutate(group = 'bipolar nondm')
c_dm_new <- c_dm_group %>%
  mutate(group = 'control dm')
c_nondm_new <- c_nondm_group %>%
  mutate(group = 'control nondm')

#join 2 dataframes
svdm_all_ppts <- rbind(bd_dm_new,bd_nondm_new,c_dm_new, c_nondm_new)
#save
save(svdm_all_ppts,file="svdm_all_ppts.Rda")
```
