---
title: "Biometric_Time_Graphs"
output: html_document
date: "2024-03-20"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages
  
lubridate is used to convert hour, min, seconds (hms) to seconds
```{r}
library(lubridate)
library(ggplot2)
library(dplyr)
library(gridExtra)
```

### Read in data and convert times to decimal hours
```{r}
data <- read.table("//wsl.localhost/Ubuntu/home/lfahey/ukb_concept/bp_times_chrono.tab", header=T, fill = TRUE)
mins <- lubridate::hms(data$time)
data$biometrics_seconds <- period_to_seconds(mins)
data$hours <- data$biometrics_seconds / 3600
```

# Plot time ranges

```{r}
# Read in file
biomet <- read.table("//wsl.localhost/Ubuntu/home/lfahey/ukb_concept/biometrics_stats.txt", header=T, fill = TRUE)

biomet

# Make barplot
barplot(biomet$NumberParticipants, names=biomet$TimeRange, xlab = "Hours", ylab = "Number of participants", main = "Times participants completed biometric tests")
```
Based on above, we could use a threshold of 10,000 samples - This would mean removing samples that completed biometric tests before 9am and after 8pm.

Create an additional column in data, called hour bin, which bins samples in bins based on the hour they completed the biometric tests.

```{r}
data <- data[data$hours > 9 & data$hours < 20,]
data$hour_bins <- cut(data$hours, breaks = c(9:20), labels = c(9:19))
```


## In the next could block, I am getting data ready to plot average measures for each time bin for different chronotypes
1. Create seperate data frames for four different chronotypes \
2. Get the average measure value for each hour bin \
3. Add chronotype as an extra column to each df \
4. Bind all 4 dfs by rows 

```{r}
# Def Evening

def_owl <- data[data$chrono == 4,]
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ hour_bins, data = def_owl, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ hour_bins, data = def_owl, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ hour_bins, data = def_owl, FUN = mean)
mean_grip_def_owls <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_def_owls$chronotype <- "DefinetlyEvening"

# More Evening
more_owl <- data[data$chrono == 3,]
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ hour_bins, data = more_owl, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ hour_bins, data = more_owl, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ hour_bins, data = more_owl, FUN = mean)
mean_grip_more_owls <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_more_owls$chronotype <- "MoreEvening"

# More Morning
more_lark <- data[data$chrono == 2,]
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ hour_bins, data = more_lark, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ hour_bins, data = more_lark, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ hour_bins, data = more_lark, FUN = mean)
mean_grip_more_larks <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_more_larks$chronotype <- "MoreMorning"

# Def Morning
def_lark <- data[data$chrono == 1,]
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ hour_bins, data = def_lark, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ hour_bins, data = def_lark, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ hour_bins, data = def_lark, FUN = mean)
mean_grip_def_larks <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_def_larks$chronotype <- "DefinetlyMorning"

mean_grip_chronotype <- rbind(mean_grip_def_larks, mean_grip_more_larks, mean_grip_def_owls, mean_grip_more_owls)
mean_grip_chronotype$chronotype <- gsub("Definetly", "Definitely", mean_grip_chronotype$chronotype)
mean_grip_chronotype <- mean_grip_chronotype[,c("hour_bins", "f.4079.0.0", "f.102.0.0", "f.4080.0.0", "chronotype")]

head(mean_grip_chronotype)

# Make hour_bins numeric (required for making the line graph)
mean_grip_chronotype$hour_bins <- as.numeric(as.character(mean_grip_chronotype$hour_bins))
```


## Make Plots or all measures

# Make hours to be used as x-axis ticks and labels
```{r}
hours <- c(9:19)
```

```{r}

systolic_bp <- ggplot(data = mean_grip_chronotype, aes(x = hour_bins, y = f.4080.0.0, colour = chronotype)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = hours, 
                     labels = hours) +
  labs(x = "Time Bins", y = "Systolic blood pressure") +
  ggtitle("Systolic blood pressure by chronotype") +
  theme_minimal()

systolic_bp
```

```{r}

pulse_rate <- ggplot(data = mean_grip_chronotype, aes(x = hour_bins, y = f.102.0.0, colour = chronotype)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = hours, 
                     labels = hours) +
  labs(x = "Time Bins", y = "Pulse rate") +
  ggtitle("Pulse rate by chronotype") +
  theme_minimal()

pulse_rate
```

```{r}

Diastolic_bp <- ggplot(data = mean_grip_chronotype, aes(x = hour_bins, y = f.4079.0.0, colour = chronotype)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = hours, 
                     labels = hours) +
  labs(x = "Time Bins", y = "Diastolic blood pressure") +
  ggtitle("Diastolic blood pressure by chronotype") +
  theme_minimal()

Diastolic_bp
```


# To add confidence interval shaded region (haven't figured out how to get CIs for mean just yet):
p<-p+geom_ribbon(aes(ymin=data$lower, ymax=data$upper), linetype=2, alpha=0.1)
