---
title: "Biometric Tests"
output:
  html_document: default
  pdf_document: default
date: "2024-02-27"
---

## Load packages
  
lubridate is used to convert hour, min, seconds (hms) to seconds
```{r}
library(lubridate)
library(ggplot2)
library(dplyr)
library(gridExtra)
```

### Read in data and convert times to decimal hours
```{r}
data <- read.table("//wsl.localhost/Ubuntu/home/lfahey/ukb_concept/grip_chrono.tab", header=T, fill = TRUE)
mins <- lubridate::hms(data$time)
data$biometrics_seconds <- period_to_seconds(mins)
data$hours <- data$biometrics_seconds / 3600
```

# Plot time ranges

```{r}
# Read in file
biomet <- read.table("//wsl.localhost/Ubuntu/home/lfahey/ukb_concept/biometrics_stats.txt", header=T, fill = TRUE)

# Make barplot
barplot(biomet$NumberParticipants, names=biomet$TimeRange, xlab = "Hours", ylab = "Number of participants", main = "Times participants completed biometric tests")
```
### Get data ready to plot average measures for each time bins for different chronotypes
1. Create seperate data frames for four different chronotypes
2. For each df, create 12 time bins, with the same number of samples in each bin
3. Get the average measure value for each time bin
4. Add chronotype as an extra column to each df
5. Bind all 4 dfs by rows

```{r}
### Add time bins col to data
data$time_bins <- ntile(data$hours, 12)

# Understanding created time bins
one <- data[data$time_bins == 1,]
two <- data[data$time_bins == 2,]
three <- data[data$time_bins == 3,]
four <- data[data$time_bins == 4,]
five <- data[data$time_bins == 5,]
six <- data[data$time_bins == 6,]
seven <- data[data$time_bins == 7,]
eight <- data[data$time_bins == 8,]
nine <- data[data$time_bins == 9,]
ten <- data[data$time_bins == 10,]
eleven <- data[data$time_bins == 11,]
twelve <- data[data$time_bins == 12,]

min(twelve$hours)
max(twelve$hours)
max(eleven$hours)
min(eleven$hours)


# Def Evening

def_owl <- data[data$chrono == 4,]

#def_owl$time_bins <- ntile(def_owl$hours, 12)
mean_grip_per_bin_right <- aggregate(right ~ time_bins, data = def_owl, FUN = mean)
mean_grip_per_bin_left <- aggregate(left ~ time_bins, data = def_owl, FUN = mean)
mean_grip_def_owls <- cbind(mean_grip_per_bin_right, mean_grip_per_bin_left)
mean_grip_def_owls$chronotype <- "DefinetlyEvening"

# Look
head(def_owl)
head(def_owl[def_owl$time_bins ==1,])

# More Evening
more_owl <- data[data$chrono == 3,]

#more_owl$time_bins <- ntile(more_owl$hours, 12)
mean_grip_per_bin_right <- aggregate(right ~ time_bins, data = more_owl, FUN = mean)
mean_grip_per_bin_left <- aggregate(left ~ time_bins, data = more_owl, FUN = mean)
mean_grip_more_owls <- cbind(mean_grip_per_bin_right, mean_grip_per_bin_left)
mean_grip_more_owls$chronotype <- "MoreEvening"

# More Morning
more_lark <- data[data$chrono == 2,]

#more_lark$time_bins <- ntile(more_lark$hours, 12)
mean_grip_per_bin_right <- aggregate(right ~ time_bins, data = more_lark, FUN = mean)
mean_grip_per_bin_left <- aggregate(left ~ time_bins, data = more_lark, FUN = mean)
mean_grip_more_larks <- cbind(mean_grip_per_bin_right, mean_grip_per_bin_left)
mean_grip_more_larks$chronotype <- "MoreMorning"

# Def Morning
def_lark <- data[data$chrono == 1,]

#def_lark$time_bins <- ntile(def_lark$hours, 12)
mean_grip_per_bin_right <- aggregate(right ~ time_bins, data = def_lark, FUN = mean)
mean_grip_per_bin_left <- aggregate(left ~ time_bins, data = def_lark, FUN = mean)
mean_grip_def_larks <- cbind(mean_grip_per_bin_right, mean_grip_per_bin_left)
mean_grip_def_larks$chronotype <- "DefinetlyMorning"



mean_grip_chronotype <- rbind(mean_grip_def_larks, mean_grip_more_larks, mean_grip_def_owls, mean_grip_more_owls)
mean_grip_chronotype$chronotype <- gsub("Definetly", "Definitely", mean_grip_chronotype$chronotype)
mean_grip_chronotype <- mean_grip_chronotype[,c("time_bins", "right", "left", "chronotype")]

head(mean_grip_chronotype)

```


#### Plot 

Get acrophase (top y-value (grip-strength measure for each time bin, for each chronotype))

```{r}
mean_grip_def_larks[which.max(mean_grip_def_larks$right), ]
mean_grip_more_larks[which.max(mean_grip_more_larks$right), ]
mean_grip_more_owls[which.max(mean_grip_more_owls$right), ]
mean_grip_def_owls[which.max(mean_grip_def_owls$right), ]
```


```{r}
right <- ggplot(data = mean_grip_chronotype, aes(x = time_bins, y = right, colour = chronotype)) +
  geom_line() +
  geom_point(aes(size = ifelse(time_bins == 1, 1, 0.5)), show.legend = FALSE) +
  scale_size_continuous(range = c(1, 5)) +  # Adjust the range of sizes as needed
  labs(x = "Time Bins", y = "Right Grip Strength") +
  ggtitle("Mean Right Grip Strength by Chronotype") +
  theme_minimal()

right

#grid.arrange(
#  right,
#  left,
#  ncol=2,
#  nrow=1)

```

```{r}
left <- ggplot(data = mean_grip_chronotype, aes(x = time_bins, y = left, colour = chronotype)) +
  geom_line() +
  geom_point(aes(size = ifelse(time_bins == 1, 1, 0.5)), show.legend = FALSE) +
  scale_size_continuous(range = c(1, 5)) +  # Adjust the range of sizes as needed
  labs(x = "Time Bins", y = "Right Grip Strength") +
  ggtitle("Mean Left Grip Strength by Chronotype") +
  theme_minimal()

left
```


```
```{r}
data <- read.table("//wsl.localhost/Ubuntu/home/lfahey/ukb_concept/bp_times_chrono.tab", header=T, fill = TRUE)
mins <- lubridate::hms(data$time)
data$biometrics_seconds <- period_to_seconds(mins)
data$hours <- data$biometrics_seconds / 3600
```

```{r}
### Add time bins col to data
data$time_bins <- ntile(data$hours, 12)

# Understanding created time bins
one <- data[data$time_bins == 1,]
two <- data[data$time_bins == 2,]
three <- data[data$time_bins == 3,]
four <- data[data$time_bins == 4,]
five <- data[data$time_bins == 5,]
six <- data[data$time_bins == 6,]
seven <- data[data$time_bins == 7,]
eight <- data[data$time_bins == 8,]
nine <- data[data$time_bins == 9,]
ten <- data[data$time_bins == 10,]
eleven <- data[data$time_bins == 11,]
twelve <- data[data$time_bins == 12,]

min(twelve$hours)
max(twelve$hours)
max(eleven$hours)
min(eleven$hours)


# Def Evening

def_owl <- data[data$chrono == 4,]

#def_owl$time_bins <- ntile(def_owl$hours, 12)
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ time_bins, data = def_owl, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ time_bins, data = def_owl, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ time_bins, data = def_owl, FUN = mean)
mean_grip_def_owls <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_def_owls$chronotype <- "DefinetlyEvening"

# Look
head(def_owl)
head(def_owl[def_owl$time_bins ==1,])

# More Evening
more_owl <- data[data$chrono == 3,]

#more_owl$time_bins <- ntile(more_owl$hours, 12)
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ time_bins, data = more_owl, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ time_bins, data = more_owl, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ time_bins, data = more_owl, FUN = mean)
mean_grip_more_owls <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_more_owls$chronotype <- "MoreEvening"

# More Morning
more_lark <- data[data$chrono == 2,]

#more_lark$time_bins <- ntile(more_lark$hours, 12)
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ time_bins, data = more_lark, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ time_bins, data = more_lark, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ time_bins, data = more_lark, FUN = mean)
mean_grip_more_larks <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_more_larks$chronotype <- "MoreMorning"

# Def Morning
def_lark <- data[data$chrono == 1,]

#def_lark$time_bins <- ntile(def_lark$hours, 12)
mean_grip_per_bin_f.4079.0.0 <- aggregate(f.4079.0.0 ~ time_bins, data = def_lark, FUN = mean)
mean_grip_per_bin_f.102.0.0 <- aggregate(f.102.0.0 ~ time_bins, data = def_lark, FUN = mean)
mean_grip_per_bin_f.4080.0.0 <- aggregate(f.4080.0.0 ~ time_bins, data = def_lark, FUN = mean)
mean_grip_def_larks <- cbind(mean_grip_per_bin_f.4079.0.0, mean_grip_per_bin_f.102.0.0, mean_grip_per_bin_f.4080.0.0)
mean_grip_def_larks$chronotype <- "DefinetlyMorning"



mean_grip_chronotype <- rbind(mean_grip_def_larks, mean_grip_more_larks, mean_grip_def_owls, mean_grip_more_owls)
mean_grip_chronotype$chronotype <- gsub("Definetly", "Definitely", mean_grip_chronotype$chronotype)
mean_grip_chronotype <- mean_grip_chronotype[,c("time_bins", "f.4079.0.0", "f.102.0.0", "f.4080.0.0", "chronotype")]

head(mean_grip_chronotype)

```


#### Plot 

Get acrophase (top y-value (grip-strength measure for each time bin, for each chronotype))

```{r}
mean_grip_def_larks[which.max(mean_grip_def_larks$f.4079.0.0), ]
mean_grip_more_larks[which.max(mean_grip_more_larks$f.4079.0.0), ]
mean_grip_more_owls[which.max(mean_grip_more_owls$f.4079.0.0), ]
mean_grip_def_owls[which.max(mean_grip_def_owls$f.4079.0.0), ]
```

```{r}
left <- ggplot(data = mean_grip_chronotype, aes(x = time_bins, y = f.4079.0.0, colour = chronotype)) +
  geom_line() +
  geom_point() +
  #scale_size_continuous(range = c(1, 5)) +  # Adjust the range of sizes as needed
  labs(x = "Time Bins", y = "Diastolic blood pressure") +
  ggtitle("Diastolic blood pressure by chronotype (mmHg)") +
  theme_minimal()

left
```

```{r}

right <- ggplot(data = mean_grip_chronotype, aes(x = time_bins, y = f.4080.0.0, colour = chronotype)) +
  geom_line() +
  geom_point() +
  #scale_size_continuous(range = c(1, 5)) +  # Adjust the range of sizes as needed
  labs(x = "Time Bins", y = "Systolic blood pressure") +
  ggtitle("Systolic blood pressure by chronotype") +
  theme_minimal()

right
```
```{r}

right <- ggplot(data = mean_grip_chronotype, aes(x = time_bins, y = f.102.0.0, colour = chronotype)) +
  geom_line() +
  geom_point() +
  #scale_size_continuous(range = c(1, 5)) +  # Adjust the range of sizes as needed
  labs(x = "Time Bins", y = "Pulse rate") +
  ggtitle("Pulse rate by chronotype") +
  theme_minimal()

right
```




