---
title: "Diabetic markers - Diabetes analysis"
author: "agpr141"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Install packages

```{r cars}
#library(tidyverse)
library(dplyr)
library(readr)
library(yarrr)
library(season)
```

#### Load in relevant data

```{r data load, echo=FALSE}
 # load in UKB_master
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\ukb_master.Rda")

#control exclusion vars
control_excl_vars <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\normative mets\\control_exclusion_participant.tsv")
control_excl_vars <- control_excl_vars %>% 
  rename('eid' = 'Participant ID')

og_df <- ukb_master
#rm(ukb_master)

# merge above 2 datasets into 1
diab_df <- merge(og_df,control_excl_vars, by="eid")

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\excluded_numbers.Rda")

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\excluded_icd.Rda")
```

#### Make new df

```{r}
# make new df which i will now work with

### remove people who shift/night shift work ####
diab1_df <- diab_df[!grepl("Prefer not to answer|Do not know|Sometimes|Usually|Always", diab_df$shift_work),] 
diab2_df <- diab1_df[!grepl("Prefer not to answer|Do not know|Sometimes|Usually|Always", diab1_df$night_shift),]

### remove people with probable single/recurrent major depression ####
diab3_df <- diab2_df[!grepl("Probable", diab2_df$`Bipolar and major depression status | Instance 0`),]

#keep in gestational diabetes for now - not sure what to do with this
#unique(bdmet_df$`Gestational diabetes only | Instance 0`)
#no_dm_2 <- no_dm[!grepl("Yes", no_dm$'Gestational diabetes only | Instance 0'),]

#merge icd search output variables with control_excl_vars ####
diab4_df <- merge(diab3_df,exclusions_icd, by="eid")

### CONTROL group: remove people who meet any of the exclusion criteria ####
control1_df <- diab4_df %>%
  filter(F20!='1' & F21!='1'& F22!='1' & F23!='1' & F24!='1' & F25!='1'& F27!='1' &
           F28!='1' &
           F30!='1' & F31!='1'& F32!='1'& F33!='1' & F34!='1' & F38.1!='1' & 
           F39!='1' & 
           E10!='0' | E11!='0' | E12!='0' | E13!='0' | E14!='0' &
           F50!='1'&
           E70_E72!='1'& E74!='1' & E75!='1' & E76!='1' & E77!='1' & 
           E78!='1' & E79!='1' & E80!='1' &
           G47.2!='1'& F51.2!='1')


#### CONTROL group: remove people with probable bipolar disorder 
control2_df <- control1_df[!grepl("Disorder", control1_df$'Bipolar and major depression status | Instance 0'),]

# remove poeple with no self-reported diabetes (can't figure out - to do later)
#control3_df <- control2_df[!grepl("No", control2_df$'Diabetes diagnosed by doctor #| Instance 0') | grepl("No", control2_df$'Ever had diabetes (Type I or Type II)')]

#control2_df[!grepl("No", control2_df$'Diabetes diagnosed by doctor | Instance 0'#),]
#control4_df <- control3_df[!grepl("Yes", control3_df$'Diabetes diagnosed by doctor | Instance 0'),]

### BIPOLAR group: remove people who meet any of the exclusion criteria ####
bipolar1_df <- diab4_df %>%
  filter(F20!='1' & F21!='1'& F22!='1' & F23!='1' & F24!='1' & F25!='1'& 
           F27!='1' & F28!='1' &
           F32!='1'& F33!='1' & F34!='1' & F38.1!='1' & F39!='1' & 
           E10!='0' | E11!='0' | E12!='0' | E13!='0' | E14!='0' &
           F50!='1'&
           E70_E72!='1'& E74!='1' & E75!='1' & E76!='1' & E77!='1' & 
           E78!='1' & E79!='1' & E80!='1' &
           G47.2!='1'& F51.2!='1')

# create new bipolar group of people who have either: ICD10 bipolar disorder (F31) or have probable bipolar (Pell & Smith criteria)
bipolar2_df <- dplyr::filter(bipolar1_df, grepl('Bipolar I Disorder|Bipolar II Disorder', `Bipolar and major depression status | Instance 0`)|bipolar1_df$F31!='0')


# SAVE GROUP DATAFRAMES ####
save(control2_df,file="control_group_diabetes.Rda")
save(bipolar2_df,file="bipolar_group_diabetes.Rda")

# Combine both group dataframes into one for future analysis purposes ####
#add new column to both with study group
control_nc <- control2_df %>%
  mutate(Group = 'control')
bipolar_nc <- bipolar2_df %>%
  mutate(Group = 'bipolar')

#join 2 dataframes
diabetes_df <- rbind(control_nc,bipolar_nc)
#save
save(diabetes_df,file="diabetes_df.Rda")
```

#### Load in extra variables for analysis

```{r}
fasting_time <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\fasting_time.tsv")

fasting_time <- fasting_time %>% 
  rename('fasting_time' = 'p74_i0')

diabetes_df <- merge(diabetes_df,fasting_time, by="eid")

townsend <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\townsend.tsv")

townsend <- townsend %>% 
  rename('townsend' = 'p22189')

diabetes_df <- merge(diabetes_df,townsend, by="eid")


sleep_activity <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\sleep_activity.tsv")

sleep_activity <- sleep_activity %>% 
  rename('sleep_duration_sr' = '1160-0.0',
         'sleep_avg_acti' = '40042-0.0',
         'sedentary_avg_acti' = '40043-0.0',
         'moderate_avg_acti' = '40045-0.0',
         'light_avg_acti' = '40044-0.0',
         'acti_data_qual' = '90015-0.0',
         'IPAQ_group' = '22032-0.0')

diabetes_df <- merge(diabetes_df,sleep_activity, by="eid")
```

#### Load in blood biomarkers and metabolites

```{r}
blood_biochem_df <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\blood_biochemistry.tsv")
blood_datetime <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\blood_datetime_3166_participant.tsv")


#rename tsv variable columns
blood_biochem_df <- blood_biochem_df %>% 
  rename('alanine_at' = '30620-0.0',
         'albumin' = '30600-0.0',
         'alkaline_phos' = '30610-0.0',
         'apolipo_a' = '30630-0.0',
         'apolipo_b' = '30640-0.0',
         'aspartate_at' = '30650-0.0',
         'crp' = '30710-0.0',
         'calcium' = '30680-0.0',
         'cholesterol' = '30690-0.0',
         'creatinine' = '30700-0.0',
         'cystatin_c' = '30720-0.0',
         'dir_bilirubin' = '30660-0.0',
         'gamma_gmt' = '30730-0.0',
         'glucose' = '30740-0.0',
         'g_haem' = '30750-0.0',
         'hdl_chol' = '30760-0.0',
         'igf_1' = '30770-0.0',
         'ldl_direct' = '30780-0.0',
         'lipoprotein_a' = '30790-0.0',
         'oestradiol' = '30800-0.0',
         'phosphate' = '30810-0.0',
         'rheum_factor' = '30820-0.0',
         'shbg' = '30830-0.0',
         'testosterone' = '30850-0.0',
         'tot_bilirubin' = '30840-0.0',
         'tot_protein' = '30860-0.0',
         'triglycerides' = '30870-0.0',
         'urate' = '30880-0.0',
         'urea' = '30670-0.0',
         'vit_d' = '30890-0.0',
         )

#add comments to variable to link with biobank datafield id

comment(blood_biochem_df$alanine_at) <- c("Datafield = 30620-0.0")
comment(blood_biochem_df$albumin) <- c("Datafield = 30600-0.0")
comment(blood_biochem_df$alkaline_phos) <- c("Datafield = 30610-0.0")
comment(blood_biochem_df$apolipo_a) <- c("Datafield = 30630-0.0")
comment(blood_biochem_df$apolipo_b) <- c("Datafield = 30640-0.0")
comment(blood_biochem_df$aspartate_at) <- c("Datafield = 30650-0.0")
comment(blood_biochem_df$crp) <- c("Datafield = 30710-0.0")
comment(blood_biochem_df$calcium) <- c("Datafield = 30680-0.0")
comment(blood_biochem_df$cholesterol) <- c("Datafield = 30690-0.0")
comment(blood_biochem_df$creatinine) <- c("Datafield = 30700-0.0")
comment(blood_biochem_df$cystatin_c) <- c("Datafield = 30720-0.0")
comment(blood_biochem_df$dir_bilirubin) <- c("Datafield = 30660-0.0")
comment(blood_biochem_df$gamma_gmt) <- c("Datafield = 30730-0.0")
comment(blood_biochem_df$glucose) <- c("Datafield = 30740-0.0")
comment(blood_biochem_df$g_haem) <- c("Datafield = 30750-0.0")
comment(blood_biochem_df$hdl_chol) <- c("Datafield = 30760-0.0")
comment(blood_biochem_df$igf_1) <- c("Datafield = 30770-0.0")
comment(blood_biochem_df$ldl_direct) <- c("Datafield = 30780-0.0")
comment(blood_biochem_df$lipoprotein_a) <- c("Datafield = 30790-0.0")
comment(blood_biochem_df$oestradiol) <- c("Datafield = 30800-0.0")
comment(blood_biochem_df$phosphate) <- c("Datafield = 30810-0.0")
comment(blood_biochem_df$rheum_factor) <- c("Datafield = 30820-0.0")
comment(blood_biochem_df$shbg) <- c("Datafield = 30830-0.0")
comment(blood_biochem_df$testosterone) <- c("Datafield = 30850-0.0")
comment(blood_biochem_df$tot_bilirubin) <- c("Datafield = 30840-0.0")
comment(blood_biochem_df$tot_protein) <- c("Datafield = 30860-0.0")
comment(blood_biochem_df$triglycerides) <- c("Datafield = 30870-0.0")
comment(blood_biochem_df$urate) <- c("Datafield = 30880-0.0")
comment(blood_biochem_df$urea) <- c("Datafield = 30670-0.0")
comment(blood_biochem_df$vit_d) <- c("Datafield = 30890-0.0")


# exclude markers that are not of interest to us
metabolites <- blood_biochem_df[c("eid", "alanine_at","albumin","apolipo_a", "apolipo_b", "crp", "cholesterol","creatinine", "cystatin_c", "dir_bilirubin", "gamma_gmt", "glucose", "g_haem", "hdl_chol", "ldl_direct", "lipoprotein_a", "phosphate", "rheum_factor", "tot_bilirubin", "triglycerides", "vit_d")]


# merge with existing bdmet_df
diabetes_df <- merge(diabetes_df,metabolites, by="eid")
diabetes2_df <- merge(diabetes_df,blood_datetime, by="eid")

diabetes2_df <- diabetes2_df %>% 
  rename('blood sample datetime' = '3166-0.0')

```

#### Get blood sample time data

```{r}
# load participant dataframe for analysis
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Metabolite Analyses\\analysis_met.Rda")

# get blood sample timings
bs_vars <- readr::read_tsv("C:/Users/arogusk2/OneDrive - University of Edinburgh./HELIOS-BD/Side Projects/BioBank Project/core dataset/Sample timings/blood_datetime_3166_participant.tsv")

#rename variable
bs_vars <- bs_vars %>% 
  rename('BS_datetime' = '3166-0.0')

# initialise columns
bs_vars$bs_year <- as.numeric(NA)
bs_vars$bs_month <- as.numeric(NA)
#bs_vars$bs_mday <- as.numeric(NA)
#bs_vars$bs_wday <- as.numeric(NA)
bs_vars$bs_time <- as.numeric(NA)

#create forloop to pull out year, month, time and day of week
for (i in 1:nrow(bs_vars)){
  row <- bs_vars[i,]
  
  #convert to POSIXlt
  lt <- as.POSIXlt(bs_vars$BS_datetime[i])
  
  #pull out year from data (lt format = # years since 1900)
  year <- 1900+lt$year
  bs_vars$bs_year[i] <- year
  
  # pull out month (zero indexed, add 1)
  month <- 1+lt$mon
  bs_vars$bs_month[i] <- month
  
  # pull out day of the month
  #mday <- lt$mday
  #bs_vars$bs_mday[i] <- mday
  
  # pull out day of the week
  #wday <- lt$wday
  #bs_vars$bs_wday[i] <- wday
  
  # pull out ToD
  time <- format(as.POSIXct(bs_vars$BS_datetime[i]), format = "%H:%M")
  bs_vars$bs_time[i] <- time
}

#merge bs_vars into main df
ghaem_intermediate_df <- merge(diabetes2_df,bs_vars, by="eid")
```

Tidy data

```{r}
# remove people with NA glycated haemoglobin
ghaem_df <- diabetes2_df %>%  filter(!is.na(g_haem))

# remove people with NA glucose
ghaem_df <- ghaem_df %>%  filter(!is.na(glucose))

#remove people with glycated haemoglobin levels outside of manufacturer's reportable range
ghaem_df<- ghaem_df %>%  filter(!g_haem>184)
ghaem_df<- ghaem_df %>%  filter(!g_haem<15)

# remove people with NA variables: bs_datetime (removes 2)
#ghaem_df<- ghaem_df %>%  filter(!is.na(BS_datetime))

# make season numeric
#ghaem_df$bs_season <- as.numeric(as.character(ghaem_df$bs_season))

# tidy time outdoors data to remove '-10', '-3', '-1' values:
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-10] <- 0
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-3] <- NA
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-1] <- NA

ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-10] <- 0
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-3] <- NA
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-1] <- NA

# tidy self-report sleep time to remove -3', '-1' values:
ghaem_df$sleep_duration[ghaem_df$sleep_duration ==-3] <- NA
ghaem_df$sleep_duration[ghaem_df$sleep_duration ==-1] <- NA


# recode ethnicity into 7 groups
library('plyr')
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer
ghaem_df$ethnicity_recode <- revalue(ghaem_df$ethnicity, c("British"="1", "Any other white background"="1", "White"="1", "Irish"="1", "Mixed"="2", "White and Black Caribbean"="2", "White and Black African"="2", "White and Asian"="2", "Any other mixed background"="2", "Asian or Asian British"="3", "Indian"="3", "Pakistani"="3", "Bangladeshi"="3", "Any other Asian background"="3", "Black or Black British"="4", "Caribbean"="4", "African"="4", "Any other Black background"="4", "Chinese"="5", "Other ethnic group"="6", "Do not know"="7", "Prefer not to answer"="7"))
#ghaem_df$ethnicity_recode <- as.numeric(ghaem_df$ethnicity_recode)

#ethnicity dummy vars code
#white
ghaem_df$ethnicity_white <- revalue(ghaem_df$ethnicity_recode, c("1"="1", "2"="0", "3"="0", "4"="0", "5"="0", "6"="0", "7"="0"))

#mixed
ghaem_df$ethnicity_mixed <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="1", "3"="0", "4"="0", "5"="0", "6"="0", "7"="0"))

#asian/asian british
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="1", "4"="0", "5"="0", "6"="0", "7"="0"))

#black/black british
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="1", "5"="0", "6"="0", "7"="0"))

#chinese
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="0", "5"="1", "6"="0", "7"="0"))

#other
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="0", "5"="0", "6"="1", "7"="0"))

# recode employment into 5 groups
#1=employed, 2=retired, 3=unable to work because of sickness/disability, 4=other unemployment, 5=do not know/prefer not to answer
ghaem_df$employment_recode <- revalue(ghaem_df$employed, c("In paid employment or self-employed"="1", "Retired"="2", "Doing unpaid or voluntary work"="4", "Unable to work because of sickness or disability"="3", "Full or part-time student"="4", "Unemployed"="4", "Looking after home and/or family"="4", "None of the above"="4", "Prefer not to answer"="5"))

#remove plyr package as has masking for later functions
detach("package:plyr", unload=TRUE)

# collapse bs_time into 1 day time variable - n=161 people with whacky time between 00:00 - 08:00
#ghaem_df$bs_time_collapsed <- as.POSIXct(ghaem_df$bs_time,format="%H:%M")

    #create variables containing sequence of times to exclude from data
    # pull out the data date 
    #start_time <- ymd_hms("2024-04-19 00:00:00")
    #end_time <- ymd_hms("2024-04-19 08:00:00")
    #exclude_dt_range <- seq(start_time, end_time, "min")
    #exclude_time_range <- format(as.POSIXct(exclude_dt_range,format='%Y-%m-%d %H:%M:%S'),format='%Y-%m-%d %H:%M:%S')
#exclude_time_range <- as.POSIXct(exclude_time_range)
    
    #remove rows where sample time is outside of 23:00:00-06:00:00 range
    #bloodtime_df <- ghaem_df %>%
    #  filter(!bs_time_collapsed %in% exclude_time_range)

#create new date var
#ghaem_df$bs_date <- as.Date(ghaem_df$BS_datetime)

#remove people with wrong datetime
#bs_time to integer
#library(data.table)
#ghaem_df$bs_itime <- as.ITime(ghaem_df$bs_time)
#ghaem_df$bs_itime <- as.integer(ghaem_df$bs_itime)
#ghaem_df<- ghaem_df %>%  filter(!bs_itime<1)
#detach("package:data.table", unload=TRUE) 
    
# make control and bipolar dataframes
#ungroup main df
ghaem_df = as.data.frame(ghaem_df)

#create groups
control_df <- ghaem_df %>%  filter(Group=='control')
bipolar_df <- ghaem_df %>%  filter(Group=='bipolar')
```

```{r recode for multivar lr}

library('plyr')

##Group
#control
ghaem_df$group <- revalue(ghaem_df$Group, c("control"="1", "bipolar"="2"))
ghaem_df$group_f <- factor(ghaem_df$group, levels= c("1", "2"), labels= c("control", "bipolar"), ordered=FALSE)

### LEVEL 1
#Sex
#sex
#ghaem_df$sex_recode <- revalue(ghaem_df$sex, c("Male"="1", "Female"="2"))
ghaem_df$sex_f <- factor(ghaem_df$sex, ordered=FALSE)

#ghaem_df$sex_fact <- factor(ghaem_df$sex_recode, levels= c("Male", "Female"), labels=c("male", "female"))

#ethnicity
# set values to NA if not relevant
ghaem_df$ethnicity_recode[ghaem_df$ethnicity_recode == -1] <- NA
ghaem_df$ethnicity_recode[ghaem_df$ethnicity_recode == -3] <- NA
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer
ghaem_df$ethnicity_f <- factor(ghaem_df$ethnicity_recode, levels= c("1", "2", "3", "4", "5", "6"), labels= c("white", "mixed", "asian", "black", "chinese", "other"), ordered=FALSE)

### LEVEL 2 
#smoking
# set values to NA if not relevant
ghaem_df$smoking_f <- ghaem_df$smoking_status
ghaem_df$smoking_f[ghaem_df$smoking_f == -3] <- NA
ghaem_df$smoking_f <- factor(ghaem_df$smoking_f, levels= c("Never", "Previous", "Current"), labels= c("Never", "Previous", "Current"), ordered=FALSE)

#alcohol - showing up strange!
#recode: never/special occasions=1 (LOW), 1-3month/1-2week=2 (MODERATE), 3-4week/daily=3 (HIGH)
ghaem_df$alcohol_f <- as.character(ghaem_df$alcohol_intake)
#ghaem_df$alcohol_f[ghaem_df$alcohol_f == 'Prefer not to answer'] <- NA
ghaem_df$alcohol_f <- revalue(ghaem_df$alcohol_f, c("Never"="Low", "Special occasions only"="Low", "One to three times a month"="Moderate", "Once or twice a week"="Moderate","Three or four times a week"="High", "Daily or almost daily"="High", "Prefer not to answer"=NA))
ghaem_df$alcohol_f <- factor(ghaem_df$alcohol_f, levels= c("Low", "Moderate", "High"), labels= c("Low", "Moderate", "High"), ordered=TRUE)

#health self-report
ghaem_df$health_f <- as.character(ghaem_df$health_self_report)
ghaem_df$health_f[ghaem_df$health_f == 'Prefer not to answer'] <- NA
ghaem_df$health_f[ghaem_df$health_f == 'Do not know'] <- NA
ghaem_df$health_f <- factor(ghaem_df$health_f, levels= c("Excellent", "Good", "Fair", "Poor"), labels= c("Excellent", "Good", "Fair", "Poor"), ordered=TRUE)


#disbility
ghaem_df$disability_f <- as.character(ghaem_df$disability_self_report)
ghaem_df$disability_f[ghaem_df$disability_f == 'Prefer not to answer'] <- NA
ghaem_df$disability_f[ghaem_df$disability_f == 'Do not know'] <- NA
ghaem_df$disability_f <- factor(ghaem_df$disability_f, levels= c("No", "Yes"), labels= c("No", "Yes"), ordered=FALSE)


#chronotype
ghaem_df$chrono_f <- as.character(ghaem_df$chronotype)
ghaem_df$chrono_f[ghaem_df$chrono_f == 'Prefer not to answer'] <- NA
ghaem_df$chrono_f[ghaem_df$chrono_f == 'Do not know'] <- NA
ghaem_df$chrono_f <- factor(ghaem_df$chrono_f, ordered=TRUE)

### Level 4 
#assess_centre to factor
ghaem_df$centre_f <- factor(ghaem_df$assess_centre, ordered=FALSE)

#remove plyr package as has masking for later functions
detach("package:plyr", unload=TRUE)

# re-make control and bipolar dataframes
#ungroup main df
ghaem_df = as.data.frame(ghaem_df)

#create groups
control_df <- ghaem_df %>%  filter(Group=='control')
bipolar_df <- ghaem_df %>%  filter(Group=='bipolar')
```

#### Run analysis to test HbA1C rhythm

### 8. Calculate cosinor and plot marker levels using pirate plot

```{r initialise pirate plot}

#colour gradient
#col <- inferno(12, begin=0.3, end=0.7, direction=1) 
#as.data.frame(col)

dayCOL<-c("#50576CFF", "#989278FF", "#ADA274FF", "#D8C563FF", "#EED654FF", "#FFEA46FF", "#EED654FF", "#D8C563FF", "#ADA274FF", "#858379FF", "#50576CFF", "#1B3B6DFF")


SeasonPal <- c("darkseagreen3","darkseagreen3","darkseagreen3","lightgoldenrod2","lightgoldenrod2","lightgoldenrod2","indianred3","indianred3","indianred3","lightsteelblue2","lightsteelblue2","lightsteelblue2") 

names <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
namesshort <-c("J","F","M","A","M","J","J","A","S","O","N","D")
```

#### [HbA1c - control cosinor]{.underline}

```{r HbA1c all control}

#run cosinor
hba1c_control_cos <- cosinor(g_haem~1 #+ control_df$age + control_df$BMI + control_df$sex_f + control_df$ethnicity_f
                             , date='assess_month', type='monthly',data=control_df)
control_cos_summary <- summary(hba1c_control_cos)
summary(hba1c_control_cos$glm)
mesor_control <- mean(hba1c_control_cos$glm$fitted.values) # for mesor
#plot(hba1c_control_cos)
#seasrescheck(hba1c_control_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(hba1c_control_cos$glm$data, select = hba1c_control_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control

pirateplot(formula =  g_haem ~ assess_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(45,50),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "HbA1c Seasonal Profile - Diabetic Control population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],hba1c_control_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], hba1c_control_cos$fitted.values[o_c], pch = 19)
legend(10, 60, legend=c("Spring", "Summer","Autumn", "Winter"),
       col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"), lty=1, lwd=5, cex=0.8)
```

#### HbA1c - matched cosinors

```{r matched cosinor}
#HbA1c - matched control cosinor
hba1c_mc_cos <- cosinor(g_haem~1 + mc_df$age + mc_df$BMI + mc_df$sex_f + mc_df$bs_itime + mc_df$ethnicity_f + mc_df$alcohol_f + mc_df$health_f, date='bs_month', type='monthly',data=mc_df)
mc3_cos_summary <- summary(hba1c_mc_cos)
summary(hba1c_mc_cos$glm)
mesor_mc <- mean(hba1c_mc_cos$glm$fitted.values) # for mesor

#plot(hba1c_mc_cos)
#seasrescheck(hba1c_mc_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_mc_3 = subset(hba1c_mc_cos$glm$data, select = hba1c_mc_cos$date)[, 1]
o_mc3 = order(time_fit_mc_3)

# HbA1c - control

pirateplot(formula =  g_haem ~ bs_month,
           data = mc_df,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(34.4,35),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_mc_3[o_mc3],hba1c_mc_cos$fitted.values[o_mc3], lty=2,lwd=1, col="black")
points(time_fit_mc_3[o_mc3], hba1c_mc_cos$fitted.values[o_mc3], pch = 19)



```

#### HbA1c - bipolar cosinor

```{r}
# HbA1c - bipolar

#run cosinor
hba1c_bipolar_cos <- cosinor(g_haem~1, #+ bipolar_df$age + bipolar_df$BMI + bipolar_df$sex_f + bipolar_df$bs_itime + bipolar_df$ethnicity_f + bipolar_df$alcohol_f + bipolar_df$health_f, 
                             date='assess_month', type='monthly', data=bipolar_df, cycles=2)
bd_cos_summary <- summary(hba1c_bipolar_cos)
summary(hba1c_bipolar_cos$glm)
mesor_bipolar <- mean(hba1c_bipolar_cos$glm$fitted.values) # for mesor
#plot(hba1c_bipolar_cos)
#seasrescheck(hba1c_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(hba1c_bipolar_cos$glm$data, select = hba1c_bipolar_cos$date)[, 1]
o_bd = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  g_haem ~ assess_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(32,36.5) # Adjust y axis limits
           ylim=c(38,60),
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "HbA1c Seasonal Profile - Diabetic Bipolar population"
           )
#axis.break()
#axis.break(0,38, style='zigzag')
axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o_bd],hba1c_bipolar_cos$fitted.values[o_bd], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o_bd], hba1c_bipolar_cos$fitted.values[o_bd], pch = 19)
legend(10, 60, legend=c("Spring", "Summer","Autumn", "Winter"),
       col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"), lty=1, lwd=5, cex=0.8)
```

#### HbA1c - plot all groups (year collapsed)

nts(control_phase, (hba1c_control_cos\$fitted.values[control_phase])+0.2, pch = 25, col="coral3") \# phase

```{r}
#run significance test between groups for: phase, nadir, amplitude, mesor
#using circacompare - alpha threshold set to 0.5 because bipolar group is non-rhythmic
#cos_stats_ghaem <- circacompare(x=ghaem_df, col_time="bs_month", col_group="Group"#, col_outcome="g_haem", period=12, alpha_threshold=0.5)

#pull out phase and nadir information as numeric
control_phase <- parse_number(control_cos_summary$phase)
control_nadir <- parse_number(control_cos_summary$lphase)
#mc3_phase <- parse_number(mc3_cos_summary$phase)
#mc3_nadir <- parse_number(mc3_cos_summary$lphase)
bd_phase <- parse_number(bd_cos_summary$phase)
bd_nadir <- parse_number(bd_cos_summary$lphase)

#create plot with fitted sinusoid
pirateplot(formula = g_haem ~ time_fit_control,
           data = control_df,
           theme = 0, # set theme to 0 for blank plot
           xlab = "Month", # add x axis label
           ylab = expression(bold(paste("HbA1c (mmol/mol)"))), # add y axis label
           xaxt = "n", # remove x axis check marks
           yaxt = "n", # remove y axis check markers
           ylim=c(20,50), # define y axis limits
           main = "HbA1c Seasonal Profile")

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
#add control
lines(time_fit_control[o_c],hba1c_control_cos$fitted.values[o_c], lty=2,lwd=1.2, col="#D05F5F") # line
points(time_fit_control[o_c], hba1c_control_cos$fitted.values[o_c], pch = 16, col="#D05F5F") # points for each month
points(control_phase, max(hba1c_control_cos$fitted.values)+0.05, pch =25, col="black", bg="#D05F5F", cex=1.3) # phase
text(control_phase, max(hba1c_control_cos$fitted.values)+0.05, sprintf("Phase: %s",control_phase), cex=0.7, pos=3, col="black") # add text to phase point
segments(x0=control_phase, x1=control_phase, y0=max(hba1c_control_cos$fitted.values), y1=(max(hba1c_control_cos$fitted.values))-control_cos_summary$amp, lty=2, lwd=1, col="#D05F5F") # add amplitude vertical line
text(control_phase, ((max(hba1c_control_cos$fitted.values))-control_cos_summary$amp), sprintf("Amplitude: %s",round(control_cos_summary$amp, digits=2)), cex=0.7, pos=1, offset=-0.05,col="black") # add text to phase point
text(1, hba1c_control_cos$fitted.values[1], "Control", cex=1, pos=3,col="#D05F5F") # add text to label line

#add matched control (3:1)
#lines(time_fit_mc_3[o_mc3],hba1c_mc_cos$fitted.values[o_mc3], lty=2,lwd=1.2, col="#EA6535") #line
#points(time_fit_mc_3[o_mc3], hba1c_mc_cos$fitted.values[o_mc3], pch = 16, col="#EA6535") #points for each month
#points(mc3_phase, max(hba1c_mc_cos$fitted.values)+0.05, pch =25, col="black", bg="#EA6535", cex=1.3) # phase
#text(mc3_phase, max(hba1c_mc_cos$fitted.values)+0.05, sprintf("Phase: %s", mc3_phase), cex=0.7, pos=3, col="black") # add text to phase point
#segments(x0=mc3_phase, x1=mc3_phase, y0=max(hba1c_mc_cos$fitted.values), y1=(max(hba1c_mc_cos$fitted.values))-mc3_cos_summary$amp, lty=2,lwd=1, col="#EA6535") # add amplitude vertical line
#text(mc3_phase, ((max(hba1c_mc_cos$fitted.values))-mc3_cos_summary$amp), sprintf("Amplitude: %s",round(mc3_cos_summary$amp, digits=2)), cex=0.7, pos=1, offset=-0.05,col="black") # add text to phase point
#text(1, hba1c_mc_cos$fitted.values[1], "Matched\ncontrol", cex=1, pos=3,col="#EA6535") # add text to label line

#add bipolar
lines(time_fit_bipolar[o_bd],hba1c_bipolar_cos$fitted.values[o_bd], lty=2,lwd=1.2, col="#196AA9")
points(time_fit_bipolar[o_bd], hba1c_bipolar_cos$fitted.values[o_bd], pch = 16, col="#196AA9")
points(bd_phase, max(hba1c_bipolar_cos$fitted.values)+0.05, pch =25, col="black", bg="#196AA9", cex=1.3) # phase
text(bd_phase, max(hba1c_bipolar_cos$fitted.values)+0.05, sprintf("Phase: %s",bd_phase), cex=0.7, pos=3, col="black") # add text to phase point
segments(x0=bd_phase, x1=bd_phase, y0=max(hba1c_bipolar_cos$fitted.values), y1=(max(hba1c_bipolar_cos$fitted.values))-bd_cos_summary$amp, lty=2,lwd=1, col="#196AA9") # add amplitude vertical line
text(bd_phase, ((max(hba1c_bipolar_cos$fitted.values))-bd_cos_summary$amp), sprintf("Amplitude: %s",round(bd_cos_summary$amp, digits=2)), cex=0.7, pos=1, offset=-0.05,col="black") # add text to phase point
text(1, hba1c_bipolar_cos$fitted.values[1], "Bipolar", cex=1, pos=3,col="#196AA9") # add text to label line

#legend(x="topright", legend=c("Control", "Matched Control", "Bipolar"), col=c("#D05F5F", "#EA6535", "#196AA9"), lty=1, cex=0.8)
```

#### [Glucose - control cosinor]{.underline}

```{r Glucose cosinor+plot}

#run cosinor
glu_control_cos <- cosinor(glucose~1 #+ control_df$age + control_df$BMI + control_df$sex_f + control_df$bs_itime + control_df$ethnicity_f + control_df$alcohol_f + control_df$health_f
                           , date='assess_month', type='monthly', data=control_df)
glu_control_summary <- summary(glu_control_cos)
#summary(glu_control_cos$glm)
#glu_control_mesor <- mean(glu_control_cos$glm$fitted.values)
#plot(glucose_control_cos)
#seasrescheck(glucose_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
glu_time_fit_control = subset(glu_control_cos$glm$data, select = glu_control_cos$date)[, 1]
glu_o_control = order(glu_time_fit_control)

# Glucose - control
dev.new()
pirateplot(formula =  glucose ~ assess_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(4.85,5),  # Adjust y axis limits
           ylim=c(4,9),
           main = "Glucose Seasonal Profile - Diabetic Control population"
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(glu_time_fit_control[glu_o_control],glu_control_cos$fitted.values[glu_o_control], lty=2,lwd=1, col="black")
points(glu_time_fit_control[glu_o_control], glu_control_cos$fitted.values[glu_o_control], pch = 19)
legend(10, 9, legend=c("Spring", "Summer","Autumn", "Winter"),
       col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"), lty=1, lwd=5, cex=0.8)       


```

#### Glucose - matched cosinors

```{r glucose matched control cosinor}
## glucose - matched control cosinor 3:1
glu_mc3_cos <- cosinor(glucose~1+ mc_df$age + mc_df$BMI + mc_df$sex_f + mc_df$bs_itime + mc_df$ethnicity_f + mc_df$alcohol_f + mc_df$health_f, date='bs_month', type='monthly', data=mc_df)
glu_mc3_summary <- summary(glu_mc3_cos)
#summary(glu_mc3_cos$glm)
#glu_control_mesor <- mean(glu_mc_cos$glm$fitted.values)
#plot(glucose_mc_cos)
#seasrescheck(glucose_mc_cos$residuals) # check the residuals

glu_time_fit_mc3 = subset(glu_mc3_cos$glm$data, select = glu_mc3_cos$date)[, 1]
glu_o_mc3 = order(glu_time_fit_mc3)

#1:1
#glucose_mc1_cos <- cosinor(glucose~1, date='bs_month', type='monthly', data#=mc_df_1)
#summary(glucose_mc1_cos)
#summary(glucose_mc1_cos$glm)
#mesor_control1 <- mean(glucose_mc1_cos$glm$fitted.values)

#create new vars for plotting fitted sinusoid from a Cosinor object


# Glucose - matched control
dev.new()
pirateplot(formula =  glucose ~ bs_month,
           data = mc_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(4.85,5),  # Adjust y axis limits
           ylim=c(4.5,5.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_mc[o],glucose_mc_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_mc[o], glucose_mc_cos$fitted.values[o], pch = 19)
       
```

#### Glucose - bipolar cosinor

```{r glucose bipolar cosinor}
# Glucose - bipolar

#run cosinor
glu_bipolar_cos <- cosinor(glucose~1,#+ bipolar_df$age + bipolar_df$BMI + bipolar_df$sex_f + bipolar_df$bs_itime + bipolar_df$ethnicity_f + bipolar_df$alcohol_f + bipolar_df$health_f, 
                           date='assess_month', type='monthly', data=bipolar_df, cycles=1)
glu_bd_summary <- summary(glu_bipolar_cos)
#summary(glu_bipolar_cos$glm)
#glu_bd_mesor <- mean(glu_bipolar_cos$glm$fitted.values)
#plot(glucose_bipolar_cos)
#seasrescheck(glucose_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
glu_time_fit_bipolar = subset(glu_bipolar_cos$glm$data, select = glu_bipolar_cos$date)[, 1]
glu_o_bipolar = order(glu_time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  glucose ~ assess_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(4.5,5.5) # Adjust y axis limits
           ylim=c(4,9),
           main = "Glucose Seasonal Profile - Diabetic Bipolar population"
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(glu_time_fit_bipolar[glu_o_bipolar],glu_bipolar_cos$fitted.values[glu_o_bipolar], lty=2,lwd=1, col="black")
points(glu_time_fit_bipolar[glu_o_bipolar], glu_bipolar_cos$fitted.values[glu_o_bipolar], pch = 19)
legend(10, 9, legend=c("Spring", "Summer","Autumn", "Winter"),
       col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"), lty=1, lwd=5, cex=0.8)

#run significance test between groups for: phase, nadir, amplitude, mesor
#using circacompare - alpha threshold set to 0.7 because bipolar group is non-rhythmic
cos_stats_glucose <- circacompare(x=ghaem_df, col_time="bs_month", col_group="Group", col_outcome="glucose", period=12, alpha_threshold=0.7)
```

#### Glucose - plot all groups

```{r glucose all-group plot}
#pull out phase and nadir information as numeric
glu_control_phase <- parse_number(glu_control_summary$phase)
#glu_mc3_phase <- parse_number(glu_mc3_summary$phase)
glu_bd_phase <- parse_number(glu_bd_summary$phase)

#create plot with fitted sinusoid
pirateplot(formula = glucose ~ glu_time_fit_control,
           data = control_df,
           theme = 0, # set theme to 0 for blank plot
           xlab = "Month", # add x axis label
           ylab = expression(bold(paste("Glucose (mmol/L)"))), # add y axis label
           xaxt = "n", # remove x axis check marks
           yaxt = "n", # remove y axis check markers
           ylim=c(4,9), # define y axis limits
           main = "Glucose Seasonal Profile")

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
#add control
lines(glu_time_fit_control[glu_o_control],glu_control_cos$fitted.values[glu_o_control], lty=2,lwd=1.2, col="#D05F5F") # line
points(glu_time_fit_control[glu_o_control], glu_control_cos$fitted.values[glu_o_control], pch = 16, col="#D05F5F") # points for each month
points(glu_control_phase, max(glu_control_cos$fitted.values)+0.02, pch =25, col="black", bg="#D05F5F", cex=1.3) # phase
text(glu_control_phase, max(glu_control_cos$fitted.values)+0.02, sprintf("Phase: %s",glu_control_phase), cex=0.7, pos=3, col="black") # add text to phase point
segments(x0=glu_control_phase, x1=glu_control_phase, y0=max(glu_control_cos$fitted.values), y1=(max(glu_control_cos$fitted.values))-glu_control_summary$amp, lty=2, lwd=1, col="#D05F5F") # add amplitude vertical line
text(glu_control_phase, ((max(glu_control_cos$fitted.values))-glu_control_summary$amp), sprintf("Amplitude: %s",round(glu_control_summary$amp, digits=2)), cex=0.7, pos=1, offset=-0.05,col="black") # add text to phase point
text(1, glu_control_cos$fitted.values[1]+0.02, "Control", cex=1, pos=3,col="#D05F5F") # add text to label line

#add matched control (3:1)
#lines(glu_time_fit_mc3[glu_o_mc3],glu_mc3_cos$fitted.values[glu_o_mc3], lty=2,lwd=1.2, col="#EA6535") #line
#points(glu_time_fit_mc3[glu_o_mc3], glu_mc3_cos$fitted.values[glu_o_mc3], pch = 16, col="#EA6535") #points for each month
#points(glu_mc3_phase, max(glu_mc3_cos$fitted.values)+0.02, pch =25, col="black", bg="#EA6535", cex=1.3) # phase
#text(glu_mc3_phase, max(glu_mc3_cos$fitted.values)+0.02, sprintf("Phase: %s", glu_mc3_phase), cex=0.7, pos=3, col="black") # add text to phase point
#segments(x0=glu_mc3_phase, x1=glu_mc3_phase, y0=max(glu_mc3_cos$fitted.values), y1=(max(glu_mc3_cos$fitted.values))-glu_mc3_summary$amp, lty=2,lwd=1, col="#EA6535") ## add amplitude vertical line
#text(glu_mc3_phase, ((max(glu_mc3_cos$fitted.values))-glu_mc3_summary$amp), sprintf("Amplitude: %s",round(glu_mc3_summary$amp, digits=2)), cex=0.7, pos=1, offset=-0.05,col="black") # add text to phase point
#text(1, glu_mc3_cos$fitted.values[1], "Matched\ncontrol", cex=1, pos=3,col="#EA6535") # add text to label line

#add bipolar
lines(glu_time_fit_bipolar[glu_o_bipolar],glu_bipolar_cos$fitted.values[glu_o_bipolar], lty=2,lwd=1.2, col="#196AA9")
points(glu_time_fit_bipolar[glu_o_bipolar], glu_bipolar_cos$fitted.values[glu_o_bipolar], pch = 16, col="#196AA9")
points(glu_bd_phase, max(glu_bipolar_cos$fitted.values)+0.02, pch =25, col="black", bg="#196AA9", cex=1.3) # phase
text(glu_bd_phase, max(glu_bipolar_cos$fitted.values)+0.02, sprintf("Phase: %s",glu_bd_phase), cex=0.7, pos=3, col="black") # add text to phase point
segments(x0=glu_bd_phase, x1=glu_bd_phase, y0=max(glu_bipolar_cos$fitted.values), y1=(max(glu_bipolar_cos$fitted.values))-glu_bd_summary$amp, lty=2,lwd=1, col="#196AA9") # add amplitude vertical line
text(glu_bd_phase, ((max(glu_bipolar_cos$fitted.values))-glu_bd_summary$amp), sprintf("Amplitude: %s",round(glu_bd_summary$amp, digits=2)), cex=0.7, pos=1, offset=-0.05,col="black") # add text to phase point
text(1, glu_bipolar_cos$fitted.values[1], "Bipolar", cex=1, pos=3,col="#196AA9") # add text to label line
```

#### 
