---
title: "UKB_ghaem"
author: "agpr141"
date: "2024-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BD Seasonality & glycated haemoglobin
##load data - section complete

```{r load, echo=FALSE}
#load libraries
library(tidyverse)
library(dplyr)
library(readr)
library(season)
library(MatchIt)
library(ggplot2)

# load participant dataframe for analysis
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Final study participant dataframe\\analysis_df.Rda")

```

## Tidy data for Glycated Haemoglobin analysis - section complete

Remove:
* People with NA for glycated haemoglobin
* People with glycated haemoglobin levels >200mmol/mol

```{r pressure, echo=FALSE}
# remove people with NA glycated haemoglobin
ghaem_df <- study_population %>%  filter(!is.na(g_haem))

#remove people with glycated haemoglobin levels outside of manufacturer's reportable range
ghaem_df<- ghaem_df %>%  filter(!g_haem>184)
ghaem_df<- ghaem_df %>%  filter(!g_haem<15)

# remove people with NA variables: bs_datetime (removes 2)
ghaem_df<- ghaem_df %>%  filter(!is.na(BS_datetime))

```

## Create matched control group - section complete

1:1 matching based on sex, age, ethnicity, BMI, assess_centre, bs_month, bs_year

``` {r match}
# convert matching variables to type 'factor' for matchit function to work

for_match_df <- ghaem_df
for_match_df$Group<- as.factor(as.character(for_match_df$Group))

# remove people with NA variables: ethnicity
for_match_df<- for_match_df %>%  filter(!is.na(ethnicity))
#for_match_df<- for_match_df %>%  filter(!is.na(BMI))

m.out1 <- matchit(Group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year, data = for_match_df)

match_control_df <- match.data(m.out1)
```

## Table 1 - section in progress

```{r table 1}
# summary statistics - mean for continuous variables:
demo_mean_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales'),
               mean, na.rm=TRUE)

# summary statistics - std.dev for continuous variables:
demo_sd_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales'),
               sd, na.rm=TRUE)

#counts for discrete variables:
demo_sex_df <- ghaem_df %>%
  group_by(Group) %>%
  #mutate(sex = factor(sex)) %>% 
  count(sex)
demo_ethnicity_df <- ghaem_df %>%
  group_by(Group) %>%
  #mutate(sex = factor(sex)) %>% 
  count(ethnicity)
demo_disabled_df <- ghaem_df %>%
  group_by(Group) %>%
  #mutate(sex = factor(sex)) %>% 
  count(disability_self_report)    
```


## Explore blood sample timings - section in progress

```{r explore sample timings/distribution}

# month
ggplot(ghaem_df, aes(x = bs_month)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# day
ggplot(ghaem_df, aes(x = bs_wday)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")
```

## Explore glycated haemoglobin - section in progress

