---
title: "UKB_ghaem"
author: "agpr141"
date: "2024-03-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BD Seasonality & glycated haemoglobin

## Import libs & load data - section complete

```{r load, echo=FALSE, verbose=FALSE}
#load libraries
library(tidyverse)
library(dplyr)
library(readr)
library(season)
library(MatchIt)
library(ggplot2)
library(lubridate)
library(broom)
library(hrbrthemes)
library(viridis)
library(yarrr)
library(ggpubr)

# load participant dataframe for analysis
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Final study participant dataframe\\analysis_df.Rda")

```

## Tidy data for diabetic marker analysis - section complete

Remove:

-   People with NA for glycated haemoglobin

-   People with glycated haemoglobin levels outside of analytic
    reference range 15-184mmol/mol

-   People with glycated haemoglobin levels \>48mmol/mol (diagnostic
    level for diabetes)

Make:

-   New outlier variable for glycated haemoglobin

-   Season variable from month data (1=Spring, 2=Summer, 3=Autumn,
    4=Winter)

-   Tidy 'time_outdoors...' variables to remove '-10', '-3', '-1' values

```{r pressure, echo=FALSE}
# remove people with NA glycated haemoglobin
ghaem_df <- analysis_df %>%  filter(!is.na(g_haem))

#remove people with glycated haemoglobin levels outside of manufacturer's reportable range
ghaem_df<- ghaem_df %>%  filter(!g_haem>184)
ghaem_df<- ghaem_df %>%  filter(!g_haem<15)

#remove people with glycated haemoglobin levels above 48mmol/ml (diabetes diagnostic criteria)
ghaem_df<- ghaem_df %>%  filter(!g_haem>48)

# add new column to df to indicate if ghaem value is IQR-defined outlier
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

ghaem_df <- ghaem_df %>%
        group_by(Group) %>%
        mutate(outlier = ifelse(find_outlier(g_haem), g_haem, NA))

# remove people with NA variables: bs_datetime (removes 2)
ghaem_df<- ghaem_df %>%  filter(!is.na(BS_datetime))

# make season numeric
ghaem_df$bs_season <- as.numeric(as.character(ghaem_df$bs_season))

# tidy time outdoors data to remove '-10', '-3', '-1' values:
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-10] <- 0
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-3] <- NA
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-1] <- NA

ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-10] <- 0
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-3] <- NA
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-1] <- NA

```

## Create matched control group - section complete

1:1 matching based on sex, age, ethnicity, BMI, assess_centre, bs_month,
bs_year

CURRENT DEFAULT: do not run this code section. for optional future use
(sensitivity analyses)

```{r match, eval=FALSE, include=FALSE, verbose=FALSE}
# convert matching variables to type 'factor' for matchit function to work

for_match_df <- ghaem_df
for_match_df$Group<- as.factor(as.character(for_match_df$Group))

# remove people with NA variables: ethnicity
for_match_df<- for_match_df %>%  filter(!is.na(ethnicity))
#for_match_df<- for_match_df %>%  filter(!is.na(BMI))

#use 1:1 nearest neighbour matching (Matchit default) without replacement on a propensity score estimated using a
# logistic regression of the treatment on the covariates
m.out1 <- matchit(Group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year, data = for_match_df)

match_control_df <- match.data(m.out1)

#try different matching method: exact
m.out_exact <- matchit(Group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year, data = for_match_df, method="nearest", exact= ~ bs_month + age + sex + assess_centre)
match_exact_df <- match.data(m.out_exact)
```

## Table 1 - section complete

-   Compute counts for discrete variables, mean+sd for continous
    variables

-   code for basic statistical tests

```{r table 1}


# summary statistics - median for continuous variables:
demo_mean_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter'),
               mean, na.rm=TRUE)

# summary statistics - std.dev for continuous variables:
demo_sd_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter'),
               sd, na.rm=TRUE)

#counts for discrete variables:
demo_sex_df <- ghaem_df %>%
  group_by(Group) %>%
    #mutate(sex = factor(sex)) %>% 
  count(sex)
demo_ethnicity_df <- ghaem_df %>%
  group_by(Group) %>%
  count(ethnicity)
demo_disabled_df <- ghaem_df %>%
  group_by(Group) %>%
  count(disability_self_report)    
demo_chronotype_df <- ghaem_df %>%
  group_by(Group) %>%
  count(chronotype)    
demo_employment_df <- ghaem_df %>%
  group_by(Group) %>%
  count(employed) 
demo_smoking_df <- ghaem_df %>%
  group_by(Group) %>%
  count(smoking_status)  
demo_alcohol_df <- ghaem_df %>%
  group_by(Group) %>%
  count(alcohol_intake)  

# stats tests - chi square for categorical
sex_chi <- chisq.test(x = table(ghaem_df$Group,
                                        ghaem_df$sex))
yarrr::apa(sex_chi)# print stats test result in legible apa format

ethnicity_chi <- chisq.test(x = table(ghaem_df$Group,
                                        ghaem_df$ethnicity))
sex_chi <- chisq.test(x = table(ghaem_df$Group,
                                        ghaem_df$sex))

# stats tests - t-test for parametric continous
age_ttest <- t.test(ghaem_df$age ~ ghaem_df$Group)
yarrr::apa(age_ttest)

# stats tests - wilcoxon rank sum for non-parametric continous
age_mw <- wilcox.test(age ~ Group, data = ghaem_df, exact = FALSE)
yarrr::apa(age_mw)
```

## Explore blood sample timings - section in progress

```{r explore sample timings/distribution}

# season
ggplot(ghaem_df, aes(x = bs_season)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")


# month
ggplot(ghaem_df, aes(x = bs_month)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# day
ggplot(ghaem_df, aes(x = bs_wday)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time - edit time var from char to hm (lubridate)
ghaem_df$bs_time <- hm(ghaem_df$bs_time)

ggplot(ghaem_df, aes(x = bs_time)) +
  geom_histogram(fill = "salmon2", colour = "black", binwidth=10) +
  facet_grid(Group ~ ., scales = "free")

```

## Explore diabetic markers - section in progress

-   Glycated haemoglobin levels are reportedly higher in cooler months,
    and lower in warmer months. is this true for both our control
    groups?

    -   calculate mean + sd + CI for HbA1c levels for months
        (individually) and seasons

    -   correlation between temperature & Hba1c level - months

    -   correlation between (avg) daylength and (avg) rate of change &
        Hba1c levels - month

```{r gly_haem}

# task 1: visualise with data collapsed into months & seasons

#HBA1C
#months
ghaem_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), sd = sd(g_haem, na.rm=TRUE), n = length(g_haem)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
ghaem_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), sd = sd(g_haem, na.rm=TRUE), n = length(g_haem)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#GLUCOSE
#months
glucose_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(glucose, na.rm=TRUE), sd = sd(glucose, na.rm=TRUE), n = length(glucose)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
glucose_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(glucose, na.rm=TRUE), sd = sd(glucose, na.rm=TRUE), n = length(glucose)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#LACTATE
#months
lactate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(lactate, na.rm=TRUE), sd = sd(lactate, na.rm=TRUE), n = length(lactate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
lactate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(lactate, na.rm=TRUE), sd = sd(lactate, na.rm=TRUE), n = length(lactate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#PYRUVATE
#months
pyruvate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(pyruvate, na.rm=TRUE), sd = sd(pyruvate, na.rm=TRUE), n = length(pyruvate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
pyruvate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(pyruvate, na.rm=TRUE), sd = sd(pyruvate, na.rm=TRUE), n = length(pyruvate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#CITRATE
#months
citrate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(citrate, na.rm=TRUE), sd = sd(citrate, na.rm=TRUE), n = length(citrate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
citrate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(citrate, na.rm=TRUE), sd = sd(citrate, na.rm=TRUE), n = length(citrate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

# calculate mean & plot glycated haem - matched controls
#ghaem_month_matched <- match_exact_df %>%
#  group_by(bs_month, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))

#matched control seasons
#ghaem_season_matched <- match_exact_df %>%
#  group_by(bs_season, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))


#if wanting to calculate median + IQR
#ghaem_month <- ghaem_df %>%
#  group_by(bs_month, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))

```

Explore correlation between diabetic marker levels and different
variables

```{r}
# seasonal variables

range(ghaem_df$lactate, na.rm=TRUE)  # useful code to determine ranges for ylims

#1. multi-plot: temperature and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=average_temp, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(15, 48) #ylim for plotting points
  ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=average_temp, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(1.5, 14.5) #ylim for plotting points
  ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=average_temp, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0.5, 13) #ylim for plotting points
  ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=average_temp, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0, 2) #ylim for plotting points
  ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=average_temp, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0, 0.3) #ylim for plotting points
  ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure


model <- ghaem_df %>%
  filter(Group=="control") %>%
  lm(g_haem~average_temp)

temp_ghaem <- cor.test(ghaem_df$average_temp, ghaem_df$g_haem, method=c("pearson", "kendall", "spearman"))
```

plotting

```{r plotting}
 
#plot mean values plus 95%CI
ggplot(citrate_season, aes(bs_season, mean, color=Group))+
  geom_line()+ggtitle("citrate seasonal Profile")+
  geom_point()+
  geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), width=.2,
                position=position_dodge(0.05))+
  scale_x_continuous("bs_season", labels = as.character(citrate_season$bs_season), breaks = citrate_season$bs_season)

# plot temp correlation
ggplot(ghaem_df, aes(x=average_temp, y=g_haem, color=Group)) + 
  geom_point()+
  geom_smooth(method=lm)+  
  ylim(20, 50)
model <- ghaem_df %>%
  filter(Group=="control") %>%
  lm(g_haem~average_temp)
temp_ghaem <- cor.test(ghaem_df$average_temp, ghaem_df$g_haem, method=c("pearson", "kendall", "spearman"))
```

correlations between different variables

```{r}

```

### test sig

xyzxyz

```{r ghaem stats, echo=FALSE}

# Independent samples t-test: assumptions
# 1. dependent variable is continuous
# 2. independent variable contains 2 categorical independent groups
# 3. observations are independent e.g. no relationship between observations within the group and between the groups themselves
# 4. no significant outliers in the data
# box plot to visualise

x = factor(ghaem_df$bs_season)  # make bs_season factor variable

#boxplot for groups with outliers labelled
ggplot(ghaem_df) +
  aes(x = x, y = g_haem) +
  geom_boxplot(fill = "steelblue") +
  geom_text(aes(label=outlier), na.rm=TRUE, hjust=-.5)+
  facet_grid(Group ~ ., scales = "free") +
  theme_minimal()

# 5. dependent variable data (&residuals) is (approximately) normally distributed for each group
#for both tests - if p>0.05 data is normally distributed, use parametric test. if p<0.05, use non-parametric test.
ggplot(ghaem_df, aes(x = age)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# if n<2000 - Shapiro Wilk
sw_normality <- ghaem_df %>%
  group_by(Group) %>%
  do(tidy(shapiro.test(ghaem_df$g_haem)))

#if n>2000 - Kolmogorov-Smirnov
ks_normality <- ghaem_df %>%
  group_by(Group) %>%
  do(tidy(ks.test(ghaem_df$g_haem, "pnorm", mean(ghaem_df$g_haem),sd(ghaem_df$g_haem))))

# 6. homogeneity of variance - variance for each group is the same
hom_o_var <- var.test(ghaem_df$g_haem ~ ghaem_df$Group, ghaem_df)
```
