---
title: "UKB_ghaem"
author: "agpr141"
date: "2024-03-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BD Seasonality & glycated haemoglobin
##load data - section complete

```{r load, echo=FALSE, verbose=FALSE}
#load libraries
library(tidyverse)
library(dplyr)
library(readr)
library(season)
library(MatchIt)
library(ggplot2)

# load participant dataframe for analysis
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Final study participant dataframe\\analysis_df.Rda")

```

## Tidy data for Glycated Haemoglobin analysis - section complete

Remove:
* People with NA for glycated haemoglobin
* People with glycated haemoglobin levels >200mmol/mol

```{r pressure, echo=FALSE}
# remove people with NA glycated haemoglobin
ghaem_df <- analysis_df %>%  filter(!is.na(g_haem))

#remove people with glycated haemoglobin levels outside of manufacturer's reportable range
ghaem_df<- ghaem_df %>%  filter(!g_haem>184)
ghaem_df<- ghaem_df %>%  filter(!g_haem<15)

# remove people with NA variables: bs_datetime (removes 2)
ghaem_df<- ghaem_df %>%  filter(!is.na(BS_datetime))

```

## Create matched control group - section complete

1:1 matching based on sex, age, ethnicity, BMI, assess_centre, bs_month, bs_year

``` {r match, verbose=FALSE}
# convert matching variables to type 'factor' for matchit function to work

for_match_df <- ghaem_df
for_match_df$Group<- as.factor(as.character(for_match_df$Group))

# remove people with NA variables: ethnicity
for_match_df<- for_match_df %>%  filter(!is.na(ethnicity))
#for_match_df<- for_match_df %>%  filter(!is.na(BMI))

#use 1:1 nearest neighbour matching (Matchit default) without replacement on a propensity score estimated using a
# logistic regression of the treatment on the covariates
m.out1 <- matchit(Group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year, data = for_match_df)

match_control_df <- match.data(m.out1)

#try different matching method: exact
m.out_exact <- matchit(Group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year, data = for_match_df, method="nearest", exact= ~ bs_month + age + sex + assess_centre)
match_exact_df <- match.data(m.out_exact)
```

## Table 1 - section in progress

```{r table 1}
# summary statistics - mean for continuous variables:
demo_mean_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales'),
               mean, na.rm=TRUE)

# summary statistics - std.dev for continuous variables:
demo_sd_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales'),
               sd, na.rm=TRUE)

#counts for discrete variables:
demo_sex_df <- ghaem_df %>%
  group_by(Group) %>%
  #mutate(sex = factor(sex)) %>% 
  count(sex)
demo_ethnicity_df <- ghaem_df %>%
  group_by(Group) %>%
  #mutate(sex = factor(sex)) %>% 
  count(ethnicity)
demo_disabled_df <- ghaem_df %>%
  group_by(Group) %>%
  #mutate(sex = factor(sex)) %>% 
  count(disability_self_report)    
```


## Explore blood sample timings - section in progress

```{r explore sample timings/distribution}

# month
ggplot(ghaem_df, aes(x = bs_month)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# day
ggplot(ghaem_df, aes(x = bs_wday)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time - need to edit time var from char to iTime or similar

ggplot(ghaem_df, aes(x = bs_time)) +
  geom_histogram(fill = "salmon2", colour = "black", binwidth=10) +
  facet_grid(Group ~ ., scales = "free")

```

## Explore glycated haemoglobin - section in progress
Glycated haemoglobin levels are reportedly higher in cooler months, and lower in warmer months.
1. is this true for both our control groups?
    * calculate mean + sd for HbA1c levels for months (individually) and seasons
    * correlation between temperature & Hba1c level - months
    * correlation between (avg) daylength and (avg) rate of change & Hba1c levels - month

```{r gly_haem}

# task 1: visualise with data collapsed into months - all participants

# calculate mean & plot glycated haem - all controls
#individual months
g_haem_summary <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#seasons
#spring
g_haem_spring <- ghaem_df %>%
  filter(bs_month == "3" |bs_month =="4" |bs_month == "5") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#summer
g_haem_summer <- ghaem_df %>%
  filter(bs_month == "6" |bs_month =="7" |bs_month == "8") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#autumn
g_haem_autumn <- ghaem_df %>%
  filter(bs_month == "9" |bs_month =="10" |bs_month == "11") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#winter
g_haem_winter <- ghaem_df %>%
  filter(bs_month == "12" |bs_month =="1" |bs_month == "2") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))


# calculate mean & plot glycated haem - matched controls
g_haem_summary_matched <- match_exact_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#matched control seasons
#spring
g_haem_m_spring <- match_exact_df %>%
  filter(bs_month == "3" |bs_month =="4" |bs_month == "5") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#summer
g_haem_m_summer <- match_exact_df %>%
  filter(bs_month == "6" |bs_month =="7" |bs_month == "8") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#autumn
g_haem_m_autumn <- match_exact_df %>%
  filter(bs_month == "9" |bs_month =="10" |bs_month == "11") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

#winter
g_haem_m_winter <- match_exact_df %>%
  filter(bs_month == "12" |bs_month =="1" |bs_month == "2") %>%
  group_by(Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), SD = sd(g_haem, na.rm=TRUE), N = length(g_haem))

 
#plot
ggplot(g_haem_summary, aes(bs_month, mean, color=Group))+
  geom_line()+ggtitle("glyc.haem Annual Profile")+
  geom_point()+
  geom_errorbar(aes(ymin=mean-SD, ymax=mean+SD), width=.2,
                position=position_dodge(0.05))+
  scale_x_continuous("bs_month", labels = as.character(g_haem_summary$bs_month), breaks = g_haem_summary$bs_month)

# plot temp correlation
ggplot(match_exact_df, aes(x=average_temp, y=g_haem, color=Group)) + 
  geom_point()+
  geom_smooth(method=lm)+  
  ylim(20, 50)
model <- ghaem_df %>%
  filter(Group=="control") %>%
  lm(g_haem~average_temp)
temp_ghaem <- cor.test(ghaem_df$average_temp, ghaem_df$g_haem, method=c("pearson", "kendall", "spearman"))
```
