---
title: "UKB_diabetic"
author: "agpr141"
date: "2024-03-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, verbose=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

# BD Seasonality & Diabetic Biomarkers

### 1: Import libs

```{r load, echo=FALSE, verbose=TRUE}
#load libraries
library(tidyverse)
library(dplyr)
library(readr)
library(MatchIt)
library(ggplot2)
library(lubridate)
library(broom)
library(hrbrthemes)
library(viridis)
library(yarrr)
library(ggpubr)
library(sjstats)
library(plotrix)
library(segmented)
library(plot3D)
library(season)
```

### 2: Load data

```{r load data, echo=FALSE, verbose=TRUE}

# load participant dataframe for analysis
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Final study participant dataframe\\analysis_df.Rda")

fasting_time <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\fasting_time.tsv")

fasting_time <- fasting_time %>% 
  rename('fasting_time' = 'p74_i0')

analysis_df <- merge(analysis_df,fasting_time, by="eid")

townsend <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\townsend.tsv")

townsend <- townsend %>% 
  rename('townsend' = 'p22189')

analysis_df <- merge(analysis_df,townsend, by="eid")


sleep_activity <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\sleep_activity.tsv")

sleep_activity <- sleep_activity %>% 
  rename('sleep_duration_sr' = '1160-0.0',
         'sleep_avg_acti' = '40042-0.0',
         'sedentary_avg_acti' = '40043-0.0',
         'moderate_avg_acti' = '40045-0.0',
         'light_avg_acti' = '40044-0.0',
         'acti_data_qual' = '90015-0.0',
         'IPAQ_group' = '22032-0.0')

analysis_df <- merge(analysis_df,sleep_activity, by="eid")

```

### 3: Tidy data

Remove:

-   People with NA for glycated haemoglobin

-   People with NA for glucose

-   People with glycated haemoglobin levels outside of analytic
    reference range 15-184mmol/mol

-   People with glycated haemoglobin levels \>48mmol/mol (diagnostic
    level for diabetes)

Make:

-   New outlier variable for glycated haemoglobin

-   Season variable from month data (1=Spring, 2=Summer, 3=Autumn,
    4=Winter)

-   Tidy 'time_outdoors...' variables to remove '-10', '-3', '-1' values

-   Recode ethnicity to new collapsed variable + dummy vars for
    regression

-   Recode employment into new collapsed variable

-   recode blood sample time to collapse into one day

-   recode blood sample date into Date format

-   refactor vars for use in multivariable linear regression

-   2x new dfs: control_df and bipolar_df for later analysis use

```{r pressure, echo=FALSE, verbose=TRUE}
# remove people with NA glycated haemoglobin
ghaem_df <- analysis_df %>%  filter(!is.na(g_haem))

# remove people with NA glucose
ghaem_df <- ghaem_df %>%  filter(!is.na(glucose))

#remove people with glycated haemoglobin levels outside of manufacturer's reportable range
ghaem_df<- ghaem_df %>%  filter(!g_haem>184)
ghaem_df<- ghaem_df %>%  filter(!g_haem<15)

#remove people with glycated haemoglobin levels above 48mmol/ml (diabetes diagnostic criteria)
ghaem_df<- ghaem_df %>%  filter(!g_haem>48)

# add new column to df to indicate if ghaem value is IQR-defined outlier
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

ghaem_df <- ghaem_df %>%
        group_by(Group) %>%
        mutate(outlier = ifelse(find_outlier(g_haem), g_haem, NA))

# remove people with NA variables: bs_datetime (removes 2)
ghaem_df<- ghaem_df %>%  filter(!is.na(BS_datetime))

# make season numeric
ghaem_df$bs_season <- as.numeric(as.character(ghaem_df$bs_season))

# tidy time outdoors data to remove '-10', '-3', '-1' values:
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-10] <- 0
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-3] <- NA
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-1] <- NA

ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-10] <- 0
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-3] <- NA
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-1] <- NA

# tidy self-report sleep time to remove -3', '-1' values:
ghaem_df$sleep_duration[ghaem_df$sleep_duration ==-3] <- NA
ghaem_df$sleep_duration[ghaem_df$sleep_duration ==-1] <- NA


# recode ethnicity into 7 groups
library('plyr')
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer
ghaem_df$ethnicity_recode <- revalue(ghaem_df$ethnicity, c("British"="1", "Any other white background"="1", "White"="1", "Irish"="1", "Mixed"="2", "White and Black Caribbean"="2", "White and Black African"="2", "White and Asian"="2", "Any other mixed background"="2", "Asian or Asian British"="3", "Indian"="3", "Pakistani"="3", "Bangladeshi"="3", "Any other Asian background"="3", "Black or Black British"="4", "Caribbean"="4", "African"="4", "Any other Black background"="4", "Chinese"="5", "Other ethnic group"="6", "Do not know"="7", "Prefer not to answer"="7"))
#ghaem_df$ethnicity_recode <- as.numeric(ghaem_df$ethnicity_recode)

#ethnicity dummy vars code
#white
ghaem_df$ethnicity_white <- revalue(ghaem_df$ethnicity_recode, c("1"="1", "2"="0", "3"="0", "4"="0", "5"="0", "6"="0", "7"="0"))

#mixed
ghaem_df$ethnicity_mixed <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="1", "3"="0", "4"="0", "5"="0", "6"="0", "7"="0"))

#asian/asian british
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="1", "4"="0", "5"="0", "6"="0", "7"="0"))

#black/black british
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="1", "5"="0", "6"="0", "7"="0"))

#chinese
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="0", "5"="1", "6"="0", "7"="0"))

#other
ghaem_df$ethnicity_asian <- revalue(ghaem_df$ethnicity_recode, c("1"="0", "2"="0", "3"="0", "4"="0", "5"="0", "6"="1", "7"="0"))

# recode employment into 5 groups
#1=employed, 2=retired, 3=unable to work because of sickness/disability, 4=other unemployment, 5=do not know/prefer not to answer
ghaem_df$employment_recode <- revalue(ghaem_df$employed, c("In paid employment or self-employed"="1", "Retired"="2", "Doing unpaid or voluntary work"="4", "Unable to work because of sickness or disability"="3", "Full or part-time student"="4", "Unemployed"="4", "Looking after home and/or family"="4", "None of the above"="4", "Prefer not to answer"="5"))

#remove plyr package as has masking for later functions
detach("package:plyr", unload=TRUE)

# collapse bs_time into 1 day time variable - n=161 people with whacky time between 00:00 - 08:00
ghaem_df$bs_time_collapsed <- as.POSIXct(ghaem_df$bs_time,format="%H:%M")

    #create variables containing sequence of times to exclude from data
    # pull out the data date 
    start_time <- ymd_hms("2024-04-19 00:00:00")
    end_time <- ymd_hms("2024-04-19 08:00:00")
    exclude_dt_range <- seq(start_time, end_time, "min")
    exclude_time_range <- format(as.POSIXct(exclude_dt_range,format='%Y-%m-%d %H:%M:%S'),format='%Y-%m-%d %H:%M:%S')
exclude_time_range <- as.POSIXct(exclude_time_range)
    
    #remove rows where sample time is outside of 23:00:00-06:00:00 range
    bloodtime_df <- ghaem_df %>%
      filter(!bs_time_collapsed %in% exclude_time_range)

#create new date var
ghaem_df$bs_date <- as.Date(ghaem_df$BS_datetime)
#remove people with wrong datetime
ghaem_df<- ghaem_df %>%  filter(!bs_itime<1)
   
    
# make control and bipolar dataframes
#ungroup main df
ghaem_df = as.data.frame(ghaem_df)

#create groups
control_df <- ghaem_df %>%  filter(Group=='control')
bipolar_df <- ghaem_df %>%  filter(Group=='bipolar')
```

```{r recode for multivar lr}

library('plyr')

##Group
#control
ghaem_df$group <- revalue(ghaem_df$Group, c("control"="1", "bipolar"="2"))
ghaem_df$group_f <- factor(ghaem_df$group, levels= c("1", "2"), labels= c("control", "bipolar"), ordered=FALSE)

### LEVEL 1
#Sex
#sex
#ghaem_df$sex_recode <- revalue(ghaem_df$sex, c("Male"="1", "Female"="2"))
ghaem_df$sex_f <- factor(ghaem_df$sex, ordered=FALSE)

#ghaem_df$sex_fact <- factor(ghaem_df$sex_recode, levels= c("Male", "Female"), labels=c("male", "female"))

#ethnicity
# set values to NA if not relevant
ghaem_df$ethnicity_recode[ghaem_df$ethnicity_recode == -1] <- NA
ghaem_df$ethnicity_recode[ghaem_df$ethnicity_recode == -3] <- NA
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer
ghaem_df$ethnicity_f <- factor(ghaem_df$ethnicity_recode, levels= c("1", "2", "3", "4", "5", "6"), labels= c("white", "mixed", "asian", "black", "chinese", "other"), ordered=FALSE)

### LEVEL 2 
#smoking
# set values to NA if not relevant
ghaem_df$smoking_f <- ghaem_df$smoking_status
ghaem_df$smoking_f[ghaem_df$smoking_f == -3] <- NA
ghaem_df$smoking_f <- factor(ghaem_df$smoking_f, levels= c("Never", "Previous", "Current"), labels= c("Never", "Previous", "Current"), ordered=FALSE)

#alcohol - showing up strange!
#recode: never/special occasions=1 (LOW), 1-3month/1-2week=2 (MODERATE), 3-4week/daily=3 (HIGH)
ghaem_df$alcohol_f <- as.character(ghaem_df$alcohol_intake)
#ghaem_df$alcohol_f[ghaem_df$alcohol_f == 'Prefer not to answer'] <- NA
ghaem_df$alcohol_f <- revalue(ghaem_df$alcohol_f, c("Never"="Low", "Special occasions only"="Low", "One to three times a month"="Moderate", "Once or twice a week"="Moderate","Three or four times a week"="High", "Daily or almost daily"="High", "Prefer not to answer"=NA))
ghaem_df$alcohol_f <- factor(ghaem_df$alcohol_f, levels= c("Low", "Moderate", "High"), labels= c("Low", "Moderate", "High"), ordered=TRUE)

#health self-report
ghaem_df$health_f <- as.character(ghaem_df$health_self_report)
ghaem_df$health_f[ghaem_df$health_f == 'Prefer not to answer'] <- NA
ghaem_df$health_f[ghaem_df$health_f == 'Do not know'] <- NA
ghaem_df$health_f <- factor(ghaem_df$health_f, levels= c("Excellent", "Good", "Fair", "Poor"), labels= c("Excellent", "Good", "Fair", "Poor"), ordered=TRUE)


#disbility
ghaem_df$disability_f <- as.character(ghaem_df$disability_self_report)
ghaem_df$disability_f[ghaem_df$disability_f == 'Prefer not to answer'] <- NA
ghaem_df$disability_f[ghaem_df$disability_f == 'Do not know'] <- NA
ghaem_df$disability_f <- factor(ghaem_df$disability_f, levels= c("No", "Yes"), labels= c("No", "Yes"), ordered=FALSE)


#chronotype
ghaem_df$chrono_f <- as.character(ghaem_df$chronotype)
ghaem_df$chrono_f[ghaem_df$chrono_f == 'Prefer not to answer'] <- NA
ghaem_df$chrono_f[ghaem_df$chrono_f == 'Do not know'] <- NA
ghaem_df$chrono_f <- factor(ghaem_df$chrono_f, ordered=TRUE)

### Level 4 
#bs_time to integer
library(data.table)
ghaem_df$bs_itime <- as.ITime(ghaem_df$bs_time)
ghaem_df$bs_itime <- as.integer(ghaem_df$bs_itime)

#assess_centre to factor
ghaem_df$centre_f <- factor(ghaem_df$assess_centre, ordered=FALSE)

#remove plyr package as has masking for later functions
detach("package:plyr", unload=TRUE)
detach("package:data.table", unload=TRUE)

# re-make control and bipolar dataframes
#ungroup main df
ghaem_df = as.data.frame(ghaem_df)

#create groups
control_df <- ghaem_df %>%  filter(Group=='control')
bipolar_df <- ghaem_df %>%  filter(Group=='bipolar')
```

### 4: Table 1 outputs

-   Compute counts for discrete variables, mean+sd for continous
    variables

-   code for basic statistical tests

```{r table 1}

# summary statistics - median for continuous variables:
demo_mean_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter', 'fasting_time', 'townsend', 'sleep_duration'),
               mean, na.rm=TRUE)

# summary statistics - std.dev for continuous variables:
demo_sd_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter', 'fasting_time', 'townsend', 'sleep_duration'),
               sd, na.rm=TRUE)

#counts for discrete variables:
demo_sex_df <- ghaem_df %>%
  group_by(Group) %>%
  count(sex)
demo_ethnicity_df <- ghaem_df %>%
  group_by(Group) %>%
  count(ethnicity_recode)
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer

demo_disabled_df <- ghaem_df %>%
  group_by(Group) %>%
  count(disability_self_report)    
demo_chronotype_df <- ghaem_df %>%
  group_by(Group) %>%
  count(chronotype)    
demo_employment_df <- ghaem_df %>%
  group_by(Group) %>%
  count(employment_recode) 
demo_smoking_df <- ghaem_df %>%
  group_by(Group) %>%
  count(smoking_status)  
demo_alcohol_df <- ghaem_df %>%
  group_by(Group) %>%
  count(alcohol_intake)
demo_healthstatus_df <- ghaem_df %>%
  group_by(Group) %>%
  count(health_self_report) 
demo_IPAQ_df <- ghaem_df %>%
  group_by(Group) %>%
  count(IPAQ_group) 


#plot data to visualise distribution
# age - normal
age_plot <- ggplot(ghaem_df, aes(x = age)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# bmi - right skew
bmi_plot <- ggplot(ghaem_df, aes(x = BMI)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# age completed education - normal
age_edu_plot <- ggplot(ghaem_df, aes(x = age_completed_education)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: England - right skew
imd_e_plot <- ggplot(ghaem_df, aes(x = deprivation_index_england)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: Scotland - right skew
imd_s_plot <- ggplot(ghaem_df, aes(x = deprivation_index_scotland)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: Wales - right skew
imd_w_plot <- ggplot(ghaem_df, aes(x = deprivation_index_wales)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time outdoors summer - right skew
outsummer_plot <- ggplot(ghaem_df, aes(x = time_outdoors_summer)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time outdoors winter - right skew
outwinter_plot <- ggplot(ghaem_df, aes(x = time_outdoors_winter)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# fasting time - right skew
fastingtime_plot <- ggplot(ghaem_df, aes(x = fasting_time)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# townsend - right skew
townsend_plot <- ggplot(ghaem_df, aes(x = townsend)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")


# sleep duration - right skew
sleepdur_plot <- ggplot(ghaem_df, aes(x = sleep_duration)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# plot in multi-plot
figure <- ggarrange(age_plot,bmi_plot,age_edu_plot,imd_e_plot,imd_s_plot,imd_w_plot,outsummer_plot,outwinter_plot, fastingtime_plot,townsend_plot, labels = c("Age", "BMI", "Age Education", "IMD_E", "IMD_S", "IMD_W", "outdoors summer", "outdoors winter", "fasting time", "townsend"),
                    ncol = 5, nrow = 2)
figure


# stats tests - chi square for categorical
#sex
sex_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$sex))
yarrr::apa(sex_chi)# print stats test result in legible apa format

#ethnicity
ethnicity_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$ethnicity_recode))
yarrr::apa(ethnicity_chi)

#employment
employment_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$employment_recode))
yarrr::apa(employment_chi)

#disability
disability_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$disability_self_report=='Yes'))
yarrr::apa(disability_chi)

#smoking
smoking_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$smoking_status))
yarrr::apa(smoking_chi)

#alcohol
alcohol_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$alcohol_intake))
yarrr::apa(alcohol_chi)

#chronotype
chrono_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$chronotype))
yarrr::apa(chrono_chi)

#health self-report
health_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$health_self_report))
yarrr::apa(health_chi)

#IPAQ
IPAQ_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$IPAQ_group))
yarrr::apa(IPAQ_chi)

# stats tests - t-test for parametric continous
#age
age_ttest <- t.test(ghaem_df$age ~ ghaem_df$Group, na.rm=TRUE)
yarrr::apa(age_ttest)

# stats tests - wilcoxon rank sum for non-parametric continous
#BMI
bmi_wrs <- wilcox.test(BMI ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
bmi_wrs

#IMD ENGLAND
imd_e_wrs <- wilcox.test(deprivation_index_england ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_e_wrs

#IMD SCOTLAND
imd_s_wrs <- wilcox.test(deprivation_index_scotland ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_s_wrs

#IMD WALES
imd_w_wrs <- wilcox.test(deprivation_index_wales ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_w_wrs

#age completed education
edu_age_wrs <- wilcox.test(age_completed_education ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
edu_age_wrs

#time outdoors summer
outsummer_wrs <- wilcox.test(time_outdoors_summer ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
outsummer_wrs

#time outdoors winter
outwinter_wrs <- wilcox.test(time_outdoors_winter ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
outwinter_wrs

#fasting time
fasting_wrs <- wilcox.test(fasting_time ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
fasting_wrs

#townsend
townsend_wrs <- wilcox.test(townsend ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
townsend_wrs

#sleep duration
sleepdur_wrs <- wilcox.test(sleep_duration ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
sleepdur_wrs


### our marker levels
#hba1c
ghaem_ttest <- t.test(ghaem_df$g_haem ~ ghaem_df$Group, na.rm=TRUE)
yarrr::apa(ghaem_ttest)

#glucose
glucose_wrs <- wilcox.test(glucose ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
glucose_wrs
```

### 5. Explore blood samples - section in progress

```{r explore sample timings/distribution}

# timings
# season
ggplot(ghaem_df, aes(x = bs_season)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")


# month
ggplot(ghaem_df, aes(x = bs_month)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# day
ggplot(ghaem_df, aes(x = bs_wday)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time - edit time var from char to hm (lubridate)
ghaem_df$bs_time <- hm(ghaem_df$bs_time)

ggplot(ghaem_df, aes(x = bs_time)) +
  geom_histogram(fill = "salmon2", colour = "black", binwidth=10) +
  scale_x_continuous(breaks = seq(24*3600, 24*3600))+
  facet_grid(Group ~ ., scales = "free")

# levels
marker_mean_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('g_haem', 'glucose'), mean, na.rm=TRUE)

# summary statistics - std.dev for continuous variables:
marker_sd_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('g_haem', 'glucose'), sd, na.rm=TRUE)


# plot blood levels against time of day

#create groups with bloodtime_df to avoid excluding people from other df
bloodtime_control <- bloodtime_df %>%  filter(Group=='control')
bloodtime_bipolar <- bloodtime_df %>%  filter(Group=='bipolar')

bloodtime_control$bs_hour <- times(format(bloodtime_control$bs_time_collapsed, "%H"))

#plot glucose
ggplot(bloodtime_df, aes(bs_time_collapsed,glucose)) + geom_point(colour='coral1', size=0.1) + geom_smooth()+
scale_x_datetime(limits = ymd_hms(c('2024-04-19 07:00:00', '2024-04-19 20:00:00')))

#plot ghaem
dev.new()
ggplot(bloodtime_df, aes(bs_time_collapsed,g_haem)) + geom_point(colour='coral1', size=0.1) + geom_smooth()+
scale_x_datetime(limits = ymd_hms(c('2024-04-19 07:00:00', '2024-04-19 20:00:00')))

pirateplot(formula = glucose ~ bs_time_collapsed,
           data = bloodtime_control,
           theme = 4,
           main = "theme = 4")


#cosinor to look for circadian pattern - controls
#run cosinor on ghaem
hba1c_control_circcos <- cosinor(g_haem~1, date='bs_time_collapsed', type='hourly', data=bloodtime_control)
summary(hba1c_control_circcos)
plot(hba1c_control_circcos)
summary(hba1c_control_circcos$glm)
mesor_control <- mean(hba1c_control_circcos$glm$fitted.values) # for mesor
#
#seasrescheck(hba1c_control_circcos$residuals) # check the residuals

#run cosinor on glucose
glucose_control_circcos <- cosinor(glucose~1+ fasting_time, date='bs_time_collapsed', type='hourly', data=bloodtime_control)
summary(glucose_control_circcos)
plot(glucose_control_circcos)
summary(glucose_control_circcos$glm)
mesor_control <- mean(glucose_control_circcos$glm$fitted.values) # for mesor
#
#seasrescheck(hba1c_control_circcos$residuals) # check the residuals

#cosinor to look for circadian pattern - bipolar
#run cosinor on ghaem
hba1c_bipolar_circcos <- cosinor(g_haem~1, date='bs_time_collapsed', type='hourly', data=bloodtime_bipolar)
summary(hba1c_bipolar_circcos)
plot(hba1c_bipolar_circcos)
summary(hba1c_bipolar_circcos$glm)
mesor_control <- mean(hba1c_bipolar_circcos$glm$fitted.values) # for mesor
#
#seasrescheck(hba1c_bipolar_circcos$residuals) # check the residuals

#run cosinor on glucose
glucose_bipolar_circcos <- cosinor(glucose~1 + fasting_time, date='bs_time_collapsed', type='hourly', data=bloodtime_bipolar)
summary(glucose_bipolar_circcos)
plot(glucose_bipolar_circcos)
summary(glucose_bipolar_circcos$glm)
mesor_control <- mean(glucose_bipolar_circcos$glm$fitted.values) # for mesor
#
#seasrescheck(glucose_bipolar_circcos$residuals) # check the residuals


```

### 6. Collapse marker data into seasons and months

-   Glycated haemoglobin levels are reportedly higher in cooler months,
    and lower in warmer months. is this true for both our control
    groups?

    -   calculate mean + sd + CI for HbA1c levels for months
        (individually) and seasons

    -   correlation between temperature & Hba1c level - months

    -   correlation between (avg) daylength and (avg) rate of change &
        Hba1c levels - month

```{r gly_haem}

# task 1: visualise with data collapsed into months & seasons

#HBA1C
#months
ghaem_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), sd = sd(g_haem, na.rm=TRUE), n = length(g_haem)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
ghaem_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), sd = sd(g_haem, na.rm=TRUE), n = length(g_haem)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#GLUCOSE
#months
glucose_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(glucose, na.rm=TRUE), sd = sd(glucose, na.rm=TRUE), n = length(glucose)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
glucose_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(glucose, na.rm=TRUE), sd = sd(glucose, na.rm=TRUE), n = length(glucose)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#LACTATE
#months
lactate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(lactate, na.rm=TRUE), sd = sd(lactate, na.rm=TRUE), n = length(lactate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
lactate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(lactate, na.rm=TRUE), sd = sd(lactate, na.rm=TRUE), n = length(lactate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#PYRUVATE
#months
pyruvate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(pyruvate, na.rm=TRUE), sd = sd(pyruvate, na.rm=TRUE), n = length(pyruvate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
pyruvate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(pyruvate, na.rm=TRUE), sd = sd(pyruvate, na.rm=TRUE), n = length(pyruvate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#CITRATE
#months
citrate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(citrate, na.rm=TRUE), sd = sd(citrate, na.rm=TRUE), n = length(citrate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
citrate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(citrate, na.rm=TRUE), sd = sd(citrate, na.rm=TRUE), n = length(citrate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

# calculate mean & plot glycated haem - matched controls
#ghaem_month_matched <- match_exact_df %>%
#  group_by(bs_month, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))

#matched control seasons
#ghaem_season_matched <- match_exact_df %>%
#  group_by(bs_season, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))


#if wanting to calculate median + IQR
#ghaem_month <- ghaem_df %>%
#  group_by(bs_month, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))

```

### 7. create matched control group

1:1 matching based on sex, age, ethnicity, BMI, assess_centre, bs_month,
bs_year

CURRENT DEFAULT: do not run this code section. for optional future use
(sensitivity analyses)

```{r match, eval=FALSE, include=FALSE, verbose=FALSE}
# convert matching variables to type 'factor' for matchit function to work

for_match_df <- ghaem_df
for_match_df$group <- replace(for_match_df$group, for_match_df$group == 1, 0) 
for_match_df$group <- replace(for_match_df$group, for_match_df$group == 2, 1)
for_match_df$group <- as.integer(for_match_df$group)
# remove people with NA variables: ethnicity
for_match_df<- for_match_df %>%  filter(!is.na(ethnicity))
#for_match_df<- for_match_df %>%  filter(!is.na(BMI))

#use 1:1 nearest neighbour matching (Matchit default) without replacement on a propensity score estimated using a
# logistic regression of the treatment on the covariates
#m.out1 <- matchit(group ~ age + sex + ethnicity + assess_centre + bs_month + #bs_year, data = for_match_df)

#match_control_df <- match.data(m.out1)

#try different matching method: exact
m.out_exact <- matchit(group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year +bs_itime, data = for_match_df, method="nearest", 
                       exact= ~ bs_month + age + sex + assess_centre, 
                       ratio=3)
match_exact_df <- match.data(m.out_exact)
table(match_exact_df$Group)

table(match_exact_df$sex, match_exact_df$Group)

mc_df <- match_exact_df %>%  filter(Group=='control')


# match 1:1
m.out_exact <- matchit(group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year +bs_itime, data = for_match_df, method="nearest", 
                       exact= ~ bs_month + age + sex + assess_centre, 
                       ratio=1)
match_exact_df_1 <- match.data(m.out_exact)
table(match_exact_df_1$Group)

table(match_exact_df_1$sex, match_exact_df_1$Group)

mc_df_1 <- match_exact_df_1 %>%  filter(Group=='control')
```

```{}
```

### 8. Calculate cosinor and plot marker levels using pirate plot

```{r initialise pirate plot}

#colour gradient
col <- inferno(12, begin=0.3, end=0.7, direction=1) 
as.data.frame(col)

dayCOL<-c("#50576CFF", "#989278FF", "#ADA274FF", "#D8C563FF", "#EED654FF", "#FFEA46FF", "#EED654FF", "#D8C563FF", "#ADA274FF", "#858379FF", "#50576CFF", "#1B3B6DFF")


SeasonPal <- c("darkseagreen3","darkseagreen3","darkseagreen3","lightgoldenrod2","lightgoldenrod2","lightgoldenrod2","indianred3","indianred3","indianred3","lightsteelblue2","lightsteelblue2","lightsteelblue2") 

names <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
namesshort <-c("J","F","M","A","M","J","J","A","S","O","N","D")
```

#### [HbA1c - control cosinor]{.underline}

```{r HbA1c all control}

#run cosinor
hba1c_control_cos <- cosinor(g_haem~1 + control_df$age + control_df$BMI + control_df$sex_f + control_df$bs_itime + control_df$ethnicity_f + control_df$alcohol_f + control_df$health_f, date='bs_month', type='monthly',data=control_df)
summary(hba1c_control_cos)
summary(hba1c_control_cos$glm)
mesor_control <- mean(hba1c_control_cos$glm$fitted.values) # for mesor
#plot(hba1c_control_cos)
#seasrescheck(hba1c_control_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(hba1c_control_cos$glm$data, select = hba1c_control_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control

pirateplot(formula =  g_haem ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(34.4,35),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],hba1c_control_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], hba1c_control_cos$fitted.values[o_c], pch = 19)
```

#### HbA1c - matched cosinors

```{r matched cosinor}
#HbA1c - matched control cosinor
hba1c_mc_cos <- cosinor(g_haem~1 + mc_df$age + mc_df$BMI + mc_df$sex_f + mc_df$bs_itime + mc_df$ethnicity_f + mc_df$alcohol_f + mc_df$health_f, date='bs_month', type='monthly',data=mc_df)
summary(hba1c_mc_cos)
summary(hba1c_mc_cos$glm)
mesor_mc <- mean(hba1c_mc_cos$glm$fitted.values) # for mesor

#plot(hba1c_mc_cos)
#seasrescheck(hba1c_mc_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_mc_3 = subset(hba1c_mc_cos$glm$data, select = hba1c_mc_cos$date)[, 1]
o_mc3 = order(time_fit_mc_3)

# HbA1c - control

pirateplot(formula =  g_haem ~ bs_month,
           data = mc_df,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(34.4,35),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_mc_3[o_mc3],hba1c_mc_cos$fitted.values[o_mc3], lty=2,lwd=1, col="black")
points(time_fit_mc_3[o_mc3], hba1c_mc_cos$fitted.values[o_mc3], pch = 19)



```

#### HbA1c - bipolar cosinor

```{r}
# HbA1c - bipolar

#run cosinor
hba1c_bipolar_cos <- cosinor(g_haem~1 + bipolar_df$age + bipolar_df$BMI + bipolar_df$sex_f + bipolar_df$bs_itime + bipolar_df$ethnicity_f + bipolar_df$alcohol_f + bipolar_df$health_f, date='bs_month', type='monthly', data=bipolar_df, cycles=1)
summary(hba1c_bipolar_cos)
summary(hba1c_bipolar_cos$glm)
mesor_bipolar <- mean(hba1c_bipolar_cos$glm$fitted.values) # for mesor
#plot(hba1c_bipolar_cos)
#seasrescheck(hba1c_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(hba1c_bipolar_cos$glm$data, select = hba1c_bipolar_cos$date)[, 1]
o_bd = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  g_haem ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(32,36.5) # Adjust y axis limits
           #ylim=c(34.4,35)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o_bd],hba1c_bipolar_cos$fitted.values[o_bd], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o_bd], hba1c_bipolar_cos$fitted.values[o_bd], pch = 19)



```

#### HbA1c - plot all groups

```{r}
#run significance test between groups for: phase, nadir, amplitude, mesor
#using circacompare - alpha threshold set to 0.5 because bipolar group is non-rhythmic
cos_stats_ghaem <- circacompare(x=ghaem_df, col_time="bs_month", col_group="Group", col_outcome="g_haem", period=12, alpha_threshold=0.5)


#create plot with fitted sinusoid
pirateplot(formula = g_haem ~ time_fit_control,
           data = control_df,
           theme = 0, # set theme to 0 for blank plot
           xlab = "Month", # add x axis label
           ylab = expression(bold(paste("HbA1c (mmol/mol)"))), # add y axis label
           xaxt = "n", # remove x axis check marks
           yaxt = "n", # remove y axis check markers
           ylim=c(23,27), # define y axis limits
           main = "HbA1c fitted sinusoids")

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
#add lines for control
lines(time_fit_control[o_c],hba1c_control_cos$fitted.values[o_c], lty=2,lwd=1, col="red")
points(time_fit_control[o_c], hba1c_control_cos$fitted.values[o_c], pch = 19, col="red")
#add lines for matched control (3:1)
lines(time_fit_mc_3[o_mc3],hba1c_mc_cos$fitted.values[o_mc3], lty=2,lwd=1, col="blue")
points(time_fit_mc_3[o_mc3], hba1c_mc_cos$fitted.values[o_mc3], pch = 19, col="blue")
#add lines for bipolar
lines(time_fit_bipolar[o_bd],hba1c_bipolar_cos$fitted.values[o_bd], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o_bd], hba1c_bipolar_cos$fitted.values[o_bd], pch = 19, col="black")
```

#### [Glucose]{.underline}

```{r Glucose cosinor+plot}

#run cosinor
glucose_control_cos <- cosinor(glucose~1, date='bs_month', type='monthly', data=control_df)
summary(glucose_control_cos)
summary(glucose_control_cos$glm)
mesor_control <- mean(glucose_control_cos$glm$fitted.values)
#plot(glucose_control_cos)
#seasrescheck(glucose_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(glucose_control_cos$glm$data, select = glucose_control_cos$date)[, 1]
o = order(time_fit_control)

# Glucose - control
dev.new()
pirateplot(formula =  glucose ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(4.85,5),  # Adjust y axis limits
           ylim=c(4.5,5.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],glucose_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], glucose_control_cos$fitted.values[o], pch = 19)
       

# Glucose - bipolar

#run cosinor
glucose_bipolar_cos <- cosinor(glucose~1, date='bs_month', type='monthly', data=bipolar_df)
summary(glucose_bipolar_cos)
summary(glucose_bipolar_cos$glm)
mesor_bipolar <- mean(glucose_bipolar_cos$glm$fitted.values)
#plot(glucose_bipolar_cos)
#seasrescheck(glucose_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(glucose_bipolar_cos$glm$data, select = glucose_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  glucose ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(4.5,5.5) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],glucose_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], glucose_bipolar_cos$fitted.values[o], pch = 19)

#run significance test between groups for: phase, nadir, amplitude, mesor
#using circacompare - alpha threshold set to 0.7 because bipolar group is non-rhythmic
cos_stats_glucose <- circacompare(x=ghaem_df, col_time="bs_month", col_group="Group", col_outcome="glucose", period=12, alpha_threshold=0.7)
```

```{r}
## glucose - matched control cosinor 3:1
glucose_mc_cos <- cosinor(glucose~1, date='bs_month', type='monthly', data=mc_df)
summary(glucose_mc_cos)
summary(glucose_mc_cos$glm)
mesor_control <- mean(glucose_mc_cos$glm$fitted.values)
#plot(glucose_mc_cos)
#seasrescheck(glucose_mc_cos$residuals) # check the residuals

#1:1
glucose_mc1_cos <- cosinor(glucose~1, date='bs_month', type='monthly', data=mc_df_1)
summary(glucose_mc1_cos)
summary(glucose_mc1_cos$glm)
mesor_control1 <- mean(glucose_mc1_cos$glm$fitted.values)

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_mc = subset(glucose_mc_cos$glm$data, select = glucose_mc_cos$date)[, 1]
o = order(time_fit_mc)

# Glucose - matched control
dev.new()
pirateplot(formula =  glucose ~ bs_month,
           data = mc_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=SeasonPal, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = SeasonPal,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           #ylim=c(4.85,5),  # Adjust y axis limits
           ylim=c(4.5,5.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_mc[o],glucose_mc_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_mc[o], glucose_mc_cos$fitted.values[o], pch = 19)
       
```

### 9. Multivariable analysis between marker & daylength ( proxy season marker)

Increasing model adjustment to include:

-   Level 1: sociodemographic factors (sex, age, ethnicity, townsend)

-   Level 2: health & lifestyle/behavioural factors (BMI, smoking,
    alcohol, health self report, disability, chronotype)

-   Level 3: environmental & seasonal factors (sample time of day,
    photoperiod rate of change, latitude, temperature, time outdoors in
    summer, time outdoors in winter)

-   Level 4: analyser/location

```{=html}
<!-- -->
```

#### HbA1c

```{r HbA1c multivariable LM analysis}
# https://stats.oarc.ucla.edu/r/modules/coding-for-categorical-variables-in-regression-models/
library (olsrr)

#plot - is there a visible linear relationship?
ggplot(ghaem_df, aes(x=daylength_min, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(20, 50)


#test for interaction effects
l1_c_anova <- aov(g_haem ~ daylength_min * 
               #group_f +
               #level 1
               age * sex_f * townsend * ethnicity_f,
  data = control_df)
summary(l1_c_anova)  # sig sex:ethncity

l2_c_anova <- aov(g_haem ~ daylength_min * 
               #level 2
               BMI * smoking_f*
               alcohol_f * health_f * 
               chrono_f,
  data = control_df)
summary(l2_c_anova)

l3_c_anova <- aov(g_haem ~ daylength_min * 
               #level 3
               photoperiod_roc_min * assess_latitude *
               time_outdoors_summer * time_outdoors_winter,
  data = control_df)
summary(l3_c_anova)

l4_c_anova <- aov(g_haem ~ daylength_min * 
               #level 4
               assess_centre * bs_itime,
  data = control_df)
summary(l4_c_anova)
  
####CONTROL
#multivariate analysis - level 1
model1 <- lm(g_haem ~ daylength_min +
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f,
  data = control_df
)
summary(model1)
#abline(model1, col="red", lwd=2) # fit line to plot

# test for multicollinearity
ols_vif_tol(model1)

#multivariate analysis - level 2
model2 <- lm(g_haem ~ daylength_min + 
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f,
  data = control_df
)
summary(model2)

# test for multicollinearity
ols_vif_tol(model2)

#multivariate analysis - level 3
# average temp removed due to high collinearity (VIF: 11.944)
model3 <- lm(g_haem ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter, 
  data = control_df
)
summary(model3)

# test for multicollinearity
ols_vif_tol(model3)

#multivariate analysis - level 4
model4 <- lm(g_haem ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter +
               #level 4
               centre_f + bs_itime,
  data = control_df
)
summary(model4)
# test for multicollinearity
ols_vif_tol(model4)

####BIPOLAR
#multivariate analysis - level 1
model1_b <- lm(g_haem ~ daylength_min +
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f,
  data = bipolar_df
)
summary(model1_b)
# test for multicollinearity
ols_vif_tol(model1_b)

#multivariate analysis - level 2
model2_b <- lm(g_haem ~ daylength_min + 
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f,
  data = bipolar_df
)
summary(model2_b)
# test for multicollinearity
ols_vif_tol(model2_b)

#multivariate analysis - level 3
# average temp removed due to high collinearity (VIF: 13.166)
model3_b <- lm(g_haem ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter, 
  data = bipolar_df
)
summary(model3_b)
# test for multicollinearity
ols_vif_tol(model3_b)

#multivariate analysis - level 4
model4_b <- lm(g_haem ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter +
               #level 4
               centre_f + bs_itime,
  data = bipolar_df
)
summary(model4_b)
# test for multicollinearity
ols_vif_tol(model4_b)
```

#### Glucose

```{r glucose multivariable LM analysis}

#plot - is there a visible linear relationship?
dev.new()
ggplot(ghaem_df, aes(x=daylength_min, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(20, 50)

####CONTROL
#multivariate analysis - level 1
model1_glu <- lm(glucose ~ daylength_min +
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f,
  data = control_df
)
summary(model1_glu)

# test for multicollinearity
ols_vif_tol(model1_glu)

#multivariate analysis - level 2
model2_glu <- lm(glucose ~ daylength_min + 
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f,
  data = control_df
)
summary(model2_glu)

# test for multicollinearity
ols_vif_tol(model2_glu)

#multivariate analysis - level 3
# average temp removed due to high collinearity (VIF: 11.944)
model3_glu <- lm(glucose ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter, 
  data = control_df
)
summary(model3_glu)
# test for multicollinearity
ols_vif_tol(model3_glu)


#multivariate analysis - level 4
model4_glu <- lm(glucose ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter +
               #level 4
               centre_f + bs_itime,
  data = control_df
)
summary(model4_glu)
# test for multicollinearity
ols_vif_tol(model4_glu)


####BIPOLAR
#multivariate analysis - level 1
model1_glu_b <- lm(glucose ~ daylength_min +
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f,
  data = bipolar_df
)
summary(model1_glu_b)
# test for multicollinearity
ols_vif_tol(model1_glu_b)

#multivariate analysis - level 2
model2_glu_b <- lm(glucose ~ daylength_min + 
               #group +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f,
  data = bipolar_df
)
summary(model2_glu_b)
# test for multicollinearity
ols_vif_tol(model2_glu_b)

#multivariate analysis - level 3
# average temp removed due to high collinearity (VIF: 11.944)
model3_glu_b <- lm(glucose ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               #bs_time + (NEED TO BIN BS_TIME)
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter, 
  data = bipolar_df
)
summary(model3_glu_b)
# test for multicollinearity
ols_vif_tol(model3_glu_b)

#multivariate analysis - level 4
model4_glu_b <- lm(glucose ~ daylength_min + 
               #group_f +
               #level 1
               age + sex_f + townsend + ethnicity_f +
               #level 2
               BMI + smoking_f +
               alcohol_f + health_f + 
               chrono_f +
               #level 3
               photoperiod_roc_min + assess_latitude +
               time_outdoors_summer + time_outdoors_winter +
               #level 4
               centre_f + bs_itime,
  data = bipolar_df
)
summary(model4_glu_b)
# test for multicollinearity
ols_vif_tol(model4_glu_b)
```

## Extra code for different tasks

#### extra: code for multivariable cosinor

```{r cosinor additional variables}

# test run cosinor with additional vars
test_res=cosinor(glucose~1 + fasting_time, date='bs_month', type='monthly', data=control_df)
summary(test_res)
summary(test_res$glm)
plot(test_res)
seasrescheck(test_res$residuals) # check the residuals

```

#### extra: code to explore correlation between diabetic marker levels and different variables

```{r}
# seasonal variables

#### temperature
range(ghaem_df$lactate, na.rm=TRUE)  # useful code to determine ranges for ylims

#1. multi-plot: temperature and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=average_temp, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(15, 48) #ylim for plotting points
  ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=average_temp, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(1.5, 14.5) #ylim for plotting points
  ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=average_temp, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0.5, 13) #ylim for plotting points
  ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=average_temp, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0, 2) #ylim for plotting points
  ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=average_temp, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0, 0.3) #ylim for plotting points
  ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for temp and diabetic markers
temp_ghaem <- cor.test(ghaem_df$average_temp, ghaem_df$g_haem, method="pearson")
yarrr::apa(temp_ghaem)
temp_glucose <- cor.test(ghaem_df$average_temp, ghaem_df$glucose, method="pearson")
yarrr::apa(temp_glucose)
temp_lactate <- cor.test(ghaem_df$average_temp, ghaem_df$lactate, method="pearson")
yarrr::apa(temp_lactate)
temp_pyruvate <- cor.test(ghaem_df$average_temp, ghaem_df$pyruvate, method="pearson")
yarrr::apa(temp_pyruvate)
temp_citrate <- cor.test(ghaem_df$average_temp, ghaem_df$citrate, method="pearson")
yarrr::apa(temp_citrate)

####  daylength
#1. multi-plot: daylength and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=daylength_min, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=daylength_min, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=daylength_min, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=daylength_min, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=daylength_min, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for daylength and diabetic markers
day_ghaem <- cor.test(ghaem_df$daylength_min, ghaem_df$g_haem, method="pearson")
yarrr::apa(day_ghaem)
day_glucose <- cor.test(ghaem_df$daylength_min, ghaem_df$glucose, method="pearson")
yarrr::apa(day_glucose)
day_lactate <- cor.test(ghaem_df$daylength_min, ghaem_df$lactate, method="pearson")
yarrr::apa(day_lactate)
day_pyruvate <- cor.test(ghaem_df$daylength_min, ghaem_df$pyruvate, method="pearson")
yarrr::apa(day_pyruvate)
day_citrate <- cor.test(ghaem_df$daylength_min, ghaem_df$citrate, method="pearson")
yarrr::apa(day_citrate)


####  ROC
#1. multi-plot: photoperiod roc and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for photoperiod roc and diabetic markers
roc_ghaem <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$g_haem, method="pearson")
yarrr::apa(roc_ghaem)
roc_glucose <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$glucose, method="pearson")
yarrr::apa(roc_glucose)
roc_lactate <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$lactate, method="pearson")
yarrr::apa(roc_lactate)
roc_pyruvate <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$pyruvate, method="pearson")
yarrr::apa(roc_pyruvate)
roc_citrate <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$citrate, method="pearson")
yarrr::apa(roc_citrate)

#####geographic
####  latitude
#1. multi-plot: latitude and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=assess_latitude, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=assess_latitude, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=assess_latitude, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=assess_latitude, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=assess_latitude, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for latitude and diabetic markers
latitude_ghaem <- cor.test(ghaem_df$assess_latitude, ghaem_df$g_haem, method="pearson")
yarrr::apa(latitude_ghaem)
latitude_glucose <- cor.test(ghaem_df$assess_latitude, ghaem_df$glucose, method="pearson")
yarrr::apa(latitude_glucose)
latitude_lactate <- cor.test(ghaem_df$assess_latitude, ghaem_df$lactate, method="pearson")
yarrr::apa(latitude_lactate)
latitude_pyruvate <- cor.test(ghaem_df$assess_latitude, ghaem_df$pyruvate, method="pearson")
yarrr::apa(latitude_pyruvate)
latitude_citrate <- cor.test(ghaem_df$assess_latitude, ghaem_df$citrate, method="pearson")
yarrr::apa(latitude_citrate)

#####time - does not work yet, time format issues!
####  time of day
test_time <- ghaem_df
test_time$bs_time <- as.POSIXct(test_time$bs_time)
ghaem_df$bs_time <- as.POSIXct(ghaem_df$bs_time)
#1. multi-plot: sample TOD and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=bs_time, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=bs_time, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=bs_time, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=bs_time, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=bs_time, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for time and diabetic markers
time_ghaem <- cor.test(ghaem_df$bs_time, ghaem_df$g_haem, method="pearson")
yarrr::apa(time_ghaem)
time_glucose <- cor.test(ghaem_df$bs_time, ghaem_df$glucose, method="pearson")
yarrr::apa(time_glucose)
time_lactate <- cor.test(ghaem_df$bs_time, ghaem_df$lactate, method="pearson")
yarrr::apa(time_lactate)
time_pyruvate <- cor.test(ghaem_df$bs_time, ghaem_df$pyruvate, method="pearson")
yarrr::apa(time_pyruvate)
time_citrate <- cor.test(ghaem_df$bs_time, ghaem_df$citrate, method="pearson")
yarrr::apa(time_citrate)


```

#### extra: plotting

```{r plotting}
 
#plot mean values plus 95%CI
ggplot(citrate_season, aes(bs_season, mean, color=Group))+
  geom_line()+ggtitle("citrate seasonal Profile")+
  geom_point()+
  geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), width=.2,
                position=position_dodge(0.05))+
  scale_x_continuous("bs_season", labels = as.character(citrate_season$bs_season), breaks = citrate_season$bs_season)

# plot temp correlation
ggplot(ghaem_df, aes(x=photoperiod_roc, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  ylim(20, 50)
model <- ghaem_df %>%
  filter(Group=="control") %>%
  lm(g_haem~average_temp)
temp_ghaem <- cor.test(ghaem_df$average_temp, ghaem_df$g_haem, method=c("pearson", "kendall", "spearman"))
```

#### extra: correlations between different variables

```{r}
#check assumptions
#lm code
model <- ghaem_df %>%
  filter(Group=="control") %>%
  lm(g_haem~average_temp, data=ghaem_df)
check_model(ghaem_df$age)

```

#### extra: dummy coding

```{r dummy coding}
###create dummy variables

##Smoking
#previous
ghaem_df$smoking_previous <- revalue(ghaem_df$smoking_status, c("Previous"="1", "Never"="0", "Current"="0", "Prefer not to answer"="0"))

#never
ghaem_df$smoking_never <- revalue(ghaem_df$smoking_status, c("Previous"="0", "Never"="1", "Current"="0", "Prefer not to answer"="0"))

#current
ghaem_df$smoking_current <- revalue(ghaem_df$smoking_status, c("Previous"="0", "Never"="0", "Current"="1", "Prefer not to answer"="0"))

##Alcohol
#never/special
ghaem_df$alcohol_rare <- revalue(ghaem_df$alcohol_intake, c("Special occasions only"="1", "Never"="1", "One to three times a month"="0", "Once or twice a week"="0","Three or four times a week"="0", "Daily or almost daily"="0","Prefer not to answer"="0"))

#moderate
ghaem_df$alcohol_moderate <- revalue(ghaem_df$alcohol_intake, c("Special occasions only"="0", "Never"="0", "One to three times a month"="1", "Once or twice a week"="1","Three or four times a week"="0", "Daily or almost daily"="0","Prefer not to answer"="0"))


#frequent
ghaem_df$alcohol_frequent <- revalue(ghaem_df$alcohol_intake, c("Special occasions only"="0", "Never"="0", "One to three times a month"="0", "Once or twice a week"="0","Three or four times a week"="1", "Daily or almost daily"="1","Prefer not to answer"="0"))

#near daily
ghaem_df$alcohol_daily <- revalue(ghaem_df$alcohol_intake, c("Special occasions only"="0", "Never"="0", "One to three times a month"="0", "Once or twice a week"="0","Three or four times a week"="0", "Daily or almost daily"="1","Prefer not to answer"="0"))

##Disability self-report
#yes
ghaem_df$disability_yes <- revalue(ghaem_df$disability_self_report, c("Yes"="1", "No"="0", "Do not know"="0", "Prefer not to answer"="0"))
  
#no
ghaem_df$disability_no <- revalue(ghaem_df$disability_self_report, c("Yes"="0", "No"="1", "Do not know"="0", "Prefer not to answer"="0"))

##health status self-report
#Excellent
ghaem_df$health_excellent <- revalue(ghaem_df$health_self_report, c("Excellent"="1", "Good"="0", "Fair"="0", "Poor"="0","Do not know"="0", "Prefer not to answer"="0"))
  
#Good
ghaem_df$health_good <- revalue(ghaem_df$health_self_report, c("Excellent"="0", "Good"="1", "Fair"="0", "Poor"="0","Do not know"="0", "Prefer not to answer"="0"))

#Fair
ghaem_df$health_fair <- revalue(ghaem_df$health_self_report, c("Excellent"="0", "Good"="0", "Fair"="1", "Poor"="0","Do not know"="0", "Prefer not to answer"="0"))

#Poor
ghaem_df$health_poor <- revalue(ghaem_df$health_self_report, c("Excellent"="0", "Good"="0", "Fair"="0", "Poor"="1","Do not know"="0", "Prefer not to answer"="0"))

##Chronotype
#def morning
ghaem_df$chrono_defmorn <- revalue(ghaem_df$chronotype, c("Definitely a 'morning' person"="1", "More a 'morning' than 'evening' person"="0", "More an 'evening' than a 'morning' person"="0", "Definitely an 'evening' person"="0","Do not know"="0", "Prefer not to answer"="0"))

#more morning
ghaem_df$chrono_moremorn <- revalue(ghaem_df$chronotype, c("Definitely a 'morning' person"="0", "More a 'morning' than 'evening' person"="1", "More an 'evening' than a 'morning' person"="0", "Definitely an 'evening' person"="0","Do not know"="0", "Prefer not to answer"="0"))

#more evening
ghaem_df$chrono_moreeve <- revalue(ghaem_df$chronotype, c("Definitely a 'morning' person"="0", "More a 'morning' than 'evening' person"="0", "More an 'evening' than a 'morning' person"="1", "Definitely an 'evening' person"="0","Do not know"="0", "Prefer not to answer"="0"))

#def evening
ghaem_df$chrono_defeve <- revalue(ghaem_df$chronotype, c("Definitely a 'morning' person"="0", "More a 'morning' than 'evening' person"="0", "More an 'evening' than a 'morning' person"="0", "Definitely an 'evening' person"="1","Do not know"="0", "Prefer not to answer"="0"))

```

#### extra: cosinor for lactate/pyruvate/citrate

##### [Lactate]{.underline}

```{r Lactate cosinor+plot}

#run cosinor
lactate_control_cos <- cosinor(lactate~1, date='bs_month', type='monthly', data=control_df)
summary(lactate_control_cos)
summary(lactate_control_cos$glm)
#plot(lactate_control_cos)
#seasrescheck(lactate_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(lactate_control_cos$glm$data, select = lactate_control_cos$date)[, 1]
o = order(time_fit_control)

# Lactate - control
dev.new() # open new plot
pirateplot(formula =  lactate ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Lactate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(3.5,4.5),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],lactate_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], lactate_control_cos$fitted.values[o], pch = 19)
       

# Lactate - bipolar

#run cosinor
lactate_bipolar_cos <- cosinor(lactate~1, date='bs_month', type='monthly', data=bipolar_df)
summary(lactate_bipolar_cos)
summary(lactate_bipolar_cos$glm)
#plot(lactate_bipolar_cos)
#seasrescheck(lactate_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(lactate_bipolar_cos$glm$data, select = lactate_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  lactate ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Lactate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(3,5) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],lactate_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], lactate_bipolar_cos$fitted.values[o], pch = 19)

```

##### [Pyruvate]{.underline}

```{r Pyruvate cosinor+plot}

#run cosinor
pyruvate_control_cos <- cosinor(pyruvate~1, date='bs_month', type='monthly', data=control_df)
summary(pyruvate_control_cos)
summary(pyruvate_control_cos$glm)
#plot(pyruvate_control_cos)
#seasrescheck(pyruvate_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(pyruvate_control_cos$glm$data, select = pyruvate_control_cos$date)[, 1]
o = order(time_fit_control)

# Pyruvate - control
dev.new() # open new plot
pirateplot(formula =  pyruvate ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Pyruvate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.07,0.09),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],pyruvate_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], pyruvate_control_cos$fitted.values[o], pch = 19)
       

# Pyruvate - bipolar

#run cosinor
pyruvate_bipolar_cos <- cosinor(pyruvate~1, date='bs_month', type='monthly', data=bipolar_df)
summary(pyruvate_bipolar_cos)
summary(pyruvate_bipolar_cos$glm)
#plot(pyruvate_bipolar_cos)
#seasrescheck(pyruvate_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(pyruvate_bipolar_cos$glm$data, select = pyruvate_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  pyruvate ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Pyruvate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.05,0.12) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],pyruvate_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], pyruvate_bipolar_cos$fitted.values[o], pch = 19)

```

##### [Citrate]{.underline}

```{r Citrate cosinor+plot}

#run cosinor
citrate_control_cos <- cosinor(citrate~1, date='bs_month', type='monthly', data=control_df)
summary(citrate_control_cos)
summary(citrate_control_cos$glm)
#plot(citrate_control_cos)
#seasrescheck(citrate_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(citrate_control_cos$glm$data, select = citrate_control_cos$date)[, 1]
o = order(time_fit_control)

# Citrate - control
dev.new() # open new plot
pirateplot(formula =  citrate ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Citrate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.062,0.068),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],citrate_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], citrate_control_cos$fitted.values[o], pch = 19)
       

# Citrate - bipolar

#run cosinor
citrate_bipolar_cos <- cosinor(citrate~1, date='bs_month', type='monthly', data=bipolar_df)
summary(citrate_bipolar_cos)
summary(citrate_bipolar_cos$glm)
#plot(citrate_bipolar_cos)
#seasrescheck(citrate_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(citrate_bipolar_cos$glm$data, select = citrate_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  citrate ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Citrate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.05,0.08) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],citrate_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], citrate_bipolar_cos$fitted.values[o], pch = 19)

```

### 
