---
title: "UKB_ghaem"
author: "agpr141"
date: "2024-03-18"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# BD Seasonality & Diabetic Biomarkers

### 1: Import libs

```{r load, echo=FALSE, verbose=FALSE}
#load libraries
library(tidyverse)
library(dplyr)
library(readr)
library(MatchIt)
library(ggplot2)
library(lubridate)
library(broom)
library(hrbrthemes)
library(viridis)
library(yarrr)
library(ggpubr)
library(sjstats)
library(plotrix)
library(segmented)
library(plot3D)
library(season)
```

### 2: Load data

```{r load data}

# load participant dataframe for analysis
load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Final study participant dataframe\\analysis_df.Rda")

fasting_time <- readr::read_tsv("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\fasting_time.tsv")

fasting_time <- fasting_time %>% 
  rename('fasting_time' = 'p74_i0')

analysis_df <- merge(analysis_df,fasting_time, by="eid")
```

### 3: Tidy data

Remove:

-   People with NA for glycated haemoglobin

-   People with glycated haemoglobin levels outside of analytic
    reference range 15-184mmol/mol

-   People with glycated haemoglobin levels \>48mmol/mol (diagnostic
    level for diabetes)

Make:

-   New outlier variable for glycated haemoglobin

-   Season variable from month data (1=Spring, 2=Summer, 3=Autumn,
    4=Winter)

-   Tidy 'time_outdoors...' variables to remove '-10', '-3', '-1' values

-   2x new dfs: control_df and bipolar_df for later analysis use

```{r pressure, echo=FALSE}
# remove people with NA glycated haemoglobin
ghaem_df <- analysis_df %>%  filter(!is.na(g_haem))

#remove people with glycated haemoglobin levels outside of manufacturer's reportable range
ghaem_df<- ghaem_df %>%  filter(!g_haem>184)
ghaem_df<- ghaem_df %>%  filter(!g_haem<15)

#remove people with glycated haemoglobin levels above 48mmol/ml (diabetes diagnostic criteria)
ghaem_df<- ghaem_df %>%  filter(!g_haem>48)

# add new column to df to indicate if ghaem value is IQR-defined outlier
find_outlier <- function(x) {
  return(x < quantile(x, .25) - 1.5*IQR(x) | x > quantile(x, .75) + 1.5*IQR(x))
}

ghaem_df <- ghaem_df %>%
        group_by(Group) %>%
        mutate(outlier = ifelse(find_outlier(g_haem), g_haem, NA))

# remove people with NA variables: bs_datetime (removes 2)
ghaem_df<- ghaem_df %>%  filter(!is.na(BS_datetime))

# make season numeric
ghaem_df$bs_season <- as.numeric(as.character(ghaem_df$bs_season))

# tidy time outdoors data to remove '-10', '-3', '-1' values:
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-10] <- 0
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-3] <- NA
ghaem_df$time_outdoors_summer[ghaem_df$time_outdoors_summer ==-1] <- NA

ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-10] <- 0
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-3] <- NA
ghaem_df$time_outdoors_winter[ghaem_df$time_outdoors_winter ==-1] <- NA


# make control and bipolar dataframes

#ungroup main df
ghaem_df = as.data.frame(ghaem_df)

#create groups
control_df <- ghaem_df %>%  filter(Group=='control')
bipolar_df <- ghaem_df %>%  filter(Group=='bipolar')
```

### 4: Table 1 outputs

-   Compute counts for discrete variables, mean+sd for continous
    variables

-   code for basic statistical tests

```{r table 1}

# summary statistics - median for continuous variables:
demo_mean_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter'),
               mean, na.rm=TRUE)

# summary statistics - std.dev for continuous variables:
demo_sd_df <- ghaem_df %>%
  group_by(Group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter'),
               sd, na.rm=TRUE)

#counts for discrete variables:
demo_sex_df <- ghaem_df %>%
  group_by(Group) %>%
    #mutate(sex = factor(sex)) %>% 
  count(sex)
demo_ethnicity_df <- ghaem_df %>%
  group_by(Group) %>%
  count(ethnicity)
demo_disabled_df <- ghaem_df %>%
  group_by(Group) %>%
  count(disability_self_report)    
demo_chronotype_df <- ghaem_df %>%
  group_by(Group) %>%
  count(chronotype)    
demo_employment_df <- ghaem_df %>%
  group_by(Group) %>%
  count(employed) 
demo_smoking_df <- ghaem_df %>%
  group_by(Group) %>%
  count(smoking_status)  
demo_alcohol_df <- ghaem_df %>%
  group_by(Group) %>%
  count(alcohol_intake)  


#plot data to visualise distribution
# age - normal
age_plot <- ggplot(ghaem_df, aes(x = age)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# bmi - right skew
bmi_plot <- ggplot(ghaem_df, aes(x = BMI)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# age completed education - normal
age_edu_plot <- ggplot(ghaem_df, aes(x = age_completed_education)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: England - right skew
imd_e_plot <- ggplot(ghaem_df, aes(x = deprivation_index_england)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: Scotland - right skew
imd_s_plot <- ggplot(ghaem_df, aes(x = deprivation_index_scotland)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: Wales - right skew
imd_w_plot <- ggplot(ghaem_df, aes(x = deprivation_index_wales)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time outdoors summer - right skew
outsummer_plot <- ggplot(ghaem_df, aes(x = time_outdoors_summer)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time outdoors winter - right skew
outwinter_plot <- ggplot(ghaem_df, aes(x = time_outdoors_winter)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# plot in multi-plot
figure <- ggarrange(age_plot,bmi_plot,age_edu_plot,imd_e_plot,imd_s_plot,imd_w_plot,outsummer_plot,outwinter_plot,labels = c("Age", "BMI", "Age Education", "IMD_E", "IMD_S", "IMD_W", "outdoors summer", "outdoors winter"),
                    ncol = 4, nrow = 2)
figure


# stats tests - chi square for categorical
#sex
sex_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$sex))
yarrr::apa(sex_chi)# print stats test result in legible apa format

#ethnicity
ethnicity_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$ethnicity))
yarrr::apa(ethnicity_chi)

#employment
employment_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$employed))
yarrr::apa(employment_chi)

#disability
disability_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$disability_self_report=='Yes'))
yarrr::apa(disability_chi)

#smoking
smoking_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$smoking_status))
yarrr::apa(smoking_chi)

#alcohol
alcohol_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$alcohol_intake))
yarrr::apa(alcohol_chi)

#chronotype
chrono_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$chronotype))
yarrr::apa(chrono_chi)

# stats tests - t-test for parametric continous
#age
age_ttest <- t.test(ghaem_df$age ~ ghaem_df$Group, na.rm=TRUE)
yarrr::apa(age_ttest)

# stats tests - wilcoxon rank sum for non-parametric continous
#BMI
bmi_wrs <- wilcox.test(BMI ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
bmi_wrs

#IMD ENGLAND
imd_e_wrs <- wilcox.test(deprivation_index_england ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_e_wrs

#IMD SCOTLAND
imd_s_wrs <- wilcox.test(deprivation_index_scotland ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_s_wrs

#IMD WALES
imd_w_wrs <- wilcox.test(deprivation_index_wales ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_w_wrs

#age completed education
edu_age_wrs <- wilcox.test(age_completed_education ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
edu_age_wrs

#time outdoors summer
outsummer_wrs <- wilcox.test(time_outdoors_summer ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
outsummer_wrs

#time outdoors winter
outwinter_wrs <- wilcox.test(time_outdoors_winter ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
outwinter_wrs
```

### 5. Explore blood sample timings - section in progress

```{r explore sample timings/distribution}

# season
ggplot(ghaem_df, aes(x = bs_season)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")


# month
ggplot(ghaem_df, aes(x = bs_month)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# day
ggplot(ghaem_df, aes(x = bs_wday)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time - edit time var from char to hm (lubridate)
ghaem_df$bs_time <- hm(ghaem_df$bs_time)

ggplot(ghaem_df, aes(x = bs_time)) +
  geom_histogram(fill = "salmon2", colour = "black", binwidth=10) +
  scale_x_continuous(breaks = seq(24*3600, 24*3600))+
  facet_grid(Group ~ ., scales = "free")

```

### 6. Collapse marker data into seasons and months

-   Glycated haemoglobin levels are reportedly higher in cooler months,
    and lower in warmer months. is this true for both our control
    groups?

    -   calculate mean + sd + CI for HbA1c levels for months
        (individually) and seasons

    -   correlation between temperature & Hba1c level - months

    -   correlation between (avg) daylength and (avg) rate of change &
        Hba1c levels - month

```{r gly_haem}

# task 1: visualise with data collapsed into months & seasons

#HBA1C
#months
ghaem_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), sd = sd(g_haem, na.rm=TRUE), n = length(g_haem)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
ghaem_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(g_haem, na.rm=TRUE), sd = sd(g_haem, na.rm=TRUE), n = length(g_haem)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#GLUCOSE
#months
glucose_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(glucose, na.rm=TRUE), sd = sd(glucose, na.rm=TRUE), n = length(glucose)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
glucose_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(glucose, na.rm=TRUE), sd = sd(glucose, na.rm=TRUE), n = length(glucose)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#LACTATE
#months
lactate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(lactate, na.rm=TRUE), sd = sd(lactate, na.rm=TRUE), n = length(lactate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
lactate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(lactate, na.rm=TRUE), sd = sd(lactate, na.rm=TRUE), n = length(lactate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#PYRUVATE
#months
pyruvate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(pyruvate, na.rm=TRUE), sd = sd(pyruvate, na.rm=TRUE), n = length(pyruvate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
pyruvate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(pyruvate, na.rm=TRUE), sd = sd(pyruvate, na.rm=TRUE), n = length(pyruvate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#CITRATE
#months
citrate_month <- ghaem_df %>%
  group_by(bs_month, Group) %>%
  summarise(mean = mean(citrate, na.rm=TRUE), sd = sd(citrate, na.rm=TRUE), n = length(citrate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

#seasons
citrate_season <- ghaem_df %>%
  group_by(bs_season, Group) %>%
  summarise(mean = mean(citrate, na.rm=TRUE), sd = sd(citrate, na.rm=TRUE), n = length(citrate)) %>%
  mutate(se = sd / sqrt(n),
         lower.ci = mean - qt(1 - (0.05 / 2), n - 1) * se,
         upper.ci = mean + qt(1 - (0.05 / 2), n - 1) * se)

# calculate mean & plot glycated haem - matched controls
#ghaem_month_matched <- match_exact_df %>%
#  group_by(bs_month, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))

#matched control seasons
#ghaem_season_matched <- match_exact_df %>%
#  group_by(bs_season, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))


#if wanting to calculate median + IQR
#ghaem_month <- ghaem_df %>%
#  group_by(bs_month, Group) %>%
#  summarise(median = median(g_haem, na.rm=TRUE), IQR = IQR(g_haem, na.rm=TRUE), q1 #= quantile(g_haem, 0.25), q3 = quantile(g_haem, 0.75), N = length(g_haem))

```

### 7. Test for significant differences in marker levels between groups (per month)

```{r}

# is there a significant difference between marker level between groups per month?

jan <- ghaem_df %>%  filter(bs_month==1) 
jan_wrs <-  wilcox.test(citrate ~ Group, data=jan, exact = FALSE, conf.int = TRUE)
jan_wrs

feb <- ghaem_df %>%  filter(bs_month==2) 
feb_wrs <-  wilcox.test(citrate ~ Group, data=feb, exact = FALSE, conf.int = TRUE)
feb_wrs

march <- ghaem_df %>%  filter(bs_month==3) 
march_wrs <-  wilcox.test(citrate ~ Group, data=march, exact = FALSE, conf.int = TRUE)
march_wrs

april <- ghaem_df %>%  filter(bs_month==4) 
april_wrs <-  wilcox.test(citrate ~ Group, data=april, exact = FALSE, conf.int = TRUE)
april_wrs

may <- ghaem_df %>%  filter(bs_month==5) 
may_wrs <-  wilcox.test(citrate ~ Group, data=may, exact = FALSE, conf.int = TRUE)
may_wrs

june <- ghaem_df %>%  filter(bs_month==6) 
june_wrs <-  wilcox.test(citrate ~ Group, data=june, exact = FALSE, conf.int = TRUE)
june_wrs

july <- ghaem_df %>%  filter(bs_month==7) 
july_wrs <-  wilcox.test(citrate ~ Group, data=july, exact = FALSE, conf.int = TRUE)
july_wrs

aug <- ghaem_df %>%  filter(bs_month==8) 
aug_wrs <-  wilcox.test(citrate ~ Group, data=aug, exact = FALSE, conf.int = TRUE)
aug_wrs

sept <- ghaem_df %>%  filter(bs_month==9) 
sept_wrs <-  wilcox.test(citrate ~ Group, data=sept, exact = FALSE, conf.int = TRUE)
sept_wrs

oct <- ghaem_df %>%  filter(bs_month==10) 
oct_wrs <-  wilcox.test(citrate ~ Group, data=oct, exact = FALSE, conf.int = TRUE)
oct_wrs

nov <- ghaem_df %>%  filter(bs_month==11) 
nov_wrs <-  wilcox.test(citrate ~ Group, data=nov, exact = FALSE, conf.int = TRUE)
nov_wrs

dec <- ghaem_df %>%  filter(bs_month==12) 
dec_wrs <-  wilcox.test(citrate ~ Group, data=dec, exact = FALSE, conf.int = TRUE)
dec_wrs

```

### 8. Calculate cosinor and plot marker levels using pirate plot

```{r initialise pirate plot}

#colour gradient
col <- inferno(12, begin=0.3, end=0.7, direction=1) 
as.data.frame(col)

dayCOL<-c("#50576CFF", "#989278FF", "#ADA274FF", "#D8C563FF", "#EED654FF", "#FFEA46FF", "#EED654FF", "#D8C563FF", "#ADA274FF", "#858379FF", "#50576CFF", "#1B3B6DFF")


SeasonPal <- c("#191500","#4C4000","#7F6B00","#B29600","#E5C100","#FFD700","#E5C100","#CCAC00","#998100","#4C4000","#191500","#000000") 

names <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
namesshort <-c("J","F","M","A","M","J","J","A","S","O","N","D")
```

#### [HbA1c]{.underline}

```{r HbA1c cosinor+plot}

#run cosinor
hba1c_control_cos <- cosinor(g_haem~1, date='bs_month', type='monthly', data=control_df)
summary(hba1c_control_cos)
summary(hba1c_control_cos$glm)
#plot(hba1c_control_cos)
#seasrescheck(hba1c_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(hba1c_control_cos$glm$data, select = hba1c_control_cos$date)[, 1]
o = order(time_fit_control)

# HbA1c - control

pirateplot(formula =  g_haem ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(34.4,35),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],hba1c_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], hba1c_control_cos$fitted.values[o], pch = 19)
       

# HbA1c - bipolar

#run cosinor
hba1c_bipolar_cos <- cosinor(g_haem~1, date='bs_month', type='monthly', data=bipolar_df)
summary(hba1c_bipolar_cos)
summary(hba1c_bipolar_cos$glm)
#plot(hba1c_bipolar_cos)
#seasrescheck(hba1c_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(hba1c_bipolar_cos$glm$data, select = hba1c_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  g_haem ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(32,36.5) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],hba1c_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], hba1c_bipolar_cos$fitted.values[o], pch = 19)

```

#### [Glucose]{.underline}

```{r Glucose cosinor+plot}

#run cosinor
glucose_control_cos <- cosinor(glucose~1, date='bs_month', type='monthly', data=control_df)
summary(glucose_control_cos)
summary(glucose_control_cos$glm)
#plot(glucose_control_cos)
#seasrescheck(glucose_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(glucose_control_cos$glm$data, select = glucose_control_cos$date)[, 1]
o = order(time_fit_control)

# Glucose - control
dev.new()
pirateplot(formula =  glucose ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(4.85,5),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],glucose_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], glucose_control_cos$fitted.values[o], pch = 19)
       

# Glucose - bipolar

#run cosinor
glucose_bipolar_cos <- cosinor(glucose~1, date='bs_month', type='monthly', data=bipolar_df)
summary(glucose_bipolar_cos)
summary(glucose_bipolar_cos$glm)
#plot(glucose_bipolar_cos)
#seasrescheck(glucose_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(glucose_bipolar_cos$glm$data, select = glucose_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  glucose ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(4.5,5.5) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],glucose_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], glucose_bipolar_cos$fitted.values[o], pch = 19)

```

#### [Lactate]{.underline}

```{r Lactate cosinor+plot}

#run cosinor
lactate_control_cos <- cosinor(lactate~1, date='bs_month', type='monthly', data=control_df)
summary(lactate_control_cos)
summary(lactate_control_cos$glm)
#plot(lactate_control_cos)
#seasrescheck(lactate_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(lactate_control_cos$glm$data, select = lactate_control_cos$date)[, 1]
o = order(time_fit_control)

# Lactate - control
dev.new() # open new plot
pirateplot(formula =  lactate ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Lactate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(3.5,4.5),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],lactate_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], lactate_control_cos$fitted.values[o], pch = 19)
       

# Lactate - bipolar

#run cosinor
lactate_bipolar_cos <- cosinor(lactate~1, date='bs_month', type='monthly', data=bipolar_df)
summary(lactate_bipolar_cos)
summary(lactate_bipolar_cos$glm)
#plot(lactate_bipolar_cos)
#seasrescheck(lactate_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(lactate_bipolar_cos$glm$data, select = lactate_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  lactate ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Lactate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(3,5) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],lactate_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], lactate_bipolar_cos$fitted.values[o], pch = 19)

```

#### [Pyruvate]{.underline}

```{r Pyruvate cosinor+plot}

#run cosinor
pyruvate_control_cos <- cosinor(pyruvate~1, date='bs_month', type='monthly', data=control_df)
summary(pyruvate_control_cos)
summary(pyruvate_control_cos$glm)
#plot(pyruvate_control_cos)
#seasrescheck(pyruvate_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(pyruvate_control_cos$glm$data, select = pyruvate_control_cos$date)[, 1]
o = order(time_fit_control)

# Pyruvate - control
dev.new() # open new plot
pirateplot(formula =  pyruvate ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Pyruvate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.07,0.09),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],pyruvate_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], pyruvate_control_cos$fitted.values[o], pch = 19)
       

# Pyruvate - bipolar

#run cosinor
pyruvate_bipolar_cos <- cosinor(pyruvate~1, date='bs_month', type='monthly', data=bipolar_df)
summary(pyruvate_bipolar_cos)
summary(pyruvate_bipolar_cos$glm)
#plot(pyruvate_bipolar_cos)
#seasrescheck(pyruvate_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(pyruvate_bipolar_cos$glm$data, select = pyruvate_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  pyruvate ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Pyruvate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.05,0.12) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],pyruvate_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], pyruvate_bipolar_cos$fitted.values[o], pch = 19)

```

#### [Citrate]{.underline}

```{r Citrate cosinor+plot}

#run cosinor
citrate_control_cos <- cosinor(citrate~1, date='bs_month', type='monthly', data=control_df)
summary(citrate_control_cos)
summary(citrate_control_cos$glm)
#plot(citrate_control_cos)
#seasrescheck(citrate_control_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(citrate_control_cos$glm$data, select = citrate_control_cos$date)[, 1]
o = order(time_fit_control)

# Citrate - control
dev.new() # open new plot
pirateplot(formula =  citrate ~ bs_month,
           data = control_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Control Citrate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.062,0.068),  # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o],citrate_control_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_control[o], citrate_control_cos$fitted.values[o], pch = 19)
       

# Citrate - bipolar

#run cosinor
citrate_bipolar_cos <- cosinor(citrate~1, date='bs_month', type='monthly', data=bipolar_df)
summary(citrate_bipolar_cos)
summary(citrate_bipolar_cos$glm)
#plot(citrate_bipolar_cos)
#seasrescheck(citrate_bipolar_cos$residuals) # check the residuals

#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bipolar = subset(citrate_bipolar_cos$glm$data, select = citrate_bipolar_cos$date)[, 1]
o = order(time_fit_bipolar)

dev.new() # open new plot

pirateplot(formula =  citrate ~ bs_month,
           data = bipolar_df,
           #col = rgb(cr(daylen / max(daylen)), max=255)
           pal=col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Bipolar Citrate (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(0.05,0.08) # Adjust y axis limits
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3))
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bipolar[o],citrate_bipolar_cos$fitted.values[o], lty=2,lwd=1, col="black")
points(time_fit_bipolar[o], citrate_bipolar_cos$fitted.values[o], pch = 19)

```

### 9. Multivariable analysis

```{r multivariable analysis}


```

## Extra code for different tasks

#### extra: code for multivariable cosinor

```{r cosinor additional variables}

# test run cosinor with additional vars
test_res=cosinor(citrate~1 + average_temp, date='bs_month', type='monthly', data=control_df)
summary(test_res)
summary(test_res$glm)
plot(test_res)
seasrescheck(test_res$residuals) # check the residuals

```

#### extra: code to explore correlation between diabetic marker levels and different variables

```{r}
# seasonal variables

#### temperature
range(ghaem_df$lactate, na.rm=TRUE)  # useful code to determine ranges for ylims

#1. multi-plot: temperature and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=average_temp, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(15, 48) #ylim for plotting points
  ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=average_temp, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(1.5, 14.5) #ylim for plotting points
  ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=average_temp, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0.5, 13) #ylim for plotting points
  ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=average_temp, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0, 2) #ylim for plotting points
  ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=average_temp, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)+  
  #ylim(0, 0.3) #ylim for plotting points
  ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for temp and diabetic markers
temp_ghaem <- cor.test(ghaem_df$average_temp, ghaem_df$g_haem, method="pearson")
yarrr::apa(temp_ghaem)
temp_glucose <- cor.test(ghaem_df$average_temp, ghaem_df$glucose, method="pearson")
yarrr::apa(temp_glucose)
temp_lactate <- cor.test(ghaem_df$average_temp, ghaem_df$lactate, method="pearson")
yarrr::apa(temp_lactate)
temp_pyruvate <- cor.test(ghaem_df$average_temp, ghaem_df$pyruvate, method="pearson")
yarrr::apa(temp_pyruvate)
temp_citrate <- cor.test(ghaem_df$average_temp, ghaem_df$citrate, method="pearson")
yarrr::apa(temp_citrate)

####  daylength
#1. multi-plot: daylength and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=daylength_min, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=daylength_min, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=daylength_min, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=daylength_min, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=daylength_min, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for daylength and diabetic markers
day_ghaem <- cor.test(ghaem_df$daylength_min, ghaem_df$g_haem, method="pearson")
yarrr::apa(day_ghaem)
day_glucose <- cor.test(ghaem_df$daylength_min, ghaem_df$glucose, method="pearson")
yarrr::apa(day_glucose)
day_lactate <- cor.test(ghaem_df$daylength_min, ghaem_df$lactate, method="pearson")
yarrr::apa(day_lactate)
day_pyruvate <- cor.test(ghaem_df$daylength_min, ghaem_df$pyruvate, method="pearson")
yarrr::apa(day_pyruvate)
day_citrate <- cor.test(ghaem_df$daylength_min, ghaem_df$citrate, method="pearson")
yarrr::apa(day_citrate)


####  ROC
#1. multi-plot: photoperiod roc and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=photoperiod_roc_min, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for photoperiod roc and diabetic markers
roc_ghaem <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$g_haem, method="pearson")
yarrr::apa(roc_ghaem)
roc_glucose <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$glucose, method="pearson")
yarrr::apa(roc_glucose)
roc_lactate <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$lactate, method="pearson")
yarrr::apa(roc_lactate)
roc_pyruvate <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$pyruvate, method="pearson")
yarrr::apa(roc_pyruvate)
roc_citrate <- cor.test(ghaem_df$photoperiod_roc_min, ghaem_df$citrate, method="pearson")
yarrr::apa(roc_citrate)

#####geographic
####  latitude
#1. multi-plot: latitude and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=assess_latitude, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=assess_latitude, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=assess_latitude, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=assess_latitude, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=assess_latitude, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for latitude and diabetic markers
latitude_ghaem <- cor.test(ghaem_df$assess_latitude, ghaem_df$g_haem, method="pearson")
yarrr::apa(latitude_ghaem)
latitude_glucose <- cor.test(ghaem_df$assess_latitude, ghaem_df$glucose, method="pearson")
yarrr::apa(latitude_glucose)
latitude_lactate <- cor.test(ghaem_df$assess_latitude, ghaem_df$lactate, method="pearson")
yarrr::apa(latitude_lactate)
latitude_pyruvate <- cor.test(ghaem_df$assess_latitude, ghaem_df$pyruvate, method="pearson")
yarrr::apa(latitude_pyruvate)
latitude_citrate <- cor.test(ghaem_df$assess_latitude, ghaem_df$citrate, method="pearson")
yarrr::apa(latitude_citrate)

#####time - does not work yet, time format issues!
####  time of day
test_time <- ghaem_df
test_time$bs_time <- as.POSIXct(test_time$bs_time)
ghaem_df$bs_time <- as.POSIXct(ghaem_df$bs_time)
#1. multi-plot: sample TOD and diabetic markers
pl1 <- ggplot(ghaem_df, aes(x=bs_time, y=g_haem, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(15, 48) #ylim for plotting points
  #ylim(30, 40)  #ylim for plotting line of best fit only

pl2 <- ggplot(ghaem_df, aes(x=bs_time, y=glucose, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(1.5, 14.5) #ylim for plotting points
  #ylim(3, 7)  #ylim for plotting line of best fit only

pl3 <- ggplot(ghaem_df, aes(x=bs_time, y=lactate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0.5, 13) #ylim for plotting points
  #ylim(1, 5)  #ylim for plotting line of best fit only

pl4 <- ggplot(ghaem_df, aes(x=bs_time, y=pyruvate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 2) #ylim for plotting points
  #ylim(0, 0.3) #ylim for plotting line of best fit only
  
pl5 <- ggplot(ghaem_df, aes(x=bs_time, y=citrate, color=Group)) + 
  #geom_point()+
  geom_smooth(method=lm)#+  
  #ylim(0, 0.3) #ylim for plotting points
  #ylim(0, 0.1)  #ylim for plotting line of best fit only

figure <- ggarrange(pl1,pl2,pl3,pl4,pl5,
                    labels = c("HbA1c", "Glucose", "Lactate", "Pyruvate", "Citrate"),
                    ncol = 3, nrow = 2)
figure

# calculate correlation for time and diabetic markers
time_ghaem <- cor.test(ghaem_df$bs_time, ghaem_df$g_haem, method="pearson")
yarrr::apa(time_ghaem)
time_glucose <- cor.test(ghaem_df$bs_time, ghaem_df$glucose, method="pearson")
yarrr::apa(time_glucose)
time_lactate <- cor.test(ghaem_df$bs_time, ghaem_df$lactate, method="pearson")
yarrr::apa(time_lactate)
time_pyruvate <- cor.test(ghaem_df$bs_time, ghaem_df$pyruvate, method="pearson")
yarrr::apa(time_pyruvate)
time_citrate <- cor.test(ghaem_df$bs_time, ghaem_df$citrate, method="pearson")
yarrr::apa(time_citrate)


```

#### extra: plotting

```{r plotting}
 
#plot mean values plus 95%CI
ggplot(citrate_season, aes(bs_season, mean, color=Group))+
  geom_line()+ggtitle("citrate seasonal Profile")+
  geom_point()+
  geom_errorbar(aes(ymin=lower.ci, ymax=upper.ci), width=.2,
                position=position_dodge(0.05))+
  scale_x_continuous("bs_season", labels = as.character(citrate_season$bs_season), breaks = citrate_season$bs_season)

# plot temp correlation
ggplot(ghaem_df, aes(x=average_temp, y=g_haem, color=Group)) + 
  geom_point()+
  geom_smooth(method=lm)+  
  ylim(20, 50)
model <- ghaem_df %>%
  filter(Group=="control") %>%
  lm(g_haem~average_temp)
temp_ghaem <- cor.test(ghaem_df$average_temp, ghaem_df$g_haem, method=c("pearson", "kendall", "spearman"))
```

#### extra: correlations between different variables

```{r}
#check assumptions
#lm code
model <- ghaem_df %>%
  filter(Group=="control") %>%
  lm(g_haem~average_temp, data=ghaem_df)
check_model(ghaem_df$age)

```

#### extra: create matched control group

1:1 matching based on sex, age, ethnicity, BMI, assess_centre, bs_month,
bs_year

CURRENT DEFAULT: do not run this code section. for optional future use
(sensitivity analyses)

```{r match, eval=FALSE, include=FALSE, verbose=FALSE}
# convert matching variables to type 'factor' for matchit function to work

for_match_df <- ghaem_df
for_match_df$Group<- as.factor(as.character(for_match_df$Group))

# remove people with NA variables: ethnicity
for_match_df<- for_match_df %>%  filter(!is.na(ethnicity))
#for_match_df<- for_match_df %>%  filter(!is.na(BMI))

#use 1:1 nearest neighbour matching (Matchit default) without replacement on a propensity score estimated using a
# logistic regression of the treatment on the covariates
m.out1 <- matchit(Group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year, data = for_match_df)

match_control_df <- match.data(m.out1)

#try different matching method: exact
m.out_exact <- matchit(Group ~ age + sex + ethnicity + assess_centre + bs_month + bs_year, data = for_match_df, method="nearest", exact= ~ bs_month + age + sex + assess_centre)
match_exact_df <- match.data(m.out_exact)
```

## 
