---
title: "seasonal_diabetic_markers_final_analysis"
author: "agpr141"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Seasonal variation in diabetic markers

**Final analysis script**

This .Rmd is used for the analyses in [paper] by Roguski et al. [date].

#### Load packages

```{r load packages}
library(tidyverse)
library(dplyr)
library(readr)
library(yarrr)
library(season)
library(wesanderson)
library(circacompare)
```

#### Load data

```{r load data}

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Seasonal Variation Diabetic Markers - Analysis\\bd_dm_group.Rda")

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Seasonal Variation Diabetic Markers - Analysis\\bd_nondm_group.Rda")

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Seasonal Variation Diabetic Markers - Analysis\\c_dm_group.Rda")

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Seasonal Variation Diabetic Markers - Analysis\\c_nondm_group.Rda")

load(file="C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\BD-Metabolomics data\\Seasonal Variation Diabetic Markers - Analysis\\svdm_all_ppts.Rda")

```

#### Table 1

```{r table 1, echo=FALSE}

# summary statistics - mean for continuous variables:
demo_mean_df <- svdm_all_ppts %>%
  group_by(group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter', 'fasting_time', 'townsend', 'sleep_duration', 'g_haem', 'glucose'),
               mean, na.rm=TRUE)

# summary statistics - std.dev for continuous variables:
demo_sd_df <- svdm_all_ppts %>%
  group_by(group) %>%
  summarise_at(c('age', 'BMI', 'age_completed_education', 'deprivation_index_england', 'deprivation_index_scotland', 'deprivation_index_wales', 'time_outdoors_summer', 'time_outdoors_winter', 'fasting_time', 'townsend', 'sleep_duration', 'g_haem', 'glucose'),
               sd, na.rm=TRUE)

#counts for discrete variables:
demo_sex_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(sex)
demo_ethnicity_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(ethnicity_recode)
#1=White, 2=Mixed, 3=Asian/Asian British, 4=Black/Black British, 5=Chinese, 6=Other ethnic group, 7=do not know/prefer not to answer

demo_disabled_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(disability_self_report)    
demo_chronotype_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(chronotype)    
demo_employment_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(employment_recode) 
demo_smoking_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(smoking_status)  
demo_alcohol_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(alcohol_intake)
demo_healthstatus_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(health_self_report) 
demo_IPAQ_df <- svdm_all_ppts %>%
  group_by(group) %>%
  count(IPAQ_group) 


#plot data to visualise distribution
# age - normal
age_plot <- ggplot(ghaem_df, aes(x = age)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# bmi - right skew
bmi_plot <- ggplot(ghaem_df, aes(x = BMI)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# age completed education - normal
age_edu_plot <- ggplot(ghaem_df, aes(x = age_completed_education)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: England - right skew
imd_e_plot <- ggplot(ghaem_df, aes(x = deprivation_index_england)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: Scotland - right skew
imd_s_plot <- ggplot(ghaem_df, aes(x = deprivation_index_scotland)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# IMD: Wales - right skew
imd_w_plot <- ggplot(ghaem_df, aes(x = deprivation_index_wales)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time outdoors summer - right skew
outsummer_plot <- ggplot(ghaem_df, aes(x = time_outdoors_summer)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# time outdoors winter - right skew
outwinter_plot <- ggplot(ghaem_df, aes(x = time_outdoors_winter)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# fasting time - right skew
fastingtime_plot <- ggplot(ghaem_df, aes(x = fasting_time)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# townsend - right skew
townsend_plot <- ggplot(ghaem_df, aes(x = townsend)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")


# sleep duration - right skew
sleepdur_plot <- ggplot(ghaem_df, aes(x = sleep_duration)) +
  geom_histogram(fill = "salmon2", colour = "black") +
  facet_grid(Group ~ ., scales = "free")

# plot in multi-plot
figure <- ggarrange(age_plot,bmi_plot,age_edu_plot,imd_e_plot,imd_s_plot,imd_w_plot,outsummer_plot,outwinter_plot, fastingtime_plot,townsend_plot, labels = c("Age", "BMI", "Age Education", "IMD_E", "IMD_S", "IMD_W", "outdoors summer", "outdoors winter", "fasting time", "townsend"),
                    ncol = 5, nrow = 2)
figure


# stats tests - chi square for categorical
#sex
sex_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$sex))
yarrr::apa(sex_chi)# print stats test result in legible apa format

#ethnicity
ethnicity_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$ethnicity_recode))
yarrr::apa(ethnicity_chi)

#employment
employment_chi <- chisq.test(x = table(ghaem_df$Group,ghaem_df$employment_recode))
yarrr::apa(employment_chi)

#disability
disability_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$disability_self_report=='Yes'))
yarrr::apa(disability_chi)

#smoking
smoking_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$smoking_status))
yarrr::apa(smoking_chi)

#alcohol
alcohol_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$alcohol_intake))
yarrr::apa(alcohol_chi)

#chronotype
chrono_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$chronotype))
yarrr::apa(chrono_chi)

#health self-report
health_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$health_self_report))
yarrr::apa(health_chi)

#IPAQ
IPAQ_chi <- chisq.test(x = table(ghaem_df$Group, ghaem_df$IPAQ_group))
yarrr::apa(IPAQ_chi)

# stats tests - t-test for parametric continous
#age
age_ttest <- t.test(ghaem_df$age ~ ghaem_df$Group, na.rm=TRUE)
yarrr::apa(age_ttest)

# stats tests - wilcoxon rank sum for non-parametric continous
#BMI
bmi_wrs <- wilcox.test(BMI ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
bmi_wrs

#IMD ENGLAND
imd_e_wrs <- wilcox.test(deprivation_index_england ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_e_wrs

#IMD SCOTLAND
imd_s_wrs <- wilcox.test(deprivation_index_scotland ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_s_wrs

#IMD WALES
imd_w_wrs <- wilcox.test(deprivation_index_wales ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
imd_w_wrs

#age completed education
edu_age_wrs <- wilcox.test(age_completed_education ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
edu_age_wrs

#time outdoors summer
outsummer_wrs <- wilcox.test(time_outdoors_summer ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
outsummer_wrs

#time outdoors winter
outwinter_wrs <- wilcox.test(time_outdoors_winter ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
outwinter_wrs

#fasting time
fasting_wrs <- wilcox.test(fasting_time ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
fasting_wrs

#townsend
townsend_wrs <- wilcox.test(townsend ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
townsend_wrs

#sleep duration
sleepdur_wrs <- wilcox.test(sleep_duration ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
sleepdur_wrs


### our marker levels
#hba1c
ghaem_ttest <- t.test(ghaem_df$g_haem ~ ghaem_df$Group, na.rm=TRUE)
yarrr::apa(ghaem_ttest)

#glucose
glucose_wrs <- wilcox.test(glucose ~ Group, data = ghaem_df, exact = FALSE, conf.int = TRUE)
glucose_wrs
```

#### Plots

#### 

```{r initialise plot}

#colour gradient
#col <- inferno(12, begin=0.3, end=0.7, direction=1) 
#as.data.frame(col)

dayCOL<-c("#50576CFF", "#989278FF", "#ADA274FF", "#D8C563FF", "#EED654FF", "#FFEA46FF", "#EED654FF", "#D8C563FF", "#ADA274FF", "#858379FF", "#50576CFF", "#1B3B6DFF")


SeasonPal <- c("darkseagreen3","darkseagreen3","darkseagreen3","lightgoldenrod2","lightgoldenrod2","lightgoldenrod2","indianred3","indianred3","indianred3","lightsteelblue2","lightsteelblue2","lightsteelblue2") 
#0450b4, #046dc8, #1184a7, #15a2a2, #6fb1a0, #b4418e, #d94a8c, #ea515f, #fe7434, #fea802

dummy <- c_nondm_group %>% 
  dplyr::select(assess_month, average_temp) %>%
  group_by(assess_month) %>%
  summarise(colo = mean(average_temp, na.rm=TRUE))


amber_col <- colorRampPalette(as.character(wes_palette("Zissou1", n=12, type='continuous'))) #colorRampPalette(c("#15A2A2", "#ea515f"))
#detach("package:wesanderson", unload=TRUE)
amber_col_list <- amber_col(12)
dummy_order <- dummy[order(dummy$colo),]
dummy_order$hex[1]= amber_col_list[1]
dummy_order$hex[2]= amber_col_list[2]
dummy_order$hex[3]= amber_col_list[3]
dummy_order$hex[4]= amber_col_list[4]
dummy_order$hex[5]= amber_col_list[5]
dummy_order$hex[6]= amber_col_list[6]
dummy_order$hex[7]= amber_col_list[7]
dummy_order$hex[8]= amber_col_list[8]
dummy_order$hex[9]= amber_col_list[9]
dummy_order$hex[10]= amber_col_list[10]
dummy_order$hex[11]= amber_col_list[11]
dummy_order$hex[12]= amber_col_list[12]
dummy_order <- dummy_order[order(dummy_order$assess_month),]
temp_col <- as.list(dummy_order$hex)

names <- c("Jan","Feb","Mar","Apr","May","Jun","Jul","Aug","Sep","Oct","Nov","Dec")
namesshort <-c("J","F","M","A","M","J","J","A","S","O","N","D")
```

#### HbA1c control non-diabetic

```{r HbA1c all control}

#run cosinor
hba1c_c_nondm_cos <- cosinor(g_haem~1 #+ control_df$age + control_df$BMI + control_df$sex_f + control_df$ethnicity_f
                             , date='assess_month', type='monthly',data=c_nondm_group)
control_cos_summary <- summary(hba1c_c_nondm_cos)
#summary(hba1c_c_nondm_cos$glm)
mesor_control <- mean(hba1c_c_nondm_cos$glm$fitted.values) # for mesor
#plot(hba1c_c_nondm_cos)
#seasrescheck(hba1c_c_nondm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(hba1c_c_nondm_cos$glm$data, select = hba1c_c_nondm_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control

pirateplot(formula =  g_haem ~ assess_month,
           data = c_nondm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(34,35.5),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "HbA1c Seasonal Profile - Non-diabetic population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],hba1c_c_nondm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], hba1c_c_nondm_cos$fitted.values[o_c], pch = 19)
legend(10, 60, legend=c("Average Temperature (°c)"),
       #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
       lty=1, lwd=5, cex=0.8)

## potential code for 
plot.new()
season_cols <- viridis(50, option ="E")

tiff("legend_temp1.tiff", width = 7, height = 6, units = 'in', res = 300)
plot.new()
colkey(clim = range(1:12),
       col = temp_col,
      dist = -0.2, 
       #shift = 0.5,
       side = 4, 
       add = TRUE, 
       clab = "Average temperature", 
       line.clab=.5,#lower title
       cex.clab = 1,
       col.clab = "black",
       adj.clab=0,
       length = .5, 
       width = 1.5, 
       col.axis = "black", 
       col.ticks = "black",
       col.box = "white",
       at = c(6.5,17.5), 
       #labels = c("6hr","18hr"),
       mgp=c(3, .3, 0),
       tck=0,
       cex.axis = 1,
       lwd.ticks=.2)
dev.off()

```

#### HbA1c control diabetic

```{r hba1c diabetic}

#run cosinor
hba1c_c_dm_cos <- cosinor(g_haem~1 #+ c_dm_group$age + c_dm_group$BMI + c_dm_group$sex_f + c_dm_group$ethnicity_f
                             , date='assess_month', type='monthly',data=c_dm_group)
c_dm_cos_summary <- summary(hba1c_c_dm_cos)
#summary(hba1c_c_dm_cos$glm)
mesor_control <- mean(hba1c_c_dm_cos$glm$fitted.values) # for mesor
#plot(hba1c_c_dm_cos)
#seasrescheck(hba1c_c_dm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(hba1c_c_dm_cos$glm$data, select = hba1c_c_dm_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control
dev.new()
pirateplot(formula =  g_haem ~ assess_month,
           data = c_dm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(45,50),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "HbA1c Seasonal Profile - Diabetic population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],hba1c_c_dm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], hba1c_c_dm_cos$fitted.values[o_c], pch = 19)
legend(10, 60, legend=c("Average Temperature (°c)"),
       #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
       lty=1, lwd=5, cex=0.8)
```

#### Glucose non-diabetic

```{r hba1c diabetic}

#run cosinor
glu_c_nondm_cos <- cosinor(glucose~1 #+ c_dm_group$age + c_dm_group$BMI + c_dm_group$sex_f + c_dm_group$ethnicity_f
                             , date='assess_month', type='monthly',data=c_nondm_group)
c_nondm_cos_summary <- summary(glu_c_nondm_cos)
#summary(glu_c_nondm_cos$glm)
mesor_control <- mean(glu_c_nondm_cos$glm$fitted.values) # for mesor
#plot(glu_c_nondm_cos)
#seasrescheck(glu_c_nondm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(glu_c_nondm_cos$glm$data, select = glu_c_nondm_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control
dev.new()
pirateplot(formula =  glucose ~ assess_month,
           data = c_nondm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(4.4,5.4),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "Glucose Seasonal Profile - Non-diabetic population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],glu_c_nondm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], glu_c_nondm_cos$fitted.values[o_c], pch = 19)
legend(10, 60, legend=c("Average Temperature (°c)"),
       #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
       lty=1, lwd=5, cex=0.8)
```

#### Glucose diabetic

```{r glucose diabetic}

#run cosinor
glu_c_dm_cos <- cosinor(glucose~1 #+ c_dm_group$age + c_dm_group$BMI + c_dm_group$sex_f + c_dm_group$ethnicity_f
                             , date='assess_month', type='monthly',data=c_dm_group)
c_dm_cos_summary <- summary(glu_c_dm_cos)
#summary(glu_c_dm_cos$glm)
mesor_control <- mean(glu_c_dm_cos$glm$fitted.values) # for mesor
#plot(glu_c_dm_cos)
#seasrescheck(glu_c_dm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(glu_c_dm_cos$glm$data, select = glu_c_dm_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control
dev.new()
pirateplot(formula =  glucose ~ assess_month,
           data = c_dm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(6.2,7.2),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "Glucose Seasonal Profile - Diabetic population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],glu_c_dm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], glu_c_dm_cos$fitted.values[o_c], pch = 19)
legend(10, 60, legend=c("Average Temperature (°c)"),
       #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
       lty=1, lwd=5, cex=0.8)
```

#### Circacompare - control glucose

```{r}
#create new control_df
control_df <- filter(svdm_all_ppts, group == 'control dm' | group == 'control nondm')

complete
cos_stats_glucose <- circacompare(x=control_df, col_time="assess_month", 
                                  col_group="group", col_outcome="glucose", 
                                  period=12, alpha_threshold=0.05, 
                                  control = c('sex', 'age', 'BMI', 'ethnicity_f'))
```

#### Circacompare - control HbA1c

```{r}
cos_stats_ghaem <- circacompare(x=control_df, col_time="assess_month", 
                                  col_group="group", col_outcome="g_haem", 
                                  period=12, alpha_threshold=0.1, 
                                  control = c('sex', 'age', 'BMI', 'ethnicity_f')) #set alpha to 0.1 as p=0.054 for diabetic group
```

#### HbA1c - bipolar non-diabetic

```{r HbA1c all control}

#run cosinor
hba1c_bd_nondm_cos <- cosinor(g_haem~1 #+ control_df$age + control_df$BMI + control_df$sex_f + control_df$ethnicity_f
                             , date='assess_month', type='monthly',data=bd_nondm_group)
bd_cos_summary <- summary(hba1c_bd_nondm_cos)
#summary(hba1c_bd_nondm_cos$glm)
mesor_control <- mean(hba1c_bd_nondm_cos$glm$fitted.values) # for mesor
#plot(hba1c_bd_nondm_cos)
#seasrescheck(hba1c_bd_nondm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_bd = subset(hba1c_bd_nondm_cos$glm$data, select = hba1c_bd_nondm_cos$date)[, 1]
o_c = order(time_fit_bd)

# HbA1c - plot

pirateplot(formula =  g_haem ~ assess_month,
           data = bd_nondm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(30,40),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "HbA1c Seasonal Profile - Non-diabetic bipolar population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_bd[o_c],hba1c_bd_nondm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_bd[o_c], hba1c_bd_nondm_cos$fitted.values[o_c], pch = 19)
legend(10, 60, legend=c("Average Temperature (°c)"),
       #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
       lty=1, lwd=5, cex=0.8)


```

#### HbA1c bipolar diabetic

```{r hba1c diabetic}

#run cosinor
hba1c_bd_dm_cos <- cosinor(g_haem~1 #+ c_dm_group$age + c_dm_group$BMI + c_dm_group$sex_f + c_dm_group$ethnicity_f
                             , date='assess_month', type='monthly',data=bd_dm_group)
bd_dm_cos_summary <- summary(hba1c_bd_dm_cos)
#summary(hba1c_bd_dm_cos$glm)
mesor_control <- mean(hba1c_bd_dm_cos$glm$fitted.values) # for mesor
#plot(hba1c_bd_dm_cos)
#seasrescheck(hba1c_bd_dm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(hba1c_bd_dm_cos$glm$data, select = hba1c_bd_dm_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control
dev.new()
pirateplot(formula =  g_haem ~ assess_month,
           data = bd_dm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("HbA1c (mmol/mol)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(25,65),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "HbA1c Seasonal Profile - Diabetic bipolar population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],hba1c_bd_dm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], hba1c_bd_dm_cos$fitted.values[o_c], pch = 19)
#legend(10, 60, legend=c("Average Temperature (°c)"),
       #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
       #lty=1, lwd=5, cex=0.8)
```

#### Glucose non-diabetic

```{r hba1c diabetic}

#run cosinor
glu_bd_nondm_cos <- cosinor(glucose~1 #+ c_dm_group$age + c_dm_group$BMI + c_dm_group$sex_f + c_dm_group$ethnicity_f
                             , date='assess_month', type='monthly',data=bd_nondm_group)
bd_nondm_cos_summary <- summary(glu_bd_nondm_cos)
#summary(glu_bd_nondm_cos$glm)
mesor_control <- mean(glu_bd_nondm_cos$glm$fitted.values) # for mesor
#plot(glu_bd_nondm_cos)
#seasrescheck(glu_bd_nondm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(glu_bd_nondm_cos$glm$data, select = glu_bd_nondm_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control
dev.new()
pirateplot(formula =  glucose ~ assess_month,
           data = bd_nondm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(4.6,5.8),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "Glucose Seasonal Profile - Non-diabetic bipolar population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],glu_bd_nondm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], glu_bd_nondm_cos$fitted.values[o_c], pch = 19)
#legend(10, 60, legend=c("Average Temperature (°c)"),
 #      #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
  #     lty=1, lwd=5, cex=0.8)
```

#### Glucose diabetic

```{r glucose diabetic}

#run cosinor
glu_bd_dm_cos <- cosinor(glucose~1 #+ c_dm_group$age + c_dm_group$BMI + c_dm_group$sex_f + c_dm_group$ethnicity_f
                             , date='assess_month', type='monthly',data=bd_dm_group)
bd_dm_cos_summary <- summary(glu_bd_dm_cos)
#summary(glu_bd_dm_cos$glm)
mesor_control <- mean(glu_bd_dm_cos$glm$fitted.values) # for mesor
#plot(glu_bd_dm_cos)
#seasrescheck(glu_bd_dm_cos$residuals) # check the residuals


#create new vars for plotting fitted sinusoid from a Cosinor object
time_fit_control = subset(glu_bd_dm_cos$glm$data, select = glu_bd_dm_cos$date)[, 1]
o_c = order(time_fit_control)

# HbA1c - control
dev.new()
pirateplot(formula =  glucose ~ assess_month,
           data = bd_dm_group,
           #col = rgb(cr(daylen / max(daylen)), max=255),
           pal=temp_col, #set to your own color palette created previously
           xlab = "Month",
           ylab = expression(bold(paste("Glucose (mmol/L)"))),
           yaxt="n",
           xaxt = "n",
           theme=0, #set theme to 0
           bty="n",
           #bean.f.o = .4, # Bean fill
           #point.o = .4, # Points
           inf.f.o = .8, # Inference fill
           inf.b.o = 1, # Inference border
           inf.f.col = temp_col,
           inf.method = "ci",
           avg.line.o = 1, # Average line
           #bty=n,
           #point.cex = .5, # Points
           #quant = c(.1, .9), # Adjust quantiles
           ylim=c(2,10),  # Adjust y axis limits
           #ylim=c(32,36.5)
           #gl.col = gray(.2), # Gridline specifications
           #gl.lty = 1,
           #gl.lwd = c(0, .3)),
           main = "Glucose Seasonal Profile - Diabetic bipolar population"
           )

axis(2,col.axis="black", lwd=2, las=2, cex.axis = 0.6)
axis(1, at=1:12, labels = namesshort, lwd=2, cex.axis = 0.8,font=2)
lines(time_fit_control[o_c],glu_bd_dm_cos$fitted.values[o_c], lty=2,lwd=1, col="black")
points(time_fit_control[o_c], glu_bd_dm_cos$fitted.values[o_c], pch = 19)
#legend(10, 60, legend=c("Average Temperature (°c)"),
#       #col=c("darkseagreen3", "lightgoldenrod2", "indianred3","lightsteelblue2"),
#       lty=1, lwd=5, cex=0.8)
```

#### Circacompare - glucose

```{r}
#create new control_df
bd_df <- filter(svdm_all_ppts, group == 'bipolar dm' | group == 'bipolar nondm')


cos_stats_glucose <- circacompare(x=bd_df, col_time="assess_month", 
                                  col_group="group", col_outcome="glucose", 
                                  period=12, alpha_threshold=0.8, 
                                  control = c('sex', 'age', 'BMI', 'ethnicity_f'))
```

#### Circacompare - HbA1c

```{r}
cos_stats_ghaem <- circacompare(x=bd_df, col_time="assess_month", 
                                  col_group="group", col_outcome="g_haem", 
                                  period=12, alpha_threshold=0.5, 
                                  control = c('sex', 'age', 'BMI', 'ethnicity_f')) #set alpha to 0.1 as p=0.054 for diabetic group
```
