---
title: "get weather data"
author: "Cathy Wyse"
date: '2023-09-27'
output: html_document
---
This code extracts mean monthly outdoor temperature for each assessment centre and year.  The month and year are taken from the baseline assessment date.  Average monthly temperature is used to avoid variations from unusual daily temperatures.  

The nearest weather station to the assessment centres (that had historical data) were:  
  
Weather Data Table  

station | near	        |Category  |centre code 
:-------|:--------------|:---------|:----------- 
shef	  |Cheadle        |(revisit) |11024 
shef	  |Cheadle        |(imaging) |11025 
Brad  	|Leeds	        |          |11010  
Card  	|Cardiff        |          |11003 
Card	  |Swansea        |          |11022 
Durh	  |Newcastle      |          |11009 
Heat  	|Barts	        |          |11012 
Heat  	|Croydon        |          |11020 
Heat  	|Hounslow       |	         |11018 
Heat  	|Reading        |	         |11007 
Leuc  	|Edinburgh      |          |11005 
Oxfo  	|Oxford         |          |11002 
Pais  	|Glasgow        |	         |11004 
Ross  	|Bristol        |          |11011 
Shaw  	|Wrexham        |          |11023 
Shef  	|Stockport      |(pilot)   |10003 
Shef  	|Bury	          |          |11008  
Shef  	|Liverpool      |	         |11016 
Shef  	|Manchester     |          |11001 
Shef  	|Sheffield      |	         |11014 
Shef  	|Stoke          |          |11006 
Sutt  	|Birmingham     |          |11021 
Sutt  	|Nottingham     |          |11013 
Whit  	|Middlesborough |          |11017 
 
The data are taken from Met Office website
https://www.metoffice.gov.uk/research/climate/maps-and-data/historic-station-data

**Historic station data consists of:**  
Mean daily maximum temperature (tmax)  
Mean daily minimum temperature (tmin)  
Days of air frost (af)  
Total rainfall (rain)  
Total sunshine duration (sun)  
Monthly mean temperature  

The monthly mean temperature is calculated from the average of the mean daily maximum and mean daily minimum temperature i.e. (tmax+tmin)/2.  
 
Missing data  
Missing data (more than two days missing in a month) have been indicated by "---". Estimated data (where available) have been indicated by "*" after the value.  

Sunshine data  
Sunshine data using a Kipp and Zonen sensor have been indicated by a "#" after the value. All other sunshine data have been recorded using a Campbell-Stokes recorder.  

The file containing the temperature data is called weather_ukb_270923.xlsx.  Only data for average temperature have been extracted.  

```{r import data}
library(pillar)
library(readxl)
setwd("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\biobank-time-dimensions\\Projects\\BD-Metabolomics")
temp_table <- read_excel("./weather_ukb_270923.xlsx", sheet = "monthly temp")
colnames(temp_table) <- c("station","assess_year", "assess_month", "tmax", "tmin", "frost", "rain", "sun", "average_temp")
weather_stations <- read_excel("weather_ukb_270923.xlsx", sheet = "ukb centres - weather stations")
#UKB_master <- read.csv("C:/UKB-time/UKB_master") - use ghaem_df instead
colnames(weather_stations) <- c("station","assess_centre_name", "assess_centre")
```

Link eid to year, month and weather station

They are linked by the month, year and assessment centre for each eid to give a value for average monthly temperature for each participant at the baseline assessment visit.


```{r merge}
library(dplyr)

setwd("C:\\Users\\arogusk2\\OneDrive - University of Edinburgh\\HELIOS-BD\\Side Projects\\BioBank Project\\biobank-time-dimensions\\Projects\\BD-Metabolomics")

#check station names & clean up
working_df <- ghaem_df

# check main df
unique(working_df$station)  #lists 12 stations with 1st letter capital

# check weather_stations and remove shef for revisits
unique(weather_stations$station)  #lists 13 stations with 1st letter capital EXCEPT lower case shef (revisit & imaging)


unique(temp_table$station)  #lists 12 stations with 1st letter lower case EXCEPT upper case Card, Brad
#merge weather stations and centre
working_df  <- analysis_df
working_df  = subset(working_df , select = -c(station,tmax, tmin,frost, rain, sun, average_temp))
working_df <- merge(working_df,weather_stations, by="assess_centre")

#merge dataframes based on year and month and centre
working_df <- merge(working_df, temp_table, by = c("assess_year", "station", "assess_month"), all.x = TRUE)

```

